### 2013-09-23 docs/proposal/doc.tex
Added initial LaTeX structure for proposal. Also included a bibliograhpy that contains related citations.

% Needed packages
\documentclass[a4paper, 10pt, english, twocolumn]{article}
\usepackage[english]{babel}
\usepackage[cm]{fullpage}
\usepackage{cite}
\usepackage{anysize}
\usepackage[compact]{titlesec}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}

\usepackage{amssymb,amsmath}
\usepackage{algorithmicx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\newcommand{\keywords}[1]{\par\noindent 
{\bf Keywords\/}. #1}

% Margins & Headers
\marginsize{2.5cm}{2.5cm}{2cm}{1.5cm}
\columnsep 0.4in
\footskip 0.4in 
\usepackage{changepage}

% E-mail formatting
\usepackage{color,hyperref}
    \catcode`\_=11\relax
    \newcommand\email[1]{\_email #1\q_nil}
    \def\_email#1@#2\q_nil{
      \href{mailto:#1@#2}{{\emailfont #1\emailampersat #2}}
    }
    \newcommand\emailfont{\sffamily}
    \newcommand\emailampersat{{\color{red}\small@}}
    \catcode`\_=8\relax

% List modifications
\newenvironment{packed_item}{
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{itemize}}

\newenvironment{packed_enum}{
\begin{enumerate}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{enumerate}}

% ### Mathematics ###
\newcommand{\bpm}{\begin{pmatrix}}
\newcommand{\epm}{\end{pmatrix}}

\newcommand{\bbm}{\begin{bmatrix}}
\newcommand{\ebm}{\end{bmatrix}}

\newcommand{\bsm}{\bigl( \begin{smallmatrix}}
\newcommand{\esm}{ \end{smallmatrix} \bigl)}

\newcommand{\mbf}{\mathbf}

% ### Matrices and Vectors ###
\newcommand{\mtx}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand*\Let[2]{\State #1 $\gets$ #2}

% ### Sets ###
\newcommand{\set}[1]{\ensuremath{\mathcal{#1}}}

% ### Other ###

\newcommand{\transpose}{^{T}}
\newcommand{\inv}{^{-1}}
\newcommand{\pseudoinv}{^{+}}
% ############## End Macros ##############

% Title
\title{\fontfamily{phv}\selectfont{Research A Proposal}}
\author{
  \textbf{Ramon Janssen} - \href{mailto:ramon.janssen@student.ru.nl}{ramon.janssen@student.ru.nl} \\
  \textbf{Tom de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}
}

\date{\fontfamily{ptm}\selectfont{\small{\bfseries{\today - Radboud
Universiteit Nijmegen}}}\\[0.5cm]\rule{\linewidth}{0.3mm}}

\begin{document}

\maketitle

\begin{abstract}
Lorem Ipsum Dolor Sit Amet.
\keywords{Keywords}
\end{abstract}

\setlength{\parindent}{0.0cm}
\setlength{\parskip}{0.25cm}

\section{Introduction}

% NOTE: De delen hieronder zijn letterlijk overgenomen uit een eerder werk. Ik weet niet of we het nog nodig hebben, daarom heb ik het laten staan. We moeten het dan wel veranderen, anders pleeg ik plagiaat op mezelf.

% Introductory small-talk
%The human brain is a big network which we have yet to fully understand.
%Different areas of this network each have their own task and function, but complexity also arises from how these are connected.
%Our brains are a robust piece of equipment, presumably also a consequence of how different areas are interconnected, capable of adapting to continuous environmental change and learning throughout our entire lives.

% Motivation
%This is part of why it is interesting to investigate neuro-degenerative disease, which are able to cripple our usually robust networks effectively.
%We want to learn more about our brains, as well as the diseases that target them.

%Brain connectivity analysis can teach us a lot of how the human brain functions.
%The fields of functional and structural connectivity have, presumably for that reason, gained in popularity over the past few years \cite{vandenheuvel2010}.

%Problem
% ...

% Structure of the rest of proposal
% ...

\section{Theory}

\subsection*{Brain connectivity}
% Functional connectivity
%Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.
%An increasing body of studies successfully investigates functional connectivity by applying statistical methods on the level of co-activation of resting-state fMRI time-series between brain regions.
%Several studies have shown that resting-state networks reflect brain connectivity in a concise way, though their exact relation remains unknown \cite{Lowe2000, doria2010, Bullmore2009}.
%Resting state indicating that subjects are instructed to relax without thinking of something in particular, to stimulate spontaneous brain activity.

% Structural connectivity
%Similarly, there is the field of structural connectivity, which is defined as the mapping of direct anatomical brain pathways \cite{friston1994}.
%It refers to white matter tracts, describing the millions of long-distance axons or `brain highways', connecting large groups of spatially separated brain areas.
%In accordance with intuition, a recent study has shown that both fields are related.
%By demonstrating that almost all functionally linked regions of the most reported resting-state networks are also structurally connected by known white matter tracts \cite{vandenheuvel2009}.
%Also, functional connectivity is constrained by structural connectivity \cite{cabral2012}.
%Structural connectivity is less time dependent, as it involves the mapping of anatomical connections opposing activity patterns.

% Effective connectivity
% ...

\subsection*{Bayesian connectomics}
% ...

\subsection*{Causal discovery methods}
% ...

\subsection*{Data acquisition \& analysis}
% ...

%Within these areas methods are applied on connectivity patterns obtained from fMRI analysis to predict anatomical connections.
%It is possible to obtain anatomical connectivity patterns by applying partial correlation methods to resting-state fMRI data, filtering out poly-synaptic activity. %TODO: Reference needed.
%Another option is to apply Diffusion Tensor Imaging (DTI) to resting-state data, which maps water diffusion directions in the brain, as this is a more direct way of charting brain highways. %TODO: Reference needed.

%Both methods, i.e. correlation analysis and DTI, have downsides and ideally one would hope to combine them to have the best of both worlds.
%% Note: de onderstaande uitspraak klopt dus niet voor ons proposal
%Such a thing is possible when using Bayesian methods. In this proposal however, we focus on frequentistic or data driven methods, as they usually provide a computational advantage without a large performance trade-off.
%However, this also limits the amount of brain modelling we can do.

% Note: bekijk even of de onderstaande sectie Ÿberhaupt wel interessant is. Ik heb het gebruikt als introductie voor mijn formele uitleg. Die laatste ontbreekt nu natuurlijk.

% Brain robustness and neuro-degeneration
%\subsection*{Brain robustness and neuro-degeneration}
%Investigating the brain as an integrative network of interacting regions has proven to be a fruitful platform for learning more about how the brain functions.
%Its main premise is that the brain can be modelled as a graph $G=(\set{V},E)$, $\set{V}$ being a collection of brain areas and $E$ the connections between those.
%It provides ideas for brain robustness and forms a basis for approaching certain neuro-degenerative diseases, e.g. Alzheimer's disease (AD), as these are known to affect brain connectivity \cite{Bullmore2009, vandenheuvel2010}.
%Several studies have shown that the brain network is organised according to an efficient \emph{small-world} organization \cite{vandenheuvel2008, vandenheuvel2010}.
%Some of these suggest that brain connectivity is distributed according to a power-law function, possibly explaining the brain's scale-free organisation of connectivity networks.
%Scale-free networks are renowned for their robustness against random attacks and so it forms an interesting hypotheses for modelling brain networks.
%Even more, since scale-free organisation is susceptible to specialised attacks on hub nodes, it could be so that these hub nodes are affected in AD.
%This idea is based on resting state fMRI data and graph analysis, as well as MEG results \cite{buckner2009, stam2009}

% Note: voor onderstaande tekst geldt: kijk even kritisch of we het nodig hebben.

%\subsection{Analysing connectivity}
%Where brain data is usually available in abundant amounts, testing subjects are not.
%This, and large cross-subject divergence makes it difficult to compare (diseased) brain area's between subjects and conclusions between studies.
%To still pinpoint Regions of Interest (ROI) between subject, one can use so called labelling atlases, such as the Automatic Anatomical Labeling atlas, to map brain areas to known ROI's.
%Another option is to apply a custom clustering to brain data such as attempted in \cite{craddock2012}.
%They found this to be preferable interpretability can be sacrificed for accuracy.
%In short, by clustering adjacency patterns of functional connectivity maps, it is also possible to obtain precise and functionally homogenous parcels within human brains.

\section{Methods \& Strategy}
%Short general introduction to both methods

% Licht hier toe met welke methoden we gaan vergelijken en wat we nog meer besproken hebben in de outline.
% - Difficulties in causal discovery
% - Comparison methods
% - Comparison procedure
% - Properties

\section{Discussion}
% Short recap on:
% - what will be covered
% - what will not be covered
% Short discussion about proposal and methods covered.

\section{Time schedule}

\bibliography{references}{}
\addcontentsline{toc}{section}{References}
\bibliographystyle{apalike}

\end{document}

### 2013-09-24 docs/proposal/doc.tex
Added effective connectivity
Made a start to the motivation.

@@
 
+% vergelijking met granger causality, en andere methodes?
+
 % NOTE: De delen hieronder zijn letterlijk overgenomen uit een eerder werk.
@@
 % Motivation
+The brain and particularly the human brain have been studied for hundreds of years.
+Today, the secrets of our brains still belong to the most sought-after secrets.
+The techniques however have changed.
+Today, advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without having to open a single skull.
+
+Of particular interest are causal relations between brain regions and cognitive and bodily functions.
+We have come a long way since the time of Phrenology, the pseudoscience based on the idea that cognitive ability and personality could be derived from measurements of the skull.
+The question whether brain region X is somehow connected to brain region Y is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
+
+Understanding brain structure implies understanding more about the brain as a whole.
+
+Brain connectivity also finds applications in medicine.
+Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity.
+Better methods for analysing connectivity could lead to more insight in these diseases.
+
 %This is part of why it is interesting to investigate neuro-degenerative disease, which are able to cripple our usually robust networks effectively.
@@
 \subsection*{Brain connectivity}
-% Functional connectivity
+%Functional connectivity
 %Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.

### 2013-09-24 docs/proposal/doc.tex
merge

@@
 % vergelijking met granger causality, en andere methodes?
+% granger (Alard Roebroeck, Elia Formisano, Rainer Goebel, 2005) -> geen structural connectivity nodig (schaalt niet goed)
+% covariance structural equation modeling -> structural connectivity nodig (Roebroeck, Formisano, Goebel, 2005)
+% Multivariate autoregressive modeling (Harrison, Penny, Friston) -> Set of interaccting regions chosen beforehand (Roebroeck, Formisano, Goebel, 2005)
+% Dynamic causal modelling (Friston, Harrison, Penny, 2003) -> Set of interaccting regions chosen beforehand (Roebroeck, Formisano, Goebel, 2005)
 
@@
 
+Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
+Research has also been done on inferring effective connectivity, based on Granger causality (Roebroeck, Formisano, Goebel, 2005), multivariate autoregressive modeling (Harrison, Penny, Friston, 2003) and dynamic causal modelling (Friston, Harrison, Penny, 2003).
+However, many challenges are still being faced. (Ramsey, Hanson, Hanson, Halchenko, Poldrack , Glymour)
+
 %Brain connectivity analysis can teach us a lot of how the human brain functions.
@@
 % Effective connectivity
-% ...
-
-\subsection*{Bayesian connectomics}
-% ...
+%In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
+%It has been described as "the influence one neural system exerts over another" \cite{friston1994".
+%Effective connectivity indicates which brain regions stimulate other regions.
+%Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
+%It is possible to infer effective connectivity from structural and functional connectivity.
 
@@
 
+The causal discovery method which we will use is the PC-algorithm (spirtes, 2000). An advantage of this algorithm is that the scaling is relatively well.
+
 % Licht hier toe met welke methoden we gaan vergelijken en wat we nog meer besproken hebben in de outline.

### 2013-09-24 docs/proposal/doc.tex
references over andere causale methodes toegevoegd

--- 2013-09-24 docs/proposal/doc.tex
@@
 
+Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
+Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, multivariate autoregressive modeling \cite{Harrison2003} and dynamic causal modelling \cite{Friston2003}.
+However, many challenges are still being faced \cite{Ramsey2010}.
+One example of this is that not all approaches have a good schaling, and as such they cannot be applied to large datasets.
+Another problem is that not all methods can cope with latent variables, and wrong directive relations can be concluded.
+
 %Brain connectivity analysis can teach us a lot of how the human brain functions.
@@
 \subsection*{Brain connectivity}
-%Functional connectivity
-%Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.
+\subsubsection*{Functional connectivity}
+Functional
+%The Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.
 %An increasing body of studies successfully investigates functional connectivity by applying statistical methods on the level of co-activation of resting-state fMRI time-series between brain regions.
@@
 
-% Structural connectivity
+\subsubsection*{Structural connectivity}
+The structural connectivity is a mapping of the physical paths in the brain, and it shows the neural connections between anatomically separated brain regions.
+This is strongly related with functional connectivity: regions can only be functionally connected if there is a physical connection between them. \cite{vandenheuvel2009}
+
 %Similarly, there is the field of structural connectivity, which is defined as the mapping of direct anatomical brain pathways \cite{friston1994}.
@@
 
-% Effective connectivity
-% ...
-
-\subsection*{Bayesian connectomics}
-% ...
+\subsubsection*{Effective connectivity}
+In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
+It has been described as \"the influence one neural system exerts over another\" \cite{friston1994}.
+Effective connectivity indicates which brain regions stimulate other regions.
+Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
+It is possible to infer effective connectivity from structural and functional connectivity.
 

### 2013-09-24 docs/proposal/doc.tex
Further work on motivation.

--- 2013-09-24 docs/proposal/doc.tex
@@
 The brain and particularly the human brain have been studied for hundreds of years.
-Today, the secrets of our brains still belong to the most sought-after secrets.
+Today, the secrets of our brains still are one of the most sought-after.
 The techniques however have changed.
@@
 Of particular interest are causal relations between brain regions and cognitive and bodily functions.
-We have come a long way since the time of Phrenology, the pseudoscience based on the idea that cognitive ability and personality could be derived from measurements of the skull.
+We have come a long way since the time of Phrenology, the pseudoscience based on the idea that cognitive ability and personality could be derived from measurements of the skull and, post-mortem, the brain.
 The question whether brain region X is somehow connected to brain region Y is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
-
 Understanding brain structure implies understanding more about the brain as a whole.
 
-Brain connectivity also finds applications in medicine.
+Not surprisingly, brain connectivity also finds applications in medicine.
 Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity.
@@
 
-%This is part of why it is interesting to investigate neuro-degenerative disease, which are able to cripple our usually robust networks effectively.
-%We want to learn more about our brains, as well as the diseases that target them.
-
-Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
-Research has also been done on inferring effective connectivity, based on Granger causality (Roebroeck, Formisano, Goebel, 2005), multivariate autoregressive modeling (Harrison, Penny, Friston, 2003) and dynamic causal modelling (Friston, Harrison, Penny, 2003).
-However, many challenges are still being faced. (Ramsey, Hanson, Hanson, Halchenko, Poldrack , Glymour)
+The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
+One type of connectivity analysis involves in determining these causal relations between brain areas.
+However, causal analysis still faces many challenges that require solutions \cite{ramsey2010}.
 

### 2013-09-24 docs/proposal/doc.tex

--- 2013-09-24 docs/proposal/doc.tex
@@
 
-% NOTE: De delen hieronder zijn letterlijk overgenomen uit een eerder werk.
-Ik weet niet of we het nog nodig hebben, daarom heb ik het laten staan.
-We moeten het dan wel veranderen, anders pleeg ik plagiaat op mezelf.
+% Introductory small-talk
+% ...
 
-% Introductory small-talk
-%The human brain is a big network which we have yet to fully understand.
-%Different areas of this network each have their own task and function, but complexity also arises from how these are connected.
-%Our brains are a robust piece of equipment, presumably also a consequence of how different areas are interconnected, capable of adapting to continuous environmental change and learning throughout our entire lives.
-
-% Motivation
+\paragraph{Motivation}
 The brain and particularly the human brain have been studied for hundreds of years.
-Today, the secrets of our brains still belong to the most sought-after secrets.
+Today, the secrets of our brains still are one of the most sought-after.
 The techniques however have changed.
@@
 
-%This is part of why it is interesting to investigate neuro-degenerative disease, which are able to cripple our usually robust networks effectively.
-%We want to learn more about our brains, as well as the diseases that target them.
+The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
+One type of connectivity analysis involves finding causal relations between brain areas.
+However, causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{Ramsey2010}.
 
-Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
-Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, multivariate autoregressive modeling \cite{Harrison2003} and dynamic causal modelling \cite{Friston2003}.
-However, many challenges are still being faced \cite{Ramsey2010}.
-One example of this is that not all approaches have a good schaling, and as such they cannot be applied to large datasets.
-Another problem is that not all methods can cope with latent variables, and wrong directive relations can be concluded.
-
-%Brain connectivity analysis can teach us a lot of how the human brain functions.
-%The fields of functional and structural connectivity have, presumably for that reason, gained in popularity over the past few years \cite{vandenheuvel2010}.
-
-%Problem
-% ...
+\paragraph{Problem statement}
+This project seeks to combine the strengths of current state-of-the-art causal discovery methods and apply them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
 
@@
 In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
-It has been described as \"the influence one neural system exerts over another\" \cite{friston1994}.
+It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
 Effective connectivity indicates which brain regions stimulate other regions.
@@
 It is possible to infer effective connectivity from structural and functional connectivity.
+
+%Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
+%Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, multivariate autoregressive modeling \cite{Harrison2003} and dynamic causal modelling \cite{Friston2003}.
+%However, many challenges are still being faced \cite{Ramsey2010}.
+One example of this is that not all approaches have a good schaling, and as such they cannot be applied to large datasets.
+Another problem is that not all methods can cope with latent variables, and wrong directive relations can be concluded.
 
@@
 
-\section{Discussion}
-% Short recap on:
-% - what will be covered
-% - what will not be covered
-% Short discussion about proposal and methods covered.
-
 \section{Time schedule}

### 2013-09-24 docs/proposal/doc.tex
voortgang in stukje over connectivity

--- 2013-09-24 docs/proposal/doc.tex
@@
 \subsubsection*{Functional connectivity}
-Functional
+The functional connectivity describes the statistical dependance of neuronal activity of different brain regions \cite{friston1993functional}.
+This gives an insight in the organization of the brain, even though the exact functioning is still unclear \cite{Lowe2000, doria2010, Bullmore2009}.
+Functional connectivity is strongly time-dependant, as these dependancies may be short-lasting.
+
 %The Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.
@@
 The structural connectivity is a mapping of the physical paths in the brain, and it shows the neural connections between anatomically separated brain regions.
-This is strongly related with functional connectivity: regions can only be functionally connected if there is a physical connection between them. \cite{vandenheuvel2009}
+This is strongly related with functional connectivity: regions can only be functionally connected if there is a physical connection between them.
+This is an intuitive relation, and is also shown experimentally \cite{vandenheuvel2009}.
 

### 2013-09-24 docs/proposal/doc.tex
begin BCCD gemaakt

@@
 The functional connectivity describes the statistical dependance of neuronal activity of different brain regions \cite{friston1993functional}.
-This gives an insight in the organization of the brain, even though the exact functioning is still unclear \cite{Lowe2000, doria2010, Bullmore2009}.
+This gives an insight in the organization of the brain, even though the exact functioning is still unclear.
 Functional connectivity is strongly time-dependant, as these dependancies may be short-lasting.
+The functional connectivity can be deduced from resting-state networks \cite{Lowe2000, doria2010, Bullmore2009}.
 
@@
 \subsection*{Causal discovery methods}
-% ...
+The causal discovery method which we will use is the BCCD-algorithm.
+This algorithm finds both the structure of a network as well as the direction of the connections of this network.
+An advantage of this algorithm is that the scaling is relatively good.
+BCCD also finds probabilities of connections.
 
@@
 
-The causal discovery method which we will use is the PC-algorithm (spirtes, 2000). An advantage of this algorithm is that the scaling is relatively well.
-
 % Licht hier toe met welke methoden we gaan vergelijken en wat we nog meer besproken hebben in de outline.

### 2013-09-24 docs/proposal/doc.tex
Added small part on problem statement

--- 2013-09-24 docs/proposal/doc.tex
@@
 The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
-One type of connectivity analysis involves finding causal relations between brain areas.
+One type of connectivity analysis involves finding causal relations between brain areas, specifically \emph{effective connectivity}.
 However, causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{Ramsey2010}.
@@
 This project seeks to combine the strengths of current state-of-the-art causal discovery methods and apply them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
+Several methods for finding effective connectivity already exist \cite{mcIntosh1994, Harrisson2003, Friston2003, Roebroeck2005}
+However using more generally applicable causal discovery methods has not been elaborated on much.
+
+We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or PC-algorithm introduced in \cite{spirtes2000}, can solve some of the difficulties effective connectivity suffers from.
+% What problems? See \cite{Ramsey2010}.
 
@@
 \subsubsection*{Functional connectivity}
-The functional connectivity describes the statistical dependance of neuronal activity of different brain regions \cite{friston1993functional}.
-This gives an insight in the organization of the brain, even though the exact functioning is still unclear \cite{Lowe2000, doria2010, Bullmore2009}.
-Functional connectivity is strongly time-dependant, as these dependancies may be short-lasting.
-
+Functional
 %The Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.
@@
 The structural connectivity is a mapping of the physical paths in the brain, and it shows the neural connections between anatomically separated brain regions.
-This is strongly related with functional connectivity: regions can only be functionally connected if there is a physical connection between them.
-This is an intuitive relation, and is also shown experimentally \cite{vandenheuvel2009}.
+This is strongly related with functional connectivity: regions can only be functionally connected if there is a physical connection between them. \cite{vandenheuvel2009}
 
@@
 %Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
-%Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, multivariate autoregressive modeling \cite{Harrison2003} and dynamic causal modelling \cite{Friston2003}.
+%Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, Structural equation modelling \cite{mclntosh1994}, multivariate autoregressive modeling \cite{Harrison2003} and dynamic causal modelling \cite{Friston2003}.
 %However, many challenges are still being faced \cite{Ramsey2010}.

### 2013-09-24 docs/proposal/doc.tex

@@
 \subsubsection*{Functional connectivity}
-Functional
+The functional connectivity describes the statistical dependance of neuronal activity of different brain regions \cite{friston1993functional}.
+This gives an insight in the organization of the brain, even though the exact functioning is still unclear.
+Functional connectivity is strongly time-dependant, as these dependancies may be short-lasting.
+The functional connectivity can be deduced from resting-state networks \cite{Lowe2000, doria2010, Bullmore2009}.
+
 %The Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.

### 2013-09-25 docs/proposal/doc.tex
Added part to problem statement on difficulties with fMRI analysis.

--- 2013-09-24 docs/proposal/doc.tex
@@
 % Title
-\title{\fontfamily{phv}\selectfont{Research A Proposal}}
+\title{\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity}}
 \author{
@@
 The techniques however have changed.
-Today, advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without having to open a single skull.
+Today, advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
 
 Of particular interest are causal relations between brain regions and cognitive and bodily functions.
-We have come a long way since the time of Phrenology, the pseudoscience based on the idea that cognitive ability and personality could be derived from measurements of the skull and, post-mortem, the brain.
+We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
 The question whether brain region X is somehow connected to brain region Y is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
@@
 Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity.
+%TODO: Citation needed.
 Better methods for analysing connectivity could lead to more insight in these diseases.
@@
 The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
-One type of connectivity analysis involves finding causal relations between brain areas.
-However, causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{Ramsey2010}.
+One type of connectivity analysis involves finding causal relations between brain areas, specifically \emph{effective connectivity}.
+However, causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{ramsey2010}.
 
@@
 
+The work by \cite{ramsey2010} addresses several problems causal analysis in fMRI analysis suffers from.
+Methods in causal discovery have high computational bounds, making it nearly impossible to calculate networks with more than hundreds of nodes.
+Another import fact is fMRI Analysis indirectly measures brain activity.
+Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise or fail to capture underlying patterns.
+Another problem is the strong diversity between measurement subjects.
+Every brain is inherently different, making it non-trivial to combine subject data or even draw conclusions across subjects.
+Even within a single brain, how different areas react over time changes.
+
+Several methods for finding effective connectivity already exist \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
+However using more generally applicable causal discovery methods has not been elaborated on much.
+Also, none of these provide a measure of uncertainty.
+It is a fact that there are errors in any graph produced.
+This is inherent to the methods and the coarseness of the measurements.
+As there is no standard baseline to compare results with, a measure of uncertainty would be highly preferable.
+A framework for such a probabilistic approach is introduced in \cite{claassen2012}.
+
+We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as PC-algorithm \cite{spirtes2000} can solve some of the difficulties effective connectivity suffers from.
+
 % Structure of the rest of proposal
-% ...
+In the remainder of this proposal we briefly discuss some necessary background knowledge on connectivity and methods, we then discuss our strategy and finally we present a time-plan overview.
 
@@
 %Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
-%Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, multivariate autoregressive modeling \cite{Harrison2003} and dynamic causal modelling \cite{Friston2003}.
-%However, many challenges are still being faced \cite{Ramsey2010}.
+%Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, Structural equation modelling \cite{mclntosh1994}, multivariate autoregressive modeling \cite{harrison2003} and dynamic causal modelling \cite{friston2003}.
+%However, many challenges are still being faced \cite{ramsey2010}.
 One example of this is that not all approaches have a good schaling, and as such they cannot be applied to large datasets.

### 2013-09-25 docs/proposal/doc.tex
stuk over algorithmes, data en strategy toegevoegd

--- 2013-09-24 docs/proposal/doc.tex
@@
 This algorithm finds both the structure of a network as well as the direction of the connections of this network.
-An advantage of this algorithm is that the scaling is relatively good.
-BCCD also finds probabilities of connections.
+This algorithm has some useful properties; it is based on a Bayesian approach and as such, it is robust.
+It also does not need any assumptions beforehand.
+Because of this, the algorithm performs at least as good as other causal discovery algorithms such as FCI and PC in benchmarks on representative datasets, and in some cases it performs better.
+BCCD also finds the reliability of found connections, giving additional information compared to PC and FCI.
 
 \subsection*{Data acquisition \& analysis}
-% ...
+To find effective connectivity, structural and functional connectivity are needed.
+The datasets on which the causal discovery methods will be applied are DTI and fMRI-measurements of six subjects.
+Structural connectivity of these subjects follows directly from DTI-measurements.
+The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity.
+This data is already available.
 
@@
 %Short general introduction to both methods
+In this project, causal discovery algorithms will be applied to a dataset of functional and structural connectivity, in order to find effective connectivity.
+For this, implementations of PC and BCCD can be used.
+PC is a widely used algorithm and as such, many implementations already exist.
+An implementation of BCCD suitable for this project also exists at the department of Intelligent Systems of the Radboud university.
+The datasets of structural and functional connectivity are available, so the first step of this step will be to get the implementations for the algorithms working on the dataset.
+This will result in a connectivity matrix which represents the effective connectivity.
+These results will  be evaluated by cross-subject comparison, to see whether the gained data is consistent.
+If it is, then this algorithm is a potentially useful algorithm for determining effective connectivity.
+As PC is a state-of-the-art algorighm \cite{?}, and BCCD shows a good potential \cite{}, this way of determining effective connectivity may very well be the case.
 

### 2013-09-25 docs/proposal/doc.tex
merge

@@
 If it is, then this algorithm is a potentially useful algorithm for determining effective connectivity.
-As PC is a state-of-the-art algorighm \cite{?}, and BCCD shows a good potential \cite{}, this way of determining effective connectivity may very well be the case.
+As PC is a state-of-the-art algorighm \cite{spirtes2000}, and BCCD shows a good potential \cite{claassen2012}, this way of determining effective connectivity may very well be the case.
 

### 2013-09-25 docs/proposal/doc.tex
Rewrote part on functional connectivity.

@@
 \subsubsection*{Functional connectivity}
-The functional connectivity describes the statistical dependance of neuronal activity of different brain regions \cite{friston1993functional}.
-This gives an insight in the organization of the brain, even though the exact functioning is still unclear.
-Functional connectivity is strongly time-dependant, as these dependancies may be short-lasting.
-The functional connectivity can be deduced from resting-state networks \cite{Lowe2000, doria2010, Bullmore2009}.
+Functional connectivity describes the statistical dependance of neuronal activity between different brain regions \cite{friston1993functional}.
+Functional connectivity can concisely be deduced by measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
+It gives an insight in the organisation of the brain, even though the exact functioning is still unclear.
+Functional connectivity is strongly time-dependant, as activity changes rapidly providing only a small time window to measure.
+This means measured dependencies can be the result of statistical noise.
 

### 2013-09-25 docs/proposal/doc.tex
Rewrote part on functional connectivity.
wijzigingen aan causal discovery methods / strategy

--- 2013-09-25 docs/proposal/doc.tex
@@
 \subsection*{Causal discovery methods}
-The causal discovery method which we will use is the BCCD-algorithm.
-This algorithm finds both the structure of a network as well as the direction of the connections of this network.
+The first causal method that we will use to find effective connectivity will be the PC-algorithm.
+This is a state of the art used for general causal discovery, which finds the structure and the direction of connections of this network.
+The structure of a network can also be given as input, for which the PC-algorithm can determine the directions.
+Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC.
 This algorithm has some useful properties; it is based on a Bayesian approach and as such, it is robust.
@@
 To find effective connectivity, structural and functional connectivity are needed.
-The datasets on which the causal discovery methods will be applied are DTI and fMRI-measurements of six subjects.
-Structural connectivity of these subjects follows directly from DTI-measurements.
-The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity.
-This data is already available.
+The datasets on which the causal discovery methods will be applied are DTI and fMRI-measurements of six subjects. Structural connectivity of these subjects follows directly from DTI-measurements.
+The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity. This data is already available.
 

### 2013-09-25 docs/proposal/doc.tex

--- 2013-09-25 docs/proposal/doc.tex
@@
 \subsection*{Brain connectivity}
+
+%Question: should this be included?
 \subsubsection*{Functional connectivity}
 Functional connectivity describes the statistical dependance of neuronal activity between different brain regions \cite{friston1993functional}.
-Functional connectivity can concisely be deduced by measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
 It gives an insight in the organisation of the brain, even though the exact functioning is still unclear.
@@
 This means measured dependencies can be the result of statistical noise.
-
-%The Functional connectivity is defined as the temporal dependence of neuronal activity patterns of anatomically separated brain regions \cite{friston1993functional}.
-%An increasing body of studies successfully investigates functional connectivity by applying statistical methods on the level of co-activation of resting-state fMRI time-series between brain regions.
-%Several studies have shown that resting-state networks reflect brain connectivity in a concise way, though their exact relation remains unknown \cite{Lowe2000, doria2010, Bullmore2009}.
-%Resting state indicating that subjects are instructed to relax without thinking of something in particular, to stimulate spontaneous brain activity.
+Functional connectivity can concisely be deduced by measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
+Resting state indicates that subjects are instructed to relax without thinking of anything in particular, to stimulate spontaneous brain activity.
 

### 2013-09-25 docs/proposal/doc.tex
Revised sections on structural and effective connectivity. Made a start with the section on Causal Discovery Methods.

@@
 
-% Introductory small-talk
-% ...
-
 \paragraph{Motivation}
@@
 \subsubsection*{Structural connectivity}
-The structural connectivity is a mapping of the physical paths in the brain, and it shows the neural connections between anatomically separated brain regions.
-This is strongly related with functional connectivity: regions can only be functionally connected if there is a physical connection between them.
-This is an intuitive relation, and is also shown experimentally \cite{vandenheuvel2009}.
-
-%Similarly, there is the field of structural connectivity, which is defined as the mapping of direct anatomical brain pathways \cite{friston1994}.
-%It refers to white matter tracts, describing the millions of long-distance axons or `brain highways', connecting large groups of spatially separated brain areas.
-%In accordance with intuition, a recent study has shown that both fields are related.
-%By demonstrating that almost all functionally linked regions of the most reported resting-state networks are also structurally connected by known white matter tracts \cite{vandenheuvel2009}.
-%Also, functional connectivity is constrained by structural connectivity \cite{cabral2012}.
-%Structural connectivity is less time dependent, as it involves the mapping of anatomical connections opposing activity patterns.
+Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain, between anatomically separated brain regions \cite{friston1994}.
+This is strongly related with functional connectivity: regions can only be structurally connected if there is a functional relation between them.
+This is intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
+The other way around it can be demonstrated that functional connectivity is constrained by structural connectivity \cite{cabral2012}.
+Structural connectivity is less time dependent as it involves mappings of anatomical connections opposing temporal activity patterns.
 
@@
 Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
-It is possible to infer effective connectivity from structural and functional connectivity.
+It is possible to infer effective connectivity from structural and functional connectivity \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
 
-%Much progress has recently been made in the fields of structural and functional connectivity\cite{vandenheuvel2010}.
 %Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, Structural equation modelling \cite{mclntosh1994}, multivariate autoregressive modeling \cite{harrison2003} and dynamic causal modelling \cite{friston2003}.
 %However, many challenges are still being faced \cite{ramsey2010}.
-One example of this is that not all approaches have a good schaling, and as such they cannot be applied to large datasets.
+One example of this is that not all approaches have a good scaling, and as such they cannot be applied to large datasets.
 Another problem is that not all methods can cope with latent variables, and wrong directive relations can be concluded.
@@
 \subsection*{Causal discovery methods}
-The first causal method that we will use to find effective connectivity will be the PC-algorithm.
-This is a state of the art used for general causal discovery, which finds the structure and the direction of connections of this network.
-The structure of a network can also be given as input, for which the PC-algorithm can determine the directions.
-Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC.
-This algorithm has some useful properties; it is based on a Bayesian approach and as such, it is robust.
-It also does not need any assumptions beforehand.
-Because of this, the algorithm performs at least as good as other causal discovery algorithms such as FCI and PC in benchmarks on representative datasets, and in some cases it performs better.
-BCCD also finds the reliability of found connections, giving additional information compared to PC and FCI.
+The causal discovery method which we will use is the BCCD-algorithm.
+This algorithm finds both the structure of a network as well as the direction of the connections of this network.
+This algorithm has some useful properties; it is a Bayesian approach and as such, it is robust against noise.
+%It also does not need any assumptions beforehand.
+The authors have shown the method to perform as good as or better than other causal discovery algorithms such as FCI and PC on representative datasets.
+As is inherent to Bayesian methods, or approximations of Bayesian methods, the resulting network graph is not a  single structure but rather a distribution over possible structures.
+Concretely, this provides an uncertainty measure over edges in the network, giving additional information compared to PC and FCI \cite{claassen2012}.
 
@@
 To find effective connectivity, structural and functional connectivity are needed.
-The datasets on which the causal discovery methods will be applied are DTI and fMRI-measurements of six subjects. Structural connectivity of these subjects follows directly from DTI-measurements.
-The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity. This data is already available.
+The datasets on which the causal discovery methods will be applied are DTI and fMRI-measurements of six subjects.
+Structural connectivity of these subjects follows directly from DTI-measurements.
+The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity.
+This data is already available.
 

### 2013-09-25 docs/proposal/doc.tex
poging tot begin van gantt-chart

--- 2013-09-25 docs/proposal/doc.tex
@@
 \usepackage[noend]{algpseudocode}
+
+\usepackage{pgfgantt}
 
@@
 This is a state of the art used for general causal discovery, which finds the structure and the direction of connections of this network.
-The structure of a network can also be given as input, for which the PC-algorithm can determine the directions.
+If a graph skeleton of a network is already known beforehand, this can also be used as a basis such that the PC-algorithm can determine the directions.
 Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC.
@@
 An implementation of BCCD suitable for this project also exists at the department of Intelligent Systems of the Radboud university.
+
 The datasets of structural and functional connectivity are available, so the first step of this step will be to get the implementations for the algorithms working on the dataset.
+We will first try the PC-algorithm, for which a suitable implementation still needs to be found.
+This will be done by using only functional connectivity data and also by providing strucural connectivity data as a graph skeleton.
+The BCCD-algorithm will be used afterwards, which will also be used with only functional data as well as with functional and structural connectivity data.
 This will result in a connectivity matrix which represents the effective connectivity.
-These results will  be evaluated by cross-subject comparison, to see whether the gained data is consistent.
+These results will then be evaluated by cross-subject comparison, to see whether the gained data is consistent.
 If it is, then this algorithm is a potentially useful algorithm for determining effective connectivity.
@@
 
+% week 50 is deadline voor draft, 51 is presentaties, 2 inleveren van final
 \section{Time schedule}
+\begin{ganttchart}{1}{10}
+\gantttitle{2011}{10} \\
+\gantttitlelist{40,...,52,1,2}{1} \\
+\ganttgroup{Group 1}{1}{7} \\
+\ganttbar{Task 1}{1}{2} \\
+\ganttlinkedbar{Task 2}{3}{7} \ganttnewline
+\ganttmilestone{Milestone}{7} \ganttnewline
+\ganttbar{Final Task}{8}{12}
+\ganttlink{elem2}{elem3}
+\ganttlink{elem3}{elem4}
+\end{ganttchart}
 

### 2013-09-26 docs/proposal/doc.tex
time schedule toegevoegd, stukje aan strategy toegevoegd

--- 2013-09-25 docs/proposal/doc.tex
@@
 \usepackage[noend]{algpseudocode}
+
+% for coloring individual cells in a table
+\usepackage[table]{xcolor}
+%\usepackage{pgfgantt}
 
@@
 \subsection*{Causal discovery methods}
-The causal discovery method which we will use is the BCCD-algorithm.
-This algorithm finds both the structure of a network as well as the direction of the connections of this network.
+The first causal method that we will use to find effective connectivity will be the PC-algorithm.
+This is a state of the art used for general causal discovery, which finds the structure and the direction of connections of this network.
+If a graph skeleton of a network is already known beforehand, this can also be used as a basis such that the PC-algorithm can determine the directions.
+Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC.
 This algorithm has some useful properties; it is a Bayesian approach and as such, it is robust against noise.
-%It also does not need any assumptions beforehand.
 The authors have shown the method to perform as good as or better than other causal discovery algorithms such as FCI and PC on representative datasets.
 As is inherent to Bayesian methods, or approximations of Bayesian methods, the resulting network graph is not a  single structure but rather a distribution over possible structures.
-Concretely, this provides an uncertainty measure over edges in the network, giving additional information compared to PC and FCI \cite{claassen2012}.
+Concretely, this provides an uncertainty measure over edges in the network, giving additional information compared to PC and FCI.
 
@@
 An implementation of BCCD suitable for this project also exists at the department of Intelligent Systems of the Radboud university.
+
 The datasets of structural and functional connectivity are available, so the first step of this step will be to get the implementations for the algorithms working on the dataset.
+We will first try the PC-algorithm, for which a suitable implementation still needs to be found.
+This will be done by using only functional connectivity data and also by providing strucural connectivity data as a graph skeleton.
+The BCCD-algorithm will be used afterwards, which will also be used with only functional data as well as with functional and structural connectivity data.
 This will result in a connectivity matrix which represents the effective connectivity.
-These results will  be evaluated by cross-subject comparison, to see whether the gained data is consistent.
+These results will then be evaluated by cross-subject comparison, to see whether the gained data is consistent.
 If it is, then this algorithm is a potentially useful algorithm for determining effective connectivity.
 As PC is a state-of-the-art algorighm \cite{spirtes2000}, and BCCD shows a good potential \cite{claassen2012}, this way of determining effective connectivity may very well be the case.
+
+Two algorithms are used, and both are applied on two different forms of data: using either functional connectivity data only, or functional and structural connectivity data.
+Therefore four results will be obtained.
+If problems arise getting one of these results, we can still obtain enough data to draw useful conclusions.
 
@@
 
-\section{Time schedule}
+% week 50 is deadline voor draft, 51 is presentaties, 2 inleveren van final
+% gantt gaf wat problemen met packages die nodig zijn om te compileren
+% \section{Time schedule}
+% \begin{ganttchart}{1}{10}
+% \gantttitle{2011}{10} \\
+% \gantttitlelist{40,...,52,1,2}{1} \\
+% \ganttgroup{Group 1}{1}{7} \\
+% \ganttbar{Task 1}{1}{2} \\
+% \ganttlinkedbar{Task 2}{3}{7} \ganttnewline
+% \ganttmilestone{Milestone}{7} \ganttnewline
+% \ganttbar{Final Task}{8}{12}
+% \ganttlink{elem2}{elem3}
+% \ganttlink{elem3}{elem4}
+% \end{ganttchart}
+
+\section{time schedule}
+For the time schedule, this project has been divided in four main parts: applying the PC-algorithm, applying the BCCD-algorithm, analysing the data (including cross-subject comparison) and writing the article.
+The schedule for applying the algorithm has been subdivided into applying it with functional connectivity data only, and applying it to both functional and structural connectivity data.
+
+\begin{tabular}[ht]{c || c | c | c | c | c | c | c | c | c | c |}
+\setlength{\tabcolsep}{12pt}
+\renewcommand{\arraystretch}{1.3}
+week          & 41 & 42 &  43 &  44 &  45 &  46 &  47 &  48 &  49 &  50 \tabularnewline \hline
+PC            & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} find implementation} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} functional}  & \cellcolor[gray]{0.9} structural  & \multicolumn{4}{ |c| }{} & \tabularnewline \hline
+BCCD          & \multicolumn{4}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.85} functional} & \cellcolor[gray]{0.85} structural & \multicolumn{3}{ |c| }{} \tabularnewline \hline
+analysis      & \multicolumn{3}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} PC} & & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} BCCD} & \multicolumn{2}{ |c| }{} \tabularnewline \hline
+write article & \multicolumn{3}{ |c| }{} & \multicolumn{7}{ |c| }{\cellcolor[gray]{0.85}}  \tabularnewline \hline
+\end{tabular}
 

### 2013-09-26 docs/proposal/doc.tex
Small revisions to final sections. This version was sent for feedback.

--- 2013-09-25 docs/proposal/doc.tex
@@
 \usepackage{graphicx}
+\usepackage{stfloats}
 \usepackage{listings}
@@
 Functional connectivity describes the statistical dependance of neuronal activity between different brain regions \cite{friston1993functional}.
-It gives an insight in the organisation of the brain, even though the exact functioning is still unclear.
+It gives an insight in the organisation of the brain, even though functioning is not immediately visible.
 Functional connectivity is strongly time-dependant, as activity changes rapidly providing only a small time window to measure.
 This means measured dependencies can be the result of statistical noise.
-Functional connectivity can concisely be deduced by measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
-Resting state indicates that subjects are instructed to relax without thinking of anything in particular, to stimulate spontaneous brain activity.
+Functional connectivity can be deduced by fMRI measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
+Resting state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity.
 
 \subsubsection*{Structural connectivity}
-Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain, between anatomically separated brain regions \cite{friston1994}.
+Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
 This is strongly related with functional connectivity: regions can only be structurally connected if there is a functional relation between them.
@@
 
-%Research has also been done on inferring effective connectivity, based on Granger causality \cite{Roebroeck2005}, Structural equation modelling \cite{mclntosh1994}, multivariate autoregressive modeling \cite{harrison2003} and dynamic causal modelling \cite{friston2003}.
-%However, many challenges are still being faced \cite{ramsey2010}.
-One example of this is that not all approaches have a good scaling, and as such they cannot be applied to large datasets.
-Another problem is that not all methods can cope with latent variables, and wrong directive relations can be concluded.
+Research has been done on inferring effective connectivity directly with several methods.
+The most renowned methods are Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modeling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
+These are all methods dedicated to this problem and no attempt has been done from the viewpoint of causal discovery in graphs.
+These methods also suffer from similar problems of which perhaps the most important is scaling.
+Most of these methods can only handle tens of nodes in reasonable time.
 
 \subsection*{Causal discovery methods}
-The causal discovery method which we will use is the BCCD-algorithm.
-This algorithm finds both the structure of a network as well as the direction of the connections of this network.
-This algorithm has some useful properties; it is a Bayesian approach and as such, it is robust against noise.
-%It also does not need any assumptions beforehand.
-The authors have shown the method to perform as good as or better than other causal discovery algorithms such as FCI and PC on representative datasets.
+The first causal method we will apply on brain structure data will be the PC-algorithm \cite{spirtes2000}.
+It is used to find structure, as well as causal relations between vertices in directed graphs. 
+If we already provide a graph skeleton, PC-algorithm can be used to determine directions.
+An adaptation of PC that is better adapted to finding latent variables is Fast Causal Inference \cite{spirtes2000}, though this method does not scale well to tens of variables.
+
+Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC.
+This algorithm has some useful properties; it is a Bayesian approach and as such, it is more robust against noise.
+The authors have shown the method to perform as good as or better than other causal discovery algorithms such as PC on representative datasets.
 As is inherent to Bayesian methods, or approximations of Bayesian methods, the resulting network graph is not a  single structure but rather a distribution over possible structures.
-Concretely, this provides an uncertainty measure over edges in the network, giving additional information compared to PC and FCI \cite{claassen2012}.
+Concretely, this provides an uncertainty measure over edges in the network, giving additional information compared to PC.
+
+\begin{figure*}[bp]
+\begin{tabular}{c || c | c | c | c | c | c | c | c | c | c |}
+\setlength{\tabcolsep}{12pt}
+\renewcommand{\arraystretch}{1.3}
+Week          & 41 & 42 &  43 &  44 &  45 &  46 &  47 &  48 &  49 &  50 \tabularnewline \hline
+PC            & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Find implementation} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Structural}  & \cellcolor[gray]{0.9} Functional & \multicolumn{4}{ |c| }{} & \tabularnewline \hline
+BCCD          & \multicolumn{4}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.85} Structural} & \cellcolor[gray]{0.85} Functional & \multicolumn{3}{ |c| }{} \tabularnewline \hline
+Analysis      & \multicolumn{3}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} PC} & & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} BCCD} & \multicolumn{2}{ |c| }{} \tabularnewline \hline
+Write article & \multicolumn{3}{ |c| }{} & \multicolumn{7}{ |c| }{\cellcolor[gray]{0.85}}  \tabularnewline \hline
+\end{tabular}
+\caption{Project time table. `Structural' implies applying the given method on structural data only. `Functional' implies inferring structure with the given method first through functional data.}
+\end{figure*}
 
 \subsection*{Data acquisition \& analysis}
-To find effective connectivity, structural and functional connectivity are needed.
-The datasets on which the causal discovery methods will be applied are DTI and fMRI-measurements of six subjects.
+We will apply the mentioned methods on structural and functional data gathered from human subjects.
+We use the same functional and structural data as in Hinne et al. \cite{hinne2013}, of which we have selected the best 6 subjects based on MAP results.
+To find effective connectivity, both are needed.
 Structural connectivity of these subjects follows directly from DTI-measurements.
@@
 
-%Within these areas methods are applied on connectivity patterns obtained from fMRI analysis to predict anatomical connections.
-%It is possible to obtain anatomical connectivity patterns by applying partial correlation methods to resting-state fMRI data, filtering out poly-synaptic activity. %TODO: Reference needed.
-%Another option is to apply Diffusion Tensor Imaging (DTI) to resting-state data, which maps water diffusion directions in the brain, as this is a more direct way of charting brain highways.
-%TODO: Reference needed.
-
-%Both methods, i.e. correlation analysis and DTI, have downsides and ideally one would hope to combine them to have the best of both worlds.
-%% Note: de onderstaande uitspraak klopt dus niet voor ons proposal
-%Such a thing is possible when using Bayesian methods.
-In this proposal however, we focus on frequentistic or data driven methods, as they usually provide a computational advantage without a large performance trade-off.
-%However, this also limits the amount of brain modelling we can do.
-
-% Note: bekijk even of de onderstaande sectie Ÿberhaupt wel interessant is. Ik heb het gebruikt als introductie voor mijn formele uitleg. Die laatste ontbreekt nu natuurlijk.
-
-% Brain robustness and neuro-degeneration
-%\subsection*{Brain robustness and neuro-degeneration}
-%Investigating the brain as an integrative network of interacting regions has proven to be a fruitful platform for learning more about how the brain functions.
-%Its main premise is that the brain can be modelled as a graph $G=(\set{V},E)$, $\set{V}$ being a collection of brain areas and $E$ the connections between those.
-%It provides ideas for brain robustness and forms a basis for approaching certain neuro-degenerative diseases, e.g.
-Alzheimer's disease (AD), as these are known to affect brain connectivity \cite{Bullmore2009, vandenheuvel2010}.
-%Several studies have shown that the brain network is organised according to an efficient \emph{small-world} organization \cite{vandenheuvel2008, vandenheuvel2010}.
-%Some of these suggest that brain connectivity is distributed according to a power-law function, possibly explaining the brain's scale-free organisation of connectivity networks.
-%Scale-free networks are renowned for their robustness against random attacks and so it forms an interesting hypotheses for modelling brain networks.
-%Even more, since scale-free organisation is susceptible to specialised attacks on hub nodes, it could be so that these hub nodes are affected in AD.
-%This idea is based on resting state fMRI data and graph analysis, as well as MEG results \cite{buckner2009, stam2009}
-
-% Note: voor onderstaande tekst geldt: kijk even kritisch of we het nodig hebben.
-
-%\subsection{Analysing connectivity}
-%Where brain data is usually available in abundant amounts, testing subjects are not.
-%This, and large cross-subject divergence makes it difficult to compare (diseased) brain area's between subjects and conclusions between studies.
-%To still pinpoint Regions of Interest (ROI) between subject, one can use so called labelling atlases, such as the Automatic Anatomical Labeling atlas, to map brain areas to known ROI's.
-%Another option is to apply a custom clustering to brain data such as attempted in \cite{craddock2012}.
-%They found this to be preferable interpretability can be sacrificed for accuracy.
-%In short, by clustering adjacency patterns of functional connectivity maps, it is also possible to obtain precise and functionally homogenous parcels within human brains.
-
 \section{Methods \& Strategy}
 %Short general introduction to both methods
-In this project, causal discovery algorithms will be applied to a dataset of functional and structural connectivity, in order to find effective connectivity.
-For this, implementations of PC and BCCD can be used.
-PC is a widely used algorithm and as such, many implementations already exist.
-An implementation of BCCD suitable for this project also exists at the department of Intelligent Systems of the Radboud university.
-The datasets of structural and functional connectivity are available, so the first step of this step will be to get the implementations for the algorithms working on the dataset.
-This will result in a connectivity matrix which represents the effective connectivity.
-These results will  be evaluated by cross-subject comparison, to see whether the gained data is consistent.
-If it is, then this algorithm is a potentially useful algorithm for determining effective connectivity.
-As PC is a state-of-the-art algorighm \cite{spirtes2000}, and BCCD shows a good potential \cite{claassen2012}, this way of determining effective connectivity may very well be the case.
+We use existing implementations of PC and BCCD to infer effective connectivity.
+An implementation of BCCD suitable for this project is made available by the department of Intelligent Systems at the Radboud University Nijmegen.
+
+The first step of this step will be to get the implementations for the algorithms working on the datasets with functional and structural data.
+
+Firstly we will use the PC-algorithm on given structural networks to infer causal relations.
+Later we also infer network structure with the PC-algorithm to compare with the previous.
+A suitable implementation for PC-algorithm has to be found and selected.
+
+The second step is to apply the BCCD-algorithm, also to be used with and without provided network structure.
+Both methods provide a connectivity matrix representing the effective connectivity.
+
+Results will then be evaluated and compared over all six subjects to see whether gained data is consistent.
+If so, we will analyse the properties and difficulties presented by Ramsey et al. \cite{ramsey2010}.
+We will conclude whether standard causal discovery methods provide a good direction of research for deriving and evaluating effective connectivity.
+In any case, there is enough data available to achieve some useful results to draw conclusions.
 
@@
 
-\section{Time schedule}
+% week 50 is deadline voor draft, 51 is presentaties, 2 inleveren van final
+% gantt gaf wat problemen met packages die nodig zijn om te compileren
+% \section{Time schedule}
+% \begin{ganttchart}{1}{10}
+% \gantttitle{2011}{10} \\
+% \gantttitlelist{40,...,52,1,2}{1} \\
+% \ganttgroup{Group 1}{1}{7} \\
+% \ganttbar{Task 1}{1}{2} \\
+% \ganttlinkedbar{Task 2}{3}{7} \ganttnewline
+% \ganttmilestone{Milestone}{7} \ganttnewline
+% \ganttbar{Final Task}{8}{12}
+% \ganttlink{elem2}{elem3}
+% \ganttlink{elem3}{elem4}
+% \end{ganttchart}
+
+\section{time schedule}
+This project is divided in four main parts: applying the PC-algorithm, applying the BCCD-algorithm, analysing the data (including cross-subject comparison) and writing the article.
+The schedule for applying the algorithm has been subdivided into applying it with functional connectivity data only, and applying it to both functional and structural connectivity data.
 

### 2013-09-27 docs/proposal/doc.tex
begin van verwerken feedback

@@
 
-Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC.
+Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC, and is designed in an attempt to improve PC.
 This algorithm has some useful properties; it is a Bayesian approach and as such, it is more robust against noise.
@@
 Week          & 41 & 42 &  43 &  44 &  45 &  46 &  47 &  48 &  49 &  50 \tabularnewline \hline
-PC            & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Find implementation} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Structural}  & \cellcolor[gray]{0.9} Functional & \multicolumn{4}{ |c| }{} & \tabularnewline \hline
+PC            & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Make implementation} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Structural}  & \cellcolor[gray]{0.9} Functional & \multicolumn{4}{ |c| }{} & \tabularnewline \hline
 BCCD          & \multicolumn{4}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.85} Structural} & \cellcolor[gray]{0.85} Functional & \multicolumn{3}{ |c| }{} \tabularnewline \hline
@@
 To find effective connectivity, both are needed.
-Structural connectivity of these subjects follows directly from DTI-measurements.
+Structural connectivity of these subjects follows from DTI-measurements.
 The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity.
@@
 
-The first step of this step will be to get the implementations for the algorithms working on the datasets with functional and structural data.
+The first step will be to get the implementations for the algorithms working on the datasets with functional and structural data.
 
-Firstly we will use the PC-algorithm on given structural networks to infer causal relations.
-Later we also infer network structure with the PC-algorithm to compare with the previous.
-A suitable implementation for PC-algorithm has to be found and selected.
+Firstly we will use structural networks as a basis for the algorithms, and use functional information to infer causal relations. 
+Later we will use functional information only, to infer both the network and the causal relations.
+We will compare this with the previous.
+As PC is a well-described algorithm \cite{spirtes2000} and should not be difficult to implement, so this will be done within this project.
 
 The second step is to apply the BCCD-algorithm, also to be used with and without provided network structure.
-Both methods provide a connectivity matrix representing the effective connectivity.
+Both methods provide a connectivity matrix representing effective connectivity.
 
-Results will then be evaluated and compared over all six subjects to see whether gained data is consistent.
+Results will then be evaluated and compared over all six subjects to see to which extent obtained networks are consistent across subjects.
 If so, we will analyse the properties and difficulties presented by Ramsey et al. \cite{ramsey2010}.
 We will conclude whether standard causal discovery methods provide a good direction of research for deriving and evaluating effective connectivity.
-In any case, there is enough data available to achieve some useful results to draw conclusions.
 
@@
 
-% week 50 is deadline voor draft, 51 is presentaties, 2 inleveren van final
-% gantt gaf wat problemen met packages die nodig zijn om te compileren
-% \section{Time schedule}
-% \begin{ganttchart}{1}{10}
-% \gantttitle{2011}{10} \\
-% \gantttitlelist{40,...,52,1,2}{1} \\
-% \ganttgroup{Group 1}{1}{7} \\
-% \ganttbar{Task 1}{1}{2} \\
-% \ganttlinkedbar{Task 2}{3}{7} \ganttnewline
-% \ganttmilestone{Milestone}{7} \ganttnewline
-% \ganttbar{Final Task}{8}{12}
-% \ganttlink{elem2}{elem3}
-% \ganttlink{elem3}{elem4}
-% \end{ganttchart}
-
-\section{time schedule}
-This project is divided in four main parts: applying the PC-algorithm, applying the BCCD-algorithm, analysing the data (including cross-subject comparison) and writing the article.
+\section{Time schedule}
+This project is divided into four main parts: applying the PC-algorithm, applying the BCCD-algorithm, analysing the data (including cross-subject comparison) and writing the article.
 The schedule for applying the algorithm has been subdivided into applying it with functional connectivity data only, and applying it to both functional and structural connectivity data.

### 2013-09-27 docs/proposal/doc.tex
Changed title

@@
 \author{
-  \textbf{Ramon Janssen} - \href{mailto:ramon.janssen@student.ru.nl}{ramon.janssen@student.ru.nl} \\
-  \textbf{Tom de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}
+  \textbf{R. Janssen} - \href{mailto:ramon.janssen@student.ru.nl}{ramon.janssen@student.ru.nl} \\
+  \textbf{T. de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}\\
+  \textbf{T. Claassen} - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
+  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}\\
 }

### 2013-09-27 docs/proposal/doc.tex
Changed text according to feedback. Merge time.

@@
 \newcommand{\pseudoinv}{^{+}}
+
+% ### Hyphenation ###
+\hyphenation{a-na-ly-sis}
+
 % ############## End Macros ##############
@@
   \textbf{T. Claassen} - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
-  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}\\
+  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
 }
@@
 The techniques however have changed.
+We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
 Today, advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
@@
 Of particular interest are causal relations between brain regions and cognitive and bodily functions.
-We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
 The question whether brain region X is somehow connected to brain region Y is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
@@
 Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise or fail to capture underlying patterns.
-Another problem is the strong diversity between measurement subjects.
+Solutions to this particular problem have been proposed that do account for latent sources \cite{ramsey2010, waldorp2011}.
+Another problem is the strong diversity between subjects.
 Every brain is inherently different, making it non-trivial to combine subject data or even draw conclusions across subjects.
-Even within a single brain, how different areas react over time changes.
+Even within a single brain it changes how different areas react over time.
 
@@
 Structural connectivity of these subjects follows from DTI-measurements.
-The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity.
+Functional connectivity can be extracted from fMRI-measurement, possibly in combination with the structural connectivity.
 This data is already available.

### 2013-09-27 docs/proposal/doc.tex
al het commentaar van Max verwerkt

--- 2013-09-27 docs/proposal/doc.tex
@@
 The techniques however have changed.
-Today, advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
+Advanced measurement methods now exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
 
@@
 The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
-One type of connectivity analysis involves finding causal relations between brain areas, specifically \emph{effective connectivity}.
+One type of connectivity analysis involves finding causal relations between brain areas, specifically \emph{effective connectivity}; How does the activity of one region cause the activity of another?
 However, causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{ramsey2010}.
@@
 \paragraph{Problem statement}
-This project seeks to combine the strengths of current state-of-the-art causal discovery methods and apply them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
+This project seeks to utilize the strengths of current state-of-the-art causal discovery methods by applying them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
 
 The work by \cite{ramsey2010} addresses several problems causal analysis in fMRI analysis suffers from.
-Methods in causal discovery have high computational bounds, making it nearly impossible to calculate networks with more than hundreds of nodes.
-Another import fact is fMRI Analysis indirectly measures brain activity.
+Methods in causal discovery have high computational costs, making it nearly impossible to calculate networks with more than hundreds of nodes.
+Another important fact is that fMRI Analysis indirectly measures brain activity.
 Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise or fail to capture underlying patterns.
@@
 Several methods for finding effective connectivity already exist \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
-However using more generally applicable causal discovery methods has not been elaborated on much.
+However, using more generally applicable causal discovery methods has not been elaborated on much.
 Also, none of these provide a measure of uncertainty.
@@
 
-We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as PC-algorithm \cite{spirtes2000} can solve some of the difficulties effective connectivity suffers from.
+We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000} can solve some of the difficulties effective connectivity suffers from.
 
@@
 \subsubsection*{Functional connectivity}
-Functional connectivity describes the statistical dependance of neuronal activity between different brain regions \cite{friston1993functional}.
-It gives an insight in the organisation of the brain, even though functioning is not immediately visible.
-Functional connectivity is strongly time-dependant, as activity changes rapidly providing only a small time window to measure.
+Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
+This gives an insight in the organisation of the brain.
+Functional connectivity is strongly time-dependent, as activity changes rapidly providing only a low emporal resultion.
 This means measured dependencies can be the result of statistical noise.
@@
 Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
-This is strongly related with functional connectivity: regions can only be structurally connected if there is a functional relation between them.
-This is intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
-The other way around it can be demonstrated that functional connectivity is constrained by structural connectivity \cite{cabral2012}.
+This is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
+This intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
 Structural connectivity is less time dependent as it involves mappings of anatomical connections opposing temporal activity patterns.
@@
 The most renowned methods are Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modeling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
-These are all methods dedicated to this problem and no attempt has been done from the viewpoint of causal discovery in graphs.
+These are all methods dedicated to this problem.
 These methods also suffer from similar problems of which perhaps the most important is scaling.
-Most of these methods can only handle tens of nodes in reasonable time.
+Most of these methods can only handle tens of nodes in reasonable time, which becomes problematic in realistic situations with up to tens of thousands of nodes, ideally.
 
@@
 \end{tabular}
-\caption{Project time table. `Structural' implies applying the given method on structural data only. `Functional' implies inferring structure with the given method first through functional data.}
+\caption{Project time table.
+`Structural' implies applying the given method using structural data as a network skeleton.
+`Functional' implies inferring structure with the given method first through functional data.}
 \end{figure*}
@@
 We will apply the mentioned methods on structural and functional data gathered from human subjects.
-We use the same functional and structural data as in Hinne et al. \cite{hinne2013}, of which we have selected the best 6 subjects based on MAP results.
+We use the same functional and structural data as in Hinne et al. \cite{hinne2013}, of which six subjects were selected.
 To find effective connectivity, both are needed.
-Structural connectivity of these subjects follows from DTI-measurements.
+Structural connectivity of these subjects follows from DTI-measurements, using probabilistic tractography.
 The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity.
@@
 %Short general introduction to both methods
-We use existing implementations of PC and BCCD to infer effective connectivity.
+We use an existing implementation of BCCD to infer effective connectivity; PC will be implemented in this project.
 An implementation of BCCD suitable for this project is made available by the department of Intelligent Systems at the Radboud University Nijmegen.
@@
 This project is divided into four main parts: applying the PC-algorithm, applying the BCCD-algorithm, analysing the data (including cross-subject comparison) and writing the article.
-The schedule for applying the algorithm has been subdivided into applying it with functional connectivity data only, and applying it to both functional and structural connectivity data.
+The schedule for applying the algorithm has been subdivided into applying it with functional connectivity using structural data as a basis, and applying it to only functional data.
 

### 2013-09-27 docs/proposal/doc.tex

@@
 \author{
-  \textbf{Ramon Janssen} - \href{mailto:ramon.janssen@student.ru.nl}{ramon.janssen@student.ru.nl} \\
-  \textbf{Tom de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}
+  \textbf{R. Janssen} - \href{mailto:ramon.janssen@student.ru.nl}{ramon.janssen@student.ru.nl} \\
+  \textbf{T. de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}\\
+  \textbf{T. Claassen} - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
+  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
 }
@@
 The techniques however have changed.
-Advanced measurement methods now exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
+We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
+Advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
 
 Of particular interest are causal relations between brain regions and cognitive and bodily functions.
-We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
 The question whether brain region X is somehow connected to brain region Y is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
@@
 \paragraph{Problem statement}
-This project seeks to utilize the strengths of current state-of-the-art causal discovery methods by applying them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
+This project seeks to utilise the strengths of current state-of-the-art causal discovery methods by applying them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
 
@@
 To find effective connectivity, both are needed.
-Structural connectivity of these subjects follows from DTI-measurements, using probabilistic tractography.
-The functional connectivity can be extracted from these fMRI-measurement, possibly in combination with the structural connectivity.
-This data is already available.
+Structural connectivity of these subjects can be derived from DTI-measurements, using probabilistic tractography \cite{hinne2013}.
+Functional connectivity can be extracted from fMRI-measurement, possibly in combination with the structural 
+connectivity.
+Functional connectivity can be extracted from fMRI-measurement, possibly in combination with the structural connectivity.
 

### 2013-09-27 docs/proposal/doc.tex
abstract toegevoegd

@@
 \begin{abstract}
-Lorem Ipsum Dolor Sit Amet.
+In the field of brain research, different forms of connectivity are considered.
+To find effective connectivity, many different approaches have been used.
+In this project, we will apply causal discovery the methods of PC and BCCD to structural and functional connectivity, to infer effective connectivity.
+As data from multiple subjects will be used, evaluation of the result can show whether the method yields consistent results.
 \keywords{Keywords}
@@
 
-Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC, and is designed in an attempt to improve PC.
+Another causal discovery method is the Baysian Constraint-based Causal Discovery (BCCD) algorithm \cite{claassen2012}, which gives similar results as PC, and is designed in an attempt to improve PC.
 This algorithm has some useful properties; it is a Bayesian approach and as such, it is more robust against noise.

### 2013-09-27 docs/proposal/doc.tex
Processed Max's feedback.

--- 2013-09-27 docs/proposal/doc.tex
@@
 % Margins & Headers
-\marginsize{2.5cm}{2.5cm}{2cm}{1.5cm}
+\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
 \columnsep 0.4in
@@
 Of particular interest are causal relations between brain regions and cognitive and bodily functions.
-The question whether brain region X is somehow connected to brain region Y is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
+The question if brain region X is functionally or structurally connected to brain region Y and whether these are causally related is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
 Understanding brain structure implies understanding more about the brain as a whole.
@@
 The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
-One type of connectivity analysis involves finding causal relations between brain areas, specifically \emph{effective connectivity}; How does the activity of one region cause the activity of another?
-However, causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{ramsey2010}.
+One type of connectivity analysis involves finding how brain region X causes what region Y does. This is called \emph{effective connectivity}.
+More concretely: how does the measured fMRI signal, representing an underlying neuronal process, determine what other neuronal processes are activated.
+Causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{ramsey2010}.
 
@@
 Another important fact is that fMRI Analysis indirectly measures brain activity.
-Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise or fail to capture underlying patterns.
+Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise, fail to capture underlying patterns or draw erroneous conclusions.
 Solutions to this particular problem have been proposed that do account for latent sources \cite{ramsey2010, waldorp2011}.
@@
 This gives an insight in the organisation of the brain.
-Functional connectivity is strongly time-dependent, as activity changes rapidly providing only a low emporal resultion.
+Functional connectivity is strongly time-dependent, as activity changes rapidly providing and fMRI has a low temporal resolution.
 This means measured dependencies can be the result of statistical noise.
@@
 We will apply the mentioned methods on structural and functional data gathered from human subjects.
-We use the same functional and structural data as in Hinne et al. \cite{hinne2013}, of which six subjects were selected.
-To find effective connectivity, both are needed.
+We use the same functional and structural data as in Hinne et al. \cite{hinne2013}, of which six subjects were selected at random.
+To find effective connectivity, both data sources are needed.
 Structural connectivity of these subjects can be derived from DTI-measurements, using probabilistic tractography \cite{hinne2013}.

### 2013-09-27 docs/proposal/doc.tex
Added small section on derivation of structural and functional networks.

--- 2013-09-27 docs/proposal/doc.tex
@@
 \begin{abstract}
-Lorem Ipsum Dolor Sit Amet.
-\keywords{Keywords}
+We propose a new application for causal discovery methods on human brain networks in order to derive effective connectivity.
+The directive nature of effective brain networks suggest causal methods can be used.
+As current methods for effective connectivity suffer from computational difficulties and often cannot cope with latent variables, we propose an application of PC-algorithm and the Bayesian BCCD-algorithm \cite{claassen2012} to estimate effective connectivity.
+Both methods will be applied on six healthy subjects of which fMRI data and structural networks are readily available.
+\keywords{Effective connectivity, Causal Discovery}
 \end{abstract}
@@
 More concretely: how does the measured fMRI signal, representing an underlying neuronal process, determine what other neuronal processes are activated.
-Causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{ramsey2010}.
+Research regarding effective is relatively new and its application still faces many challenges that require further research \cite{ramsey2010}.
+The directed character in effective connectivity seems similar to traditional causal relations.
+In what follows, we suggest to use causal discovery methods for deriving effective connectivity.
 
@@
 
+Hinne et al. \cite{hinne2013} proposed a Bayesian method for estimating structural networks based on Diffusion Weighted MRI (DWI) and probabilistic Tractography.
+The location of white-matter tracts does not become immediately clear from imagery alone.
+Other methods rely on thresholds to estimate location of said tracts.
+The proposed method results in a measure of uncertainty about where a hypothesised connection will terminate and so provides a clearly interpretable network structure as result.
+
+\begin{figure*}[bp]
+\begin{tabular}{c || c | c | c | c | c | c | c | c | c | c |}
+\setlength{\tabcolsep}{12pt}
+\renewcommand{\arraystretch}{1.3}
+Week          & 41 & 42 &  43 &  44 &  45 &  46 &  47 &  48 &  49 &  50 \tabularnewline \hline
+PC            & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Make implementation} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Structural}  & \cellcolor[gray]{0.9} Functional & \multicolumn{4}{ |c| }{} & \tabularnewline \hline
+BCCD          & \multicolumn{4}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.85} Structural} & \cellcolor[gray]{0.85} Functional & \multicolumn{3}{ |c| }{} \tabularnewline \hline
+Analysis      & \multicolumn{3}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} PC} & & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} BCCD} & \multicolumn{2}{ |c| }{} \tabularnewline \hline
+Write article & \multicolumn{3}{ |c| }{} & \multicolumn{7}{ |c| }{\cellcolor[gray]{0.85}}  \tabularnewline \hline
+\end{tabular}
+\caption{Project time table.
+`Structural' implies applying the given method using structural data as a network skeleton.
+`Functional' implies inferring structure with the given method first through functional data.}
+\end{figure*}
+
 \subsubsection*{Effective connectivity}
@@
 
-\begin{figure*}[bp]
-\begin{tabular}{c || c | c | c | c | c | c | c | c | c | c |}
-\setlength{\tabcolsep}{12pt}
-\renewcommand{\arraystretch}{1.3}
-Week          & 41 & 42 &  43 &  44 &  45 &  46 &  47 &  48 &  49 &  50 \tabularnewline \hline
-PC            & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Make implementation} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Structural}  & \cellcolor[gray]{0.9} Functional & \multicolumn{4}{ |c| }{} & \tabularnewline \hline
-BCCD          & \multicolumn{4}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.85} Structural} & \cellcolor[gray]{0.85} Functional & \multicolumn{3}{ |c| }{} \tabularnewline \hline
-Analysis      & \multicolumn{3}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} PC} & & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} BCCD} & \multicolumn{2}{ |c| }{} \tabularnewline \hline
-Write article & \multicolumn{3}{ |c| }{} & \multicolumn{7}{ |c| }{\cellcolor[gray]{0.85}}  \tabularnewline \hline
-\end{tabular}
-\caption{Project time table.
-`Structural' implies applying the given method using structural data as a network skeleton.
-`Functional' implies inferring structure with the given method first through functional data.}
-\end{figure*}
-
 \subsection*{Data acquisition \& analysis}
 We will apply the mentioned methods on structural and functional data gathered from human subjects.
-We use the same functional and structural data as in Hinne et al. \cite{hinne2013}, of which six subjects were selected at random.
-To find effective connectivity, both data sources are needed.
-Structural connectivity of these subjects can be derived from DTI-measurements, using probabilistic tractography \cite{hinne2013}.
-Functional connectivity can be extracted from fMRI-measurement, possibly in combination with the structural 
-connectivity.
-Functional connectivity can be extracted from fMRI-measurement, possibly in combination with the structural connectivity.
+For this we are largely dependent on data collected by others.
+Hinne et al. has provided fMRI time-series data and structural network data as used and derived in \cite{hinne2013}, of which six subjects were selected at random.
+We can estimate functional connectivity from fMRI-measurement in combination with structural 
+networks according to a procedure currently submitted by Hinne et al. \cite{hinne2013submitted}.
+To find effective connectivity, both structural networks and functional data is needed.
 

### 2013-11-26 docs/paper/paper.tex
tex-file voor paper aangemaakt

--- 2013-09-27 docs/proposal/doc.tex
@@
 
-\maketitle
+\section{Causal discovery}
+An approach for finding effective connectivity may be found in the domain of causal discovery.
+A causal discovery method uses data about events to determine which of these events are correlated, and if possible, it determines which event causes which.
+The definition of a causal relation is not trivial \cite{?} and the relevant principles will be covered in this section.
+In our context, the causal discovery methods which have been used aim at inferring a Bayesian network.
+This is done by calculating statistical (in)dependencies between stochastic variables.
 
-\begin{abstract}
-Lorem Ipsum Dolor Sit Amet.
-\keywords{Keywords}
-\end{abstract}
+Such a Bayesian network can be modelled as a graph with directed edges.
+Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
+A causal relation between variables $X$ and $Y$ is denoted with a directed edge; $X \rightarrow Y$ implies that $X$ is a cause of $Y$.
+It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$.
+A Bayesian network cannot contain cycles, as that would imply that a variable is it's own (indirect) parent and thus it's own cause.
+Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
+In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
+As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
+Image {plaatje!} shows an example of a Bayesian network.
+A causal discovery method attempts at finding a real-world Bayesian network.
+But real-world situations like a brain are immensely complex and contain an unimaginable large number of atoms, which could all function as seperate variables.
+As such, we represent our real-world, underlying Bayesian network with a much simpler model.
+Such a model contains a reduced number of variables to be able to analyse and visualize the data.
 
-\setlength{\parindent}{0.0cm}
-\setlength{\parskip}{0.25cm}
+An important principle for our approach is that of conditional dependence.
+The dependency of two variables might depend on the value of a third variable (or possibly even more variables).
+This principle is best demonstrated with an example.
+When two variables, such as two switches being turned off or on, both cause a light bulb to glow, the two switches might not be dependent.
+Whether one switch is turned on or off does not cause the other to switch on or off.
+However, when the light bulb is glowing, the switches do become dependent.
+If one switch is off, the other one is probably on.
+Or in other words, one switch being turned on decreases the chance of the other switch being turned on.
+In this case, we conclude that the switches are conditionally dependent given that the light bulb is turned on.
+More generally, two variables $X$ and $Y$ can be conditionally dependent on a set of other variables $S$. % het juiste latex-symbooltje vinden :(
 
-\section{Introduction}
+A similar principle is conditional independence. If two variables are
 
-% vergelijking met granger causality, en andere methodes?
-% granger (Alard Roebroeck, Elia Formisano, Rainer Goebel, 2005) -> geen structural connectivity nodig (schaalt niet goed)
-% covariance structural equation modeling -> structural connectivity nodig (Roebroeck, Formisano, Goebel, 2005)
-% Multivariate autoregressive modeling (Harrison, Penny, Friston) -> Set of interaccting regions chosen beforehand (Roebroeck, Formisano, Goebel, 2005)
-% Dynamic causal modelling (Friston, Harrison, Penny, 2003) -> Set of interaccting regions chosen beforehand (Roebroeck, Formisano, Goebel, 2005)
+Whether there is a dependency between two variables can be concluded by using a statistical test.
+In this way, a graph can be inferred representing all dependencies between nodes.
+However, this only results in an undirected graph, as it does not give any information about causality.
+To derive causal relations, we can make use of conditional dependence.
+Three rules for finding causality are used in the PC-algorithm.
+One of those rules is based on conditional dependency
 
-\paragraph{Motivation}
-The brain and particularly the human brain have been studied for hundreds of years.
-Today, the secrets of our brains still are one of the most sought-after.
-The techniques however have changed.
-We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
-Advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
-
-Of particular interest are causal relations between brain regions and cognitive and bodily functions.
-The question if brain region X is functionally or structurally connected to brain region Y and whether these are causally related is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
-Understanding brain structure implies understanding more about the brain as a whole.
-
-Not surprisingly, brain connectivity also finds applications in medicine.
-Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity.
-%TODO: Citation needed.
-Better methods for analysing connectivity could lead to more insight in these diseases.
-
-The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
-One type of connectivity analysis involves finding how brain region X causes what region Y does. This is called \emph{effective connectivity}.
-More concretely: how does the measured fMRI signal, representing an underlying neuronal process, determine what other neuronal processes are activated.
-Causal analysis is also relatively new and its application in brain connectivity analysis still faces many challenges that require further research \cite{ramsey2010}.
-
-\paragraph{Problem statement}
-This project seeks to utilise the strengths of current state-of-the-art causal discovery methods by applying them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
-
-The work by \cite{ramsey2010} addresses several problems causal analysis in fMRI analysis suffers from.
-Methods in causal discovery have high computational costs, making it nearly impossible to calculate networks with more than hundreds of nodes.
-Another important fact is that fMRI Analysis indirectly measures brain activity.
-Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise, fail to capture underlying patterns or draw erroneous conclusions.
-Solutions to this particular problem have been proposed that do account for latent sources \cite{ramsey2010, waldorp2011}.
-Another problem is the strong diversity between subjects.
-Every brain is inherently different, making it non-trivial to combine subject data or even draw conclusions across subjects.
-Even within a single brain it changes how different areas react over time.
-
-Several methods for finding effective connectivity already exist \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
-However, using more generally applicable causal discovery methods has not been elaborated on much.
-Also, none of these provide a measure of uncertainty.
-It is a fact that there are errors in any graph produced.
-This is inherent to the methods and the coarseness of the measurements.
-As there is no standard baseline to compare results with, a measure of uncertainty would be highly preferable.
-A framework for such a probabilistic approach is introduced in \cite{claassen2012}.
-
-We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000} can solve some of the difficulties effective connectivity suffers from.
-
-% Structure of the rest of proposal
-In the remainder of this proposal we briefly discuss some necessary background knowledge on connectivity and methods, we then discuss our strategy and finally we present a time-plan overview.
-
-\section{Theory}
-
-\subsection*{Brain connectivity}
-
-%Question: should this be included?
-\subsubsection*{Functional connectivity}
-Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
-This gives an insight in the organisation of the brain.
-Functional connectivity is strongly time-dependent, as activity changes rapidly providing and fMRI has a low temporal resolution.
-This means measured dependencies can be the result of statistical noise.
-Functional connectivity can be deduced by fMRI measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
-Resting state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity.
-
-\subsubsection*{Structural connectivity}
-Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
-This is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
-This intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
-Structural connectivity is less time dependent as it involves mappings of anatomical connections opposing temporal activity patterns.
-
-\subsubsection*{Effective connectivity}
-In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
-It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
-Effective connectivity indicates which brain regions stimulate other regions.
-Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
-It is possible to infer effective connectivity from structural and functional connectivity \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
-
-Research has been done on inferring effective connectivity directly with several methods.
-The most renowned methods are Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modeling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
-These are all methods dedicated to this problem.
-These methods also suffer from similar problems of which perhaps the most important is scaling.
-Most of these methods can only handle tens of nodes in reasonable time, which becomes problematic in realistic situations with up to tens of thousands of nodes, ideally.
-
-\subsection*{Causal discovery methods}
-The first causal method we will apply on brain structure data will be the PC-algorithm \cite{spirtes2000}.
-It is used to find structure, as well as causal relations between vertices in directed graphs. 
-If we already provide a graph skeleton, PC-algorithm can be used to determine directions.
-An adaptation of PC that is better adapted to finding latent variables is Fast Causal Inference \cite{spirtes2000}, though this method does not scale well to tens of variables.
-
-Another causal discovery method is the BCCD-algorithm \cite{claassen2012}, which gives similar results as PC, and is designed in an attempt to improve PC.
-This algorithm has some useful properties; it is a Bayesian approach and as such, it is more robust against noise.
-The authors have shown the method to perform as good as or better than other causal discovery algorithms such as PC on representative datasets.
-As is inherent to Bayesian methods, or approximations of Bayesian methods, the resulting network graph is not a  single structure but rather a distribution over possible structures.
-Concretely, this provides an uncertainty measure over edges in the network, giving additional information compared to PC.
-
-\begin{figure*}[bp]
-\begin{tabular}{c || c | c | c | c | c | c | c | c | c | c |}
-\setlength{\tabcolsep}{12pt}
-\renewcommand{\arraystretch}{1.3}
-Week          & 41 & 42 &  43 &  44 &  45 &  46 &  47 &  48 &  49 &  50 \tabularnewline \hline
-PC            & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Make implementation} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} Structural}  & \cellcolor[gray]{0.9} Functional & \multicolumn{4}{ |c| }{} & \tabularnewline \hline
-BCCD          & \multicolumn{4}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.85} Structural} & \cellcolor[gray]{0.85} Functional & \multicolumn{3}{ |c| }{} \tabularnewline \hline
-Analysis      & \multicolumn{3}{ |c| }{} & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} PC} & & \multicolumn{2}{ |c| }{\cellcolor[gray]{0.9} BCCD} & \multicolumn{2}{ |c| }{} \tabularnewline \hline
-Write article & \multicolumn{3}{ |c| }{} & \multicolumn{7}{ |c| }{\cellcolor[gray]{0.85}}  \tabularnewline \hline
-\end{tabular}
-\caption{Project time table.
-`Structural' implies applying the given method using structural data as a network skeleton.
-`Functional' implies inferring structure with the given method first through functional data.}
-\end{figure*}
-
-\subsection*{Data acquisition \& analysis}
-We will apply the mentioned methods on structural and functional data gathered from human subjects.
-We use the same functional and structural data as in Hinne et al. \cite{hinne2013}, of which six subjects were selected at random.
-To find effective connectivity, both data sources are needed.
-Structural connectivity of these subjects can be derived from DTI-measurements, using probabilistic tractography \cite{hinne2013}.
-Functional connectivity can be extracted from fMRI-measurement, possibly in combination with the structural 
-connectivity.
-Functional connectivity can be extracted from fMRI-measurement, possibly in combination with the structural connectivity.
-
-\section{Methods \& Strategy}
-%Short general introduction to both methods
-We use an existing implementation of BCCD to infer effective connectivity; PC will be implemented in this project.
-An implementation of BCCD suitable for this project is made available by the department of Intelligent Systems at the Radboud University Nijmegen.
-
-The first step will be to get the implementations for the algorithms working on the datasets with functional and structural data.
-
-Firstly we will use structural networks as a basis for the algorithms, and use functional information to infer causal relations.
-Later we will use functional information only, to infer both the network and the causal relations.
-We will compare this with the previous.
-As PC is a well-described algorithm \cite{spirtes2000} and should not be difficult to implement, so this will be done within this project.
-
-The second step is to apply the BCCD-algorithm, also to be used with and without provided network structure.
-Both methods provide a connectivity matrix representing effective connectivity.
-
-Results will then be evaluated and compared over all six subjects to see to which extent obtained networks are consistent across subjects.
-If so, we will analyse the properties and difficulties presented by Ramsey et al. \cite{ramsey2010}.
-We will conclude whether standard causal discovery methods provide a good direction of research for deriving and evaluating effective connectivity.
-
-% Licht hier toe met welke methoden we gaan vergelijken en wat we nog meer besproken hebben in de outline.
-% - Difficulties in causal discovery
-% - Comparison methods
-% - Comparison procedure
-% - Properties
-
-\section{Time schedule}
-This project is divided into four main parts: applying the PC-algorithm, applying the BCCD-algorithm, analysing the data (including cross-subject comparison) and writing the article.
-The schedule for applying the algorithm has been subdivided into applying it with functional connectivity using structural data as a basis, and applying it to only functional data.
-
-\bibliography{references}{}
-\addcontentsline{toc}{section}{References}
-\bibliographystyle{apalike}
+%\bibliography{references}{}
+%\addcontentsline{toc}{section}{References}
+%\bibliographystyle{apalike}
 

### 2013-11-26 docs/paper/paper.tex
Added title and bibtex back. Properly ignored LaTeX tmp files.
Added parts from proposal that seemed slightly relevant.

@@
   \textbf{T. de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}\\
-  \textbf{T. Claassen} - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
-  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
+%  \textbf{T. Claassen} - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
+%  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
 }
@@
 
-\section{Causal discovery}
+\maketitle
+
+\begin{abstract}
+% TODO: Add abstract. Look at slides Peter to see what is asked.
+\end{abstract}
+
+\section{Introduction}
+
+\paragraph{Motivation}
+The brain and particularly the human brain have been studied for hundreds of years.
+Today, the secrets of our brains still are one of the most sought-after.
+The techniques however have changed.
+We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
+Advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
+
+Of particular interest are causal relations between brain regions and cognitive and bodily functions.
+The question if brain region X is functionally or structurally connected to brain region Y and whether these are causally related is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
+Understanding brain structure implies understanding more about the brain as a whole.
+
+Not surprisingly, brain connectivity also finds applications in medicine.
+Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity.
+%TODO: Citation needed.
+Better methods for analysing connectivity could lead to more insight in these diseases.
+
+The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
+One type of connectivity analysis involves finding how brain region X causes what region Y does. This is called \emph{effective connectivity}.
+More concretely: how does the measured fMRI signal, representing an underlying neuronal process, determine what other neuronal processes are activated.
+Research regarding effective is relatively new and its application still faces many challenges that require further research \cite{ramsey2010}.
+The directed character in effective connectivity seems similar to traditional causal relations.
+In what follows, we suggest to use causal discovery methods for deriving effective connectivity.
+
+\paragraph{Problem statement}
+This project seeks to utilise the strengths of current state-of-the-art causal discovery methods by applying them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
+
+The work by \cite{ramsey2010} addresses several problems causal analysis in fMRI analysis suffers from.
+Methods in causal discovery have high computational costs, making it nearly impossible to calculate networks with more than hundreds of nodes.
+Another important fact is that fMRI Analysis indirectly measures brain activity.
+Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise, fail to capture underlying patterns or draw erroneous conclusions.
+Solutions to this particular problem have been proposed that do account for latent sources \cite{ramsey2010, waldorp2011}.
+Another problem is the strong diversity between subjects.
+Every brain is inherently different, making it non-trivial to combine subject data or even draw conclusions across subjects.
+Even within a single brain it changes how different areas react over time.
+
+Several methods for finding effective connectivity already exist \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
+However, using more generally applicable causal discovery methods has not been elaborated on much.
+Also, none of these provide a measure of uncertainty.
+It is a fact that there are errors in any graph produced.
+This is inherent to the methods and the coarseness of the measurements.
+As there is no standard baseline to compare results with, a measure of uncertainty would be highly preferable.
+A framework for such a probabilistic approach is introduced in \cite{claassen2012}.
+
+We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000} can solve some of the difficulties effective connectivity suffers from.
+
+% Structure of the rest of proposal
+In the remainder of this proposal we briefly discuss some necessary background knowledge on connectivity and methods, we then discuss our strategy and finally we present a time-plan overview.
+
+\section{Theory}
+
+\subsection{Causal discovery}
 An approach for finding effective connectivity may be found in the domain of causal discovery.
@@
 
-%\bibliography{references}{}
-%\addcontentsline{toc}{section}{References}
-%\bibliographystyle{apalike}
+\subsection{PC Algorithm}
+
+\subsection{Brain Connectivity}
+
+%Question: should this be included?
+\subsubsection*{Functional connectivity}
+Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
+This gives an insight in the organisation of the brain.
+Functional connectivity is strongly time-dependent, as activity changes rapidly providing and fMRI has a low temporal resolution.
+This means measured dependencies can be the result of statistical noise.
+Functional connectivity can be deduced by fMRI measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
+Resting state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity.
+
+\subsubsection*{Structural connectivity}
+Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
+This is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
+This intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
+Structural connectivity is less time dependent as it involves mappings of anatomical connections opposing temporal activity patterns.
+
+Hinne et al. \cite{hinne2013} proposed a Bayesian method for estimating structural networks based on Diffusion Weighted MRI (DWI) and probabilistic Tractography.
+The location of white-matter tracts does not become immediately clear from imagery alone.
+Other methods rely on thresholds to estimate location of said tracts.
+The proposed method results in a measure of uncertainty about where a hypothesised connection will terminate and so provides a clearly interpretable network structure as result.
+
+\subsubsection*{Effective connectivity}
+In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
+It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
+Effective connectivity indicates which brain regions stimulate other regions.
+Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
+It is possible to infer effective connectivity from structural and functional connectivity \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
+
+Research has been done on inferring effective connectivity directly with several methods.
+The most renowned methods are Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modeling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
+These are all methods dedicated to this problem.
+These methods also suffer from similar problems of which perhaps the most important is scaling.
+Most of these methods can only handle tens of nodes in reasonable time, which becomes problematic in realistic situations with up to tens of thousands of nodes, ideally.
+
+\section{Methods}
+
+\subsection*{Data acquisition \& analysis}
+We will apply the mentioned methods on structural and functional data gathered from human subjects.
+For this we are largely dependent on data collected by others.
+Hinne et al. has provided fMRI time-series data and structural network data as used and derived in \cite{hinne2013}, of which six subjects were selected at random.
+We can estimate functional connectivity from fMRI-measurement in combination with structural 
+networks according to a procedure currently submitted by Hinne et al. \cite{hinne2013submitted}.
+To find effective connectivity, both structural networks and functional data is needed.
+
+\section{Results}
+
+\section{Discussion}
+
+\section{Conclusion}
+
+\bibliography{references}{}
+\addcontentsline{toc}{section}{References}
+\bibliographystyle{plain}
 

### 2013-11-26 docs/paper/paper.tex
begin aan stukje pc-algoritme

--- 2013-11-26 docs/paper/paper.tex
@@
 
-Such a Bayesian network can be modelled as a graph with directed edges.
+Such a Bayesian network can be modelled as a graph with edges which may be directed or undirected.
 Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
-A causal relation between variables $X$ and $Y$ is denoted with a directed edge; $X \rightarrow Y$ implies that $X$ is a cause of $Y$.
+A causal relation between variables $X$ and $Y$ is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
 It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$. 
-A Bayesian network cannot contain cycles, as that would imply that a variable is it's own (indirect) parent and thus it's own cause.
+When there is a directed path $X \rightarrow Z_1 \rightarrow Z_2 \dots \rightarrow Y$, $X$ is said to be an ancestor of $Y$ and $Y$ is a descendant of $X$.
+An ancestor is an (possibly indirect) cause of its descendant.
+A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
 Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
@@
 
-A similar principle is conditional independence. If two variables are
+A similar principle is conditional independence.
+If two variables are generally dependent, they can still be independent given a third variable.
+An example of this is when a variable $X$ has two children $Y$ and $Z$.
+$Y$ and $Z$ are generally dependent, as they share a common cause $X$;
+When $X$ does not vary, $Y$ and $Z$ lose their common cause and they lose their dependency.
+As such, $Y$ and $Z$ are independent given \{$X$\}.
 
+We define a trail as a set of edges which would make up a path if those edges would be undirected.
+We can now define directional separation, or d-separation.
+Two nodes $X$ and $Y$ are d-separated given a set of nodes $Z$ (with $X, Y \notin Z$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
+\begin{itemize}
+\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $Z$.
+\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $Z$.
+\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $Z$.
+\end{itemize}
+If $X$ and $Y$ are d-separated given $Z$, $Z$ is called a separating set of $X$ and $Y$.
+
+\subsection{The PC-algorithm}
 Whether there is a dependency between two variables can be concluded by using a statistical test.
@@
 To derive causal relations, we can make use of conditional dependence.
-Three rules for finding causality are used in the PC-algorithm.
-One of those rules is based on conditional dependency
+A set of rules for finding a causal relation $X \rightarrow Y$ are used in the PC-algorithm.
+
+One of those rules is based on conditional dependency given a set variables other than $X$ and $Y$, and .
 

### 2013-11-26 docs/paper/paper.tex
Initial writings for methods.

@@
 \section{Methods}
+The original PC algorithm assumes causal sufficiency, excluding common confounders between nodes.
+An adaptation of this algorithm by Spirtes et al. yielded Modified PC, which returns a Markov equivalence class, rather than one specific instance of it.
+In literature it is this algorithm that is usually cited.
+
+As mentioned earlier, PC Algorithm has stability problems as the final solution depends on an initial ordering of vertices.
+This problem decreases when considering Modified PC; as latent common confounders are not excluded by design bi-directional arrows occur in subgraphs where original PC would attempt to direct a single arrow in either direction depending on node ordering.
+
+In a more recent publication by Ramsey et al. \cite{ramsey2012} a weaker but still sufficient assumption for graph faithfulness in PC algorithm is introduced.
+The resulting algorithmic changes provide better results and is able to mark unshielded triples as unfaithful.
+However, this method also require a significant amount of additional conditional independence tests.
+
+Based on similar ideas, we answer whether an unshielded triple is faithful based on already done conditional independence tests.
 

### 2013-11-26 docs/paper/paper.tex
twee scriptjes toegevoegd om gemiddelden van permutaties uit te rekenen en om implay met een colormap aan te roepen (automatisch een groter scherm is nog niet gelukt)

--- 2013-11-26 docs/paper/paper.tex
@@
 
+% ### dots at the end of arrows and lines ###
+
+\def \orightarrow {\circ\hspace{-0.42em}\rightarrow}
+\def \oleftarrow {\leftarrow\hspace{-0.42em}\circ}
+\def \orightline {\circ\hspace{-0.16em}-}
+\def \oleftline {-\hspace{-0.16em}\circ}
+\def \oline {\circ\hspace{-0.38em}-\hspace{-0.4em}\circ}
+\def \srightarrow {\text{\textasteriskcentered}\hspace{-0.62em}\rightarrow}
+\def \sleftarrow {\leftarrow\hspace{-0.62em}\text{\textasteriskcentered}}
+\def \srightline {\text{\textasteriskcentered}\hspace{-0.16em}-}
+\def \sleftline {-\hspace{-0.16em}\text{\textasteriskcentered}}
+\def \sline {\text{\textasteriskcentered}\hspace{-0.38em}-\hspace{-0.4em}\text{\textasteriskcentered}}
+
 % ############## End Macros ##############
@@
 Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
-A causal relation between variables $X$ and $Y$ is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
+If there is an edge between two variables $X$ and $Y$, this is denoted as $X - Y$.
+A causal relation is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
 It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$. 
@@
 We can now define directional separation, or d-separation.
-Two nodes $X$ and $Y$ are d-separated given a set of nodes $Z$ (with $X, Y \notin Z$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
+Two nodes $X$ and $Y$ are d-separated given a set of nodes $S$ (with $X, Y \notin S$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
 \begin{itemize}
-\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $Z$.
-\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $Z$.
-\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $Z$.
+\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $S$.
+\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $S$.
+\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $S$.
 \end{itemize}
-If $X$ and $Y$ are d-separated given $Z$, $Z$ is called a separating set of $X$ and $Y$.
+If $X$ and $Y$ are d-separated given $S$, $S$ is called a separating set of $X$ and $Y$.
+Every two independent variables have a separating set; if they are unconditionally independent, his is the empty set.
 
 \subsection{The PC-algorithm}
+$\srightarrow$
+$\sleftarrow$
+$\srightline$
+$\sleftline$
+$\sline$
+\cite{?}
 Whether there is a dependency between two variables can be concluded by using a statistical test.
@@
 However, this only results in an undirected graph, as it does not give any information about causality.
-To derive causal relations, we can make use of conditional dependence.
-A set of rules for finding a causal relation $X \rightarrow Y$ are used in the PC-algorithm.
+A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm.
 
-One of those rules is based on conditional dependency given a set variables other than $X$ and $Y$, and .
+One of these rules attempts to find so-called v-structures, which are of the form $X \rightarrow Z \leftarrow Y$.
+V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent, .
+As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
+If $Z$ is not in $S$, then the only possible causal relation can be that of a v-structure, so the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.
 
-\subsection{PC Algorithm}
+The other rules are based on arrows that have already been found.
+\begin{itemize}
+\item If there is an edge $X \sline Y$, and if there is a directed path from $X$ to $Y$, then $X \sline Y$ can be oriented as $X \srightarrow Y$.
+\item If there are edges $X \srightarrow Y$ and $Y \sleftline Z$, if the latter is not an edge $Y \sleftarrow Z$, and if $X$ and $Z$ are not connected, orient $Y \sline Z$ as $Y \rightarrow Z$. 
+\end{itemize}
 

### 2013-11-26 code/experiments/data_analyse.m

@@ -1,0 +1,7 @@
+GsPerm = permute(Gs, [1,4,2,3]);
+
+GsMean = zeros(77,116,116); 
+for subjNr = 1: 77
+	GsMean(subjNr,:,:) = squeeze(mean(squeeze(GsPerm(subjNr,:,:,:))));
+end
+GsMean = permute(GsMean, [2,3,1]);

### 2013-11-26 code/implayScript.m

@@ -1,0 +1,2 @@
+handle = implay(GsMean, 3);
+handle.Visual.ColorMap.MapExpression = cm; % set the colormap of implay

### 2013-11-29 code/experiments/data_analyse.m
data-analyse wat uitgebreid en een colormaps-bestandje toegevoegd (ruzie met colormaps, colormap('mijn_colormap') werkt niet

@@ -1,7 +1,27 @@
-GsPerm = permute(Gs, [1,4,2,3]);
+%load('../../data/multi-subject/mult_subjects_explicit');
 
-GsMean = zeros(77,116,116); 
-for subjNr = 1: 77
-	GsMean(subjNr,:,:) = squeeze(mean(squeeze(GsPerm(subjNr,:,:,:))));
-end
-GsMean = permute(GsMean, [2,3,1]);
+% per subject, only consider connections that appear in at least a fraction
+% of 'perm_treshold' of all permutations.
+perm_treshold = 0.90;
+% in total, only consider connections that appear in at least a fraction
+% of 'subj_treshold' of all subjects.
+subj_treshold = 0.7;
+
+Gs_avg = squeeze(mean(mean(Gs, 4), 1));
+
+% make a filter by applying these tresholds
+filter = squeeze(mean(mean(Gs, 4) > perm_treshold) > subj_treshold);
+filter_per_subject = mean(Gs,4) > perm_treshold;
+
+% just take the avarage of all permutations and subjects and filter this
+PDAG_avg_filtered = ((squeeze(mean(mean(PDAGs, 4),1)) - ones(116,116)) ./ Gs_avg) .* filter;
+PDAG_per_subject_filtered = (mean(PDAGs, 4) - ones(77,116,116)).* filter_per_subject;
+PDAG_assymetric_per_subject = PDAG_per_subject_filtered - permute(PDAG_per_subject_filtered, [1,3,2]);
+%handle = implay(permute(PDAG_assymetric_per_subject, [2,3,1]), 3); % transpose the last dimensions
+%handle.Visual.ColorMap.MapExpression = 'colormap_asymmetric';
+% draw things
+
+figure; imagesc(PDAG_avg_filtered);
+colormap ('colormap_filtered');
+axis square;
+%figure; imagesc((PDAG_avg_filtered - PDAG_avg_filtered')); colormap (colormap_asymmetric); axis square;

### 2013-11-29 code/experiments/data_analyse.m
Some more experimenting.

@@ -1,7 +1,10 @@
-GsPerm = permute(Gs, [1,4,2,3]);
+GsMean = mean(Gs,4);
+GsMean = permute(GsMean,[2,3,1]);
 
-GsMean = zeros(77,116,116); 
-for subjNr = 1: 77
-	GsMean(subjNr,:,:) = squeeze(mean(squeeze(GsPerm(subjNr,:,:,:))));
+PDAGsMean = mean(PDAGs,4);
+PDAGsMean = permute(PDAGsMean,[2,3,1]);
+
+PDAGsNorm = zeros(116,116,77);
+for n = 1:77
+    PDAGsNorm(:,:,n) = PDAGsMean(:,:,n) ./ GsMean(:,:,n);
 end
-GsMean = permute(GsMean, [2,3,1]);

### 2013-11-29 code/implayScript.m

@@ -1,2 +1,2 @@
-handle = implay(GsMean, 3);
-handle.Visual.ColorMap.MapExpression = cm; % set the colormap of implay
+handle = implay(PDAGsNorm, 3);
+handle.Visual.ColorMap.MapExpression = 'hot'; % set the colormap of implay

### 2013-12-02 code/experiments/multiple_subjects.old.m
Prepped multiple subjects script to run with original 6-subject data.

@@ -1,22 +1,22 @@
 function [PDAGs,Gs,Cs] = multiple_subjects(nr_perms, nr_subjects, T, explicit, mult_sepsets, alpha, filename)
 % for nr_subjects and T: give a negative number to take all subjects/data
 
-load('../../data/data.mat');
+load('../../data/AAL_FuncData.mat');
 addpath('../pc');
 addpath('../independence');
 
-if (T <= 0 || T > size(Xs,2))
-	T = size(Xs,2); % Number of data points
+if (T <= 0 || T > size(FuncData{1},2))
+	T = size(FuncData{1},2); % Number of data points
 end
 
 % Dataset reduction
-Xs = Xs(:,1:T,:);
+FuncData = cellfun(@(x)x(:,1:T), FuncData, 'UniformOutput', false);
 
-if (nr_subjects <= 0 || nr_subjects > size(Xs,1))
-	nr_subjects = size(Xs,1);
+if (nr_subjects <= 0 || nr_subjects > length(FuncData))
+	nr_subjects = length(FuncData);
 end
 
-N = size(Xs,3); % Number of variables
+N = size(FuncData{1},1); % Number of variables
 
 Cs = zeros(nr_subjects, N, N);
 Gs = zeros(nr_subjects,N,N,nr_perms);
@@ -24,8 +24,8 @@ PDAGs = zeros(nr_subjects, N,N,nr_perms);
 
 fprintf('applying zscore and calculating covariance...\n');
 for subject_nr = 1:nr_subjects
-% 	Xs(subject_nr,:,:) = zscore(squeeze(Xs(subject_nr,:,:)));
-	C = cov(squeeze(Xs(subject_nr,:,:)));
+	FuncData{subject_nr} = zscore(transpose(FuncData{subject_nr}));
+	C = cov(FuncData{subject_nr});
 	Cs(subject_nr,:,:) = C;
 end
 
@@ -59,7 +59,6 @@ for it = 1:nr_perms
 		tmp=cputime;
 		fprintf('Finding causal directions...\n');
 		modified = 1;
-
 		PDAG_perm = step2_pc(G_perm, sepset_perm, modified, explicit, 'cond_indep_fisher_z',C,T,alpha);	
 		fprintf('Done directing arrows.\n');
 		tmp=cputime-tmp;

### 2013-12-03 code/experiments/concat_data.m
Added results on six subject experiments.

@@ -1,5 +1,5 @@
-names = 10:10:50;
-filename = '../../data/multi-subject/mult_subjects_explicit_p';
+names = 10:10:60;
+filename = '../../data/multi-subject/mult_subjects_expl_p';
 
 new_PDAGs = [];
 new_Gs = [];
@@ -15,5 +15,5 @@ end
 
 Gs = new_Gs;
 PDAGs = new_PDAGs;
-save('../../data/multi-subject/mult_subjects_explicit.mat', 'Cs', 'Gs', 'PDAGs');
+save('../../data/multi-subject/mult_subjects_expl.mat', 'Cs', 'Gs', 'PDAGs');
 clear;

### 2013-12-03 code/experiments/doodle.m
Added plot file for doodling with results.

@@ -1,0 +1,73 @@
+% perm = [1:2:90 2:2:90 91:116];
+perm = 1:116;
+
+G_avg = mean(Gs,4);
+G_avg = G_avg(:,perm,perm);
+Struct_avg = squeeze(mean(G_avg,1));
+StdDev_avg = squeeze(std(G_avg,1));
+
+PDAG_avg = mean(PDAGs,4);
+PDAG_std = std(PDAGs,0,4);
+PDAG_avg = PDAG_avg(:,perm,perm);
+Subj_avg = squeeze(mean(PDAG_avg,1));
+Subj_stddev = squeeze(std(PDAG_avg,1));
+antisym_subj_avg = (Subj_avg - Subj_avg');
+
+% STRUCTURE
+% Plot full subject mean
+figure;
+colormap 'gray';
+subplot(1,2,1);
+imagesc(Struct_avg); title('Complete mean structure'); colorbar; axis square;
+subplot(1,2,2);
+imagesc(Struct_avg - StdDev_avg); title('Structure-stdev'); colorbar; axis square;
+
+% Plot per subject mean
+h = figure; colormap 'gray'; %colormap(cmap);
+for i=1:6
+    imagesc(squeeze(G_avg(i,:,:)));
+    title(sprintf('Structure - Subj %d',i));
+    colorbar;
+    axis square;
+    pause;
+end
+close(h);
+
+% PDAG
+figure;
+colormap(cmap);
+subplot(2,2,1);
+imagesc(Subj_avg); colorbar; axis square; title('Complete mean PDAG');
+subplot(2,2,2);
+imagesc(antisym_subj_avg); colorbar; axis square; title('Antisymmetric mean PDAG');
+subplot(2,2,3);
+imagesc(Subj_stddev); colorbar; axis square; title('STD PDAG');
+subplot(2,2,4);
+imagesc(Subj_avg - Subj_stddev); colorbar; axis square; title('mean-std PDAG');
+
+% Plot per subject mean
+h = figure; colormap(cmap);
+for i=1:6
+    subplot(2,2,1);
+    imagesc(squeeze(PDAG_avg(i,:,:)));
+    title(sprintf('PDAG - Subj %d',i));
+    colorbar;
+    axis square;
+    subplot(2,2,2);
+    imagesc(squeeze(PDAG_avg(i,:,:)) - squeeze(PDAG_avg(i,:,:))');
+    title(sprintf('ASSYM PDAG - Subj %d',i));
[…]

### 2013-12-03 docs/paper/paper.tex
bezig met het stukje over causal faithfulness i.c.m. PC

@@
 If $X$ and $Y$ are d-separated given $S$, $S$ is called a separating set of $X$ and $Y$.
-Every two independent variables have a separating set; if they are unconditionally independent, his is the empty set.
+Every two independent variables have a separating set; if they are unconditionally independent, this is the empty  set.
+
+\subsection{Assumptions about the structure}
+In theory, every deterministic real-world situation must be described by some Bayesian model, as all situations conform to causality.
+But the underlying real-world data such as the measured brain activity used in this analysis usually does not conform to a simple model.
+Assumption can be made about the model to make it easier to infer it, but this increases the risk of drawing false conclusions.
+
+One assumption that can be made about the model is that of causal sufficiency. 
+When it is assumed that the model fully describes the real network, so-called V-structures can be concluded. 
+When there is a structure $X - Y - Z$, in which $X$ and $Z$ are not connected, the only possible orientation of the connections is $X \rightarrow Y \leftarrow Z$. % TODO: extra uitleg?
+An approach to finding a model is therefore to find all V-structures and orient them in this way.
+
+However, it is very unlikely that all events which are relevant for the causal graph are measured; 
+even with high temporal and spatial resolutions, the entire fysiology of the brain is too complex to capture.
+Latent variables are nodes which are present in the underlying Bayesian graph, but which are not measured.
+When there is a structure $X \leftarrow L \rightarrow Y$ in which $L$ is a latent variable, an edge between $X$ and $Y$ might be found while the these two nodes have no direct ancestor/descendant-relation between them.
+A directed edge $X \rightarrow Y$ or $X \leftarrow Y$ would be erroneous, so an arrow 
+The complex reality might be very similar to a simple model based on the measured data, but it might also be very different.
+This correctness of the data is called causal faithfulness.
+As such, the way of describing V-structures might not draw false conclusions, depending on the causal faithfulness.
+When causal sufficiency is not assumed,
 
 \subsection{The PC-algorithm}
-$\srightarrow$
-$\sleftarrow$
-$\srightline$
-$\sleftline$
-$\sline$
 \cite{?}
@@
 A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm.
+Multiple variations of this algorithm are described by Spirtes, Glymour and Scheines \ref{?}.
+For this analysis, we have used the "modified pc algorithm" which does not assume causal sufficiency, and thus is able to handle latent variables.
 
-One of these rules attempts to find so-called v-structures, which are of the form $X \rightarrow Z \leftarrow Y$.
-V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent, .
+The first of these rules attempts to find so-called v-structures, which are of the form $X \rightarrow Z \leftarrow Y$.
+V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent.
 As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
@@
 
-The other rules are based on arrows that have already been found.
+After this rules which finds V-structures, the other rules are applied in no particular order. These rules are based on arrows that have already been found.
 \begin{itemize}

### 2013-12-03 docs/paper/paper.tex
Wrote methods more neatly.

@@
 \section{Methods}
-The original PC algorithm assumes causal sufficiency, excluding common confounders between nodes.
-An adaptation of this algorithm by Spirtes et al. yielded Modified PC, which returns a Markov equivalence class, rather than one specific instance of it.
-In literature it is this algorithm that is usually cited.
+% Different adaptations of PC
+The original PC algorithm assumes causal sufficiency, excluding the existence of common confounders between nodes.
+This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
+Also measuring methods are not yet advanced enough to measure the full detail of the brain, thus skipping over details.
+In the perspective of causal inference, these missing areas can also appear as confounders.
 
-As mentioned earlier, PC Algorithm has stability problems as the final solution depends on an initial ordering of vertices.
-This problem decreases when considering Modified PC; as latent common confounders are not excluded by design bi-directional arrows occur in subgraphs where original PC would attempt to direct a single arrow in either direction depending on node ordering.
+% Modified PC
+\paragraph{Modified PC}
+To compensate for this, an adaptation of PC algorithm by Spirtes et al. returns a Markov equivalence class, rather than one specific instance of it.
+In literature it is this algorithm that is usually cited under the name PC Algorithm.
+This also makes PC algorithm more stable as the number of Markov equivalence classes is less than the number of times data can be permuted.
+However the order of vertex removal still plays a key role in the end result.
+This causal-insufficient version is named Modified PC.
 
+\paragraph{Conservative PC}
 In a more recent publication by Ramsey et al. \cite{ramsey2012} a weaker but still sufficient assumption for graph faithfulness in PC algorithm is introduced.
+Through this weaker but still adjacency-faithful assumption, the method is more cautious than PC algorithm in drawing unambiguous conclusions on causal orientations.
+That is why this method is named Conservative PC (CPC).
 The resulting algorithmic changes provide better results and is able to mark unshielded triples as unfaithful.
-However, this method also require a significant amount of additional conditional independence tests.
+However, CPC requires a significant amount of additional conditional independence tests.
 
-Based on similar ideas, we answer whether an unshielded triple is faithful based on already done conditional independence tests.
+\paragraph{...}
 
 \subsection*{Data acquisition \& analysis}
+% TODO: vraag Max naar de details van de meetmethode en subject acquisition
 We will apply the mentioned methods on structural and functional data gathered from human subjects.

### 2013-12-03 docs/paper/paper.tex
Added section on extensions of PC.

@@
 
+% Conservative PC
 \paragraph{Conservative PC}
-In a more recent publication by Ramsey et al. \cite{ramsey2012} a weaker but still sufficient assumption for graph faithfulness in PC algorithm is introduced.
+Step x in PC algorithm assigns shielded triples based on a single separating set. % TODO: Add step
+Simply put, it is assumed the single found separating set is enough to show conditional independence. 
+Ramsey et al. \cite{ramsey2012} introduce a weaker but still sufficient assumption for directing unshielded triples.
 Through this weaker but still adjacency-faithful assumption, the method is more cautious than PC algorithm in drawing unambiguous conclusions on causal orientations.
 That is why this method is named Conservative PC (CPC).
-The resulting algorithmic changes provide better results and is able to mark unshielded triples as unfaithful.
-However, CPC requires a significant amount of additional conditional independence tests.
+The resulting algorithmic changes provide better results and is able to mark incorrectly unshielded triples as unfaithful.
+Sadly, CPC requires a significant amount of additional conditional independence tests, exponential in the largest branching degree of the tested structure.
 
-\paragraph{...}
+% Multiple sepsets
+\paragraph{Multiple Separating Sets}
+The idea of CPC, to make a weaker assumption on graph faithfulness, is very potential.
+Instead of doing additional tests, one could attempt to determine causal unfaithfulness with less dependency tests.
+An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found, to test all other vertex subsets of the same order as well.
+This finds all separating sets of the lowest order without increasing complexity.
+Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
+By having more than one separating set, this constraint is strengthened.
+
+% Unfaithfulness test
+One could use these additional separating sets to test oriented unshielded triples for unfaithfulness, similar to CPC.
+Apart from serving as a sanity check, this could serve as the basis of a repair algorithm for errors made by PC.
+
+% Explicit test
+A more direct way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
+If so, the graph is likely to be more faithful to its modeled distribution.
+
+\paragraph{EMS PC}
+...
 

### 2013-12-04 docs/paper/paper.tex
Changed paragraph spacing and document type.

@@
 % Needed packages
-\documentclass[a4paper, 10pt, english, twocolumn]{article}
+\documentclass[a4paper, 10pt, english, onecolumn]{article}
 \usepackage[english]{babel}
@@
 \maketitle
+
+\setlength{\parindent}{0.0cm}
+\setlength{\parskip}{0.25cm}
 
@@
 % Explicit test
-A more direct way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
+A more explicit way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
 If so, the graph is likely to be more faithful to its modeled distribution.
 
-\paragraph{EMS PC}
-...
+\paragraph{EMS-PC}
+We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
+% TODO: Algorithmic part here.
 

### 2013-12-04 docs/paper/paper.tex
schrijven schrijven schrijven

--- 2013-12-03 docs/paper/paper.tex
@@
 An example of this is when a variable $X$ has two children $Y$ and $Z$.
-$Y$ and $Z$ are generally dependent, as they share a common cause $X$;
-When $X$ does not vary, $Y$ and $Z$ lose their common cause and they lose their dependency.
+$Y$ and $Z$ are generally dependent, as they share a common cause or \emph{confounder} or \emph{confounder} $X$;
+When $X$ does not vary, $Y$ and $Z$ lose their confoundernd they lose their dependency.
 As such, $Y$ and $Z$ are independent given \{$X$\}.
@@
 When it is assumed that the model fully describes the real network, so-called V-structures can be concluded. 
-When there is a structure $X - Y - Z$, in which $X$ and $Z$ are not connected, the only possible orientation of the connections is $X \rightarrow Y \leftarrow Z$. % TODO: extra uitleg?
+When there is a structure $X - Y - Z$, in which $X$ and $Z$ are not connected, the only possible orientation of the connections is $X \rightarrow Y \leftarrow Z$. % TODO: extra uitleg, d-separation?
 An approach to finding a model is therefore to find all V-structures and orient them in this way.
@@
 Latent variables are nodes which are present in the underlying Bayesian graph, but which are not measured.
-When there is a structure $X \leftarrow L \rightarrow Y$ in which $L$ is a latent variable, an edge between $X$ and $Y$ might be found while the these two nodes have no direct ancestor/descendant-relation between them.
-A directed edge $X \rightarrow Y$ or $X \leftarrow Y$ would be erroneous, so an arrow 
 The complex reality might be very similar to a simple model based on the measured data, but it might also be very different.
 This correctness of the data is called causal faithfulness.
+When there is a structure $X \leftarrow L \rightarrow Y$ in which $L$ is a latent variable, an edge between $X$ and $Y$ might be found while the these two nodes have no direct ancestor/descendant-relation between them;
+They might be related through a confounder, which is indicated as $X \leftrightarrow Y$.
+As such, for a V-structure for which the underlying Bayesian network has latent variables, a directed edge $X \rightarrow Y$ or $X \leftarrow Y$ would be erroneous. 
 As such, the way of describing V-structures might not draw false conclusions, depending on the causal faithfulness. 
-When causal sufficiency is not assumed,
+A weaker conclusion is that a V-structure should be oriented as $X \orightarrow Y \oleftarrow Z$, in which the circles at the end of an edge mean that the form of this end is not yet know; an arrow $\orightarrow$ might be a simple arrow $\rightarrow$ or a double arrow $\leftrightarrow$.
 
@@
 \cite{?}
-Whether there is a dependency between two variables can be concluded by using a statistical test.
-In this way, a graph can be inferred representing all dependencies between nodes.
-However, this only results in an undirected graph, as it does not give any information about causality.
+For finding directed edges, the PC-algorithm first finds the undirected structure.
+For each two nodes $X$ and $Y$ which are not neighbours in this structure, it also finds a minimal separating set, which is the smallest set acting as a separated set.
+This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.
+
+The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ (the number of neighbours decreases while edges are removed), using the fisher-test.
+If such an independence is found, this subset is a separating set of $X$ and $Y$ and they are not connected.
+If no such an independence is found, they are connected.
+To find the minimal separating set, the subsets of neighbours are traversed in order of size; the first separating set which is found, is the minimal one.
+
+After the structure and separating sets have been found, the PC-algorithm attempts to orient the edges.
 A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm.
@@
 
-The first of these rules attempts to find so-called v-structures, which are of the form $X \rightarrow Z \leftarrow Y$.
+The first of these rules attempts to find and orient V-structures.
 V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent.
@@
 
-After this rules which finds V-structures, the other rules are applied in no particular order. These rules are based on arrows that have already been found.
+After this rule which finds V-structures, the other rules are applied in no particular order. These rules are based on arrows that have already been found.
 \begin{itemize}

### 2013-12-06 docs/paper/paper.tex
stukje aan PC toegevoegd
meer pseudocode (modified het nuttigst?)

@@
 \usepackage[english]{babel}
-\usepackage[cm]{fullpage}
+%\usepackage[cm]{fullpage}
 \usepackage{cite}
 \usepackage{anysize}
-\usepackage[compact]{titlesec}
+%\usepackage[compact]{titlesec}
 \usepackage{graphicx}
@@
 \usepackage{algorithmicx}
+%\usepackage{algorithmic}
 \usepackage{algorithm}
@@
 
-\newcommand{\keywords}[1]{\par\noindent 
-{\bf Keywords\/}. #1}
+%\newcommand{\keywords}[1]{\par\noindent 
+%{\bf Keywords\/}. #1}
 
 % Margins & Headers
-\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
-\columnsep 0.4in
-\footskip 0.4in 
-\usepackage{changepage}
+%\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
+%\columnsep 0.4in
+%\footskip 0.4in 
+%\usepackage{changepage}
 
@@
 \subsection{The PC-algorithm}
+
+\subsection{Informal description}
 \cite{?}
@@
 \end{itemize}
+
+\subsection{Pseudocode}
+
+The first part of the algorithm finds an undirected graph $G$. For each pair of points which are not connected in this graph, it also finds a separating set:
+
+\begin{algorithmic}
+\Function{Structure\_PC}{$data$ used to test independence}
+\State connect all pairs of point in undirected graph $G$.
+\State $n \gets 0$
+\While{there are connected pairs $X$, $Y$ such that $n < |\operatorname{Adjecencies}(X)\setminus\{Y\}| $}
+  \ForAll{such connected pairs $X$, $Y$}
+    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
+      \State test whether $X$ and $Y$ are d-separated given $S$ (Fisher-Z-test on $data$)
+      \If{$X$ and $Y$ are d-separated given $S$}
+        \State remove the edge $X - Y$ from G
+        \State record $S$ as the separating set of $X$ and $Y$
+        \State continue with the next pair $X$, $Y$
+      \EndIf
+    \EndFor
+  \EndFor
+  n \gets n+1
+\EndWhile
+\Return G and the separating sets
+\EndFunction
+\end{algorithmic}
+
+The second part orients edges in this graph $G$ based on the separating sets:
+
+\begin{algorithmic}
+\Function{Directional\_PC}{$G$, $sepsets$}
+ \State $\text{PDAG} \gets G$
+  \ForAll{triples $X$, $Y$, $Z$ with the structure $X - Y - Z$ in $G$ (where $X$ and $Z$ are not connected)}
+    \If{$Y$ is not in the separating set of $X$ and $Z$}
+      \State Orient $X - Y - Z$ as $X \rightarrow Y \leftarrow Z$ in $\text{PDAG}$
+    \EndIf
+  \EndFor
+  \Repeat
+    \ForAll{ordered, connected pairs $X$, $Y$}
+      %\If{there is a node $Z$ such that }
+
+      %\EndIf
+    \EndFor
+  \Until{no more edges are oriented}
+\EndFunction
+\end{algorithmic}
 

### 2013-12-08 docs/paper/paper.tex
bla

--- 2013-12-04 docs/paper/paper.tex
@@
 As such, $Y$ and $Z$ are independent given \{$X$\}.
+In real-world data, independence can be found with methods such as the Fisher-Z-test.
 
@@
 
-The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ (the number of neighbours decreases while edges are removed), using the fisher-test.
+The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ (the number of neighbours decreases while edges are removed), using the fisher-Z-test.
 If such an independence is found, this subset is a separating set of $X$ and $Y$ and they are not connected.
 If no such an independence is found, they are connected.
-To find the minimal separating set, the subsets of neighbours are traversed in order of size; the first separating set which is found, is the minimal one.
+To find the minimal separating set, all subsets of neighbours are traversed in order of size; the first separating set which is found, is the minimal one.
 
@@
 As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
-If $Z$ is not in $S$, then the only possible causal relation can be that of a v-structure, so the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.
+If $Z$ is not in $S$, then the only possible causal relation can be that of a V-structure, so the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.
 
@@
 \end{itemize}
+
+%TODO: pseudocode nodig? of een verwijzing naar het boek (die pseudocode is wat onduidelijk)?
+
+\usepackage{algorithmic}
+
+%\begin{algorithm}[PC-algorithm]
+\STATE Starting state of undirected graph G: connect all pairs of points. Set n = 0.
+\WHILE{there are connected pairs $X$, $Y$ such that |Adjecencies(G,X)\{Y}| < }
+
+\ENDWHILE
+%\end{algorithm}
 

### 2013-12-08 docs/paper/paper.tex

--- 2013-12-06 docs/paper/paper.tex
@@
 \end{algorithmic}
+
+\ENDWHILE
+%\end{algorithm}
 

### 2013-12-08 docs/paper/paper.tex
stukje over EMS aangevuld, pseudocode af

@@
 \def \srightline {\text{\textasteriskcentered}\hspace{-0.16em}-}
-\def \sleftline {-\hspace{-0.16em}\text{\textasteriskcentered}}
+\def \sleftline {-\hspace{-0.40em}\text{\textasteriskcentered}}
 \def \sline {\text{\textasteriskcentered}\hspace{-0.38em}-\hspace{-0.4em}\text{\textasteriskcentered}}
+\def \soline {\text{\textasteriskcentered}\hspace{-0.16em}-\hspace{-0.4em}\circ}
+\def \osline {\circ\hspace{-0.38em}-\hspace{-0.16em}\text{\textasteriskcentered}}
 
@@
 \subsection{Informal description}
-\cite{?}
+The PC algorithm as used for these experiments is the version which is modified to handle latent variables, as also described by Spirtes, Glymour and Scheines \cite{spirtes2000}, pages 165-167.
 For finding directed edges, the PC-algorithm first finds the undirected structure.
@@
 
-The first part of the algorithm finds an undirected graph $G$. For each pair of points which are not connected in this graph, it also finds a separating set:
+The first part of the algorithm finds an undirected graph $G$.
+For each pair of points which are not connected in this graph, it also finds a separating set.
+$\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph G.
+This graph changes during the algorithm and so does $\operatorname{Adjacencies} (X)$.
 
 \begin{algorithmic}
-\Function{Structure\_PC}{$data$ used to test independence}
+\Function{Structure\_PC}{$data$}
 \State connect all pairs of point in undirected graph $G$.
@@
   \EndFor
-  n \gets n+1
+  \State $n \gets n+1$
 \EndWhile
-\Return G and the separating sets
+\Return $G$ and the separating sets
 \EndFunction
@@
 
-The second part orients edges in this graph $G$ based on the separating sets:
+The second part orients edges in this graph $G$ based on the separating sets.
+In this pseudocode, an edge with a star at the end means that that end may be either an arrowhead, an "o" or an empty mark.
+When an edge is oriented as an edge with a star at the end, this denotes that this end is not changed.
 
@@
 \Function{Directional\_PC}{$G$, $sepsets$}
- \State $\text{PDAG} \gets G$
-  \ForAll{triples $X$, $Y$, $Z$ with the structure $X - Y - Z$ in $G$ (where $X$ and $Z$ are not connected)}
+  \State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
+  \Comment{Find and orient V-structures}
+  \ForAll{triples $X$, $Y$, $Z$ with the structure $X\sline Y \sline Z$ in $G$ (where $X$ and $Z$ are not connected)}
     \If{$Y$ is not in the separating set of $X$ and $Z$}
-      \State Orient $X - Y - Z$ as $X \rightarrow Y \leftarrow Z$ in $\text{PDAG}$
+      \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
     \EndIf
@@
     \ForAll{ordered, connected pairs $X$, $Y$}
-      %\If{there is a node $Z$ such that }
-
-      %\EndIf
+      \Comment{Rule 1}
+      \If{$X \osline Y$ or $X \sleftline Y$, and there is a node $Z$ such that $Z \srightarrow X$}
+        \State Orient $X \sline Y$ as $X \rightarrow Y$
+      \EndIf
+      \Comment{Rule 2}
+      \If{there is a directed path from $X$ to $Y$}
+        \State Orient $X \sline Y$ as $X \srightarrow Y$
+      \EndIf
     \EndFor
   \Until{no more edges are oriented}
+  \State \Return PDAG
 \EndFunction
 \end{algorithmic}
-
-\ENDWHILE
-%\end{algorithm}
 

### 2013-12-09 docs/paper/paper.tex
Merge.

@@
 \subsection{Pseudocode}
+\label{sec:pseudocode}
 
@@
 
+\begin{algorithm}
+\caption{Structure PC}
 \begin{algorithmic}
@@
     \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
-      \State test whether $X$ and $Y$ are d-separated given $S$ (Fisher-Z-test on $data$)
+      \Comment{(Fisher-Z-test on $data$)}
+      \State test whether $X$ and $Y$ are d-separated given $S$
       \If{$X$ and $Y$ are d-separated given $S$}
@@
 \end{algorithmic}
+\end{algorithm}
 
@@
 When an edge is oriented as an edge with a star at the end, this denotes that this end is not changed.
-
+\begin{algorithm}
+\caption{Directional PC}
 \begin{algorithmic}
@@
   \State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
-  \Comment{Find and orient V-structures}
+  \Comment{V-structure-rule}
   \ForAll{triples $X$, $Y$, $Z$ with the structure $X\sline Y \sline Z$ in $G$ (where $X$ and $Z$ are not connected)}
@@
 We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
-% TODO: Algorithmic part here.
+Two changes have been made in this adaption with respect to modified PC as described by Spirtes et al \cite{Spirtes2000}.
+In the algorithm as described in section ~\ref{sec:pseudocode}, the search for a separating sets for $X$ and $Y$ stops when such a set is found; instead, EMS continues the search for more separating sets of the same order.
+
+Also, the condition to orient $X - Y - Z$ according to the V-structure-rule has been made more strict.
+As multiple separating sets have been found instead of only one, it needs to be checked whether $Y$ is in not in \emph{any} of those sets before the V-structure can be oriented.
+Also, in addition to the original condition, $X$ and $Z$ are tested for d-separation given $\{Y\}$.
+If they are found to be conditionally dependent, the triple is oriented as as $X \srightarrow Y \sleftarrow Z$.
 
@@
 % TODO: vraag Max naar de details van de meetmethode en subject acquisition
+In order 
 We will apply the mentioned methods on structural and functional data gathered from human subjects.

### 2013-12-09 docs/paper/paper.tex
kleine wijzigingetjes aan pseudocode
beginnetje results

@@
 After the structure and separating sets have been found, the PC-algorithm attempts to orient the edges.
+After Spirtes et al, we use a star-symbol at the end of an edge to denote that any mark may be present: an arrowhead, an "o" or an empty mark.
+When an edge is oriented as an edge with a star, this denotes that this end is not changed.
 A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm.
@@
 
-The first of these rules attempts to find and orient V-structures.
-V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent.
+The first of these rules attempts to find and orient unshielded triples.
+A triple $<X, Y, Z>$ is considered unshielded if $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent.
 As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
-If $Z$ is not in $S$, then the only possible causal relation can be that of a V-structure, so the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.
+If $Z$ is not in $S$, then the only possible causal relation within this structure can be that of a so-called V-structur; the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.
 
@@
 \end{itemize}
-
-\subsection{Pseudocode}
-\label{sec:pseudocode}
-
-The first part of the algorithm finds an undirected graph $G$.
-For each pair of points which are not connected in this graph, it also finds a separating set.
-$\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph G.
-This graph changes during the algorithm and so does $\operatorname{Adjacencies} (X)$.
-
-\begin{algorithm}
-\caption{Structure PC}
-\begin{algorithmic}
-\Function{Structure\_PC}{$data$}
-\State connect all pairs of point in undirected graph $G$.
-\State $n \gets 0$
-\While{there are connected pairs $X$, $Y$ such that $n < |\operatorname{Adjecencies}(X)\setminus\{Y\}| $}
-  \ForAll{such connected pairs $X$, $Y$}
-    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
-      \Comment{(Fisher-Z-test on $data$)}
-      \State test whether $X$ and $Y$ are d-separated given $S$
-      \If{$X$ and $Y$ are d-separated given $S$}
-        \State remove the edge $X - Y$ from G
-        \State record $S$ as the separating set of $X$ and $Y$
-        \State continue with the next pair $X$, $Y$
-      \EndIf
-    \EndFor
-  \EndFor
-  \State $n \gets n+1$
-\EndWhile
-\Return $G$ and the separating sets
-\EndFunction
-\end{algorithmic}
-\end{algorithm}
-
-The second part orients edges in this graph $G$ based on the separating sets.
-In this pseudocode, an edge with a star at the end means that that end may be either an arrowhead, an "o" or an empty mark.
-When an edge is oriented as an edge with a star at the end, this denotes that this end is not changed.
-\begin{algorithm}
-\caption{Directional PC}
-\begin{algorithmic}
-\Function{Directional\_PC}{$G$, $sepsets$}
-  \State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
-  \Comment{V-structure-rule}
-  \ForAll{triples $X$, $Y$, $Z$ with the structure $X\sline Y \sline Z$ in $G$ (where $X$ and $Z$ are not connected)}
-    \If{$Y$ is not in the separating set of $X$ and $Z$}
-      \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
-    \EndIf
-  \EndFor
-  \Repeat
-    \ForAll{ordered, connected pairs $X$, $Y$}
-      \Comment{Rule 1}
-      \If{$X \osline Y$ or $X \sleftline Y$, and there is a node $Z$ such that $Z \srightarrow X$}
-        \State Orient $X \sline Y$ as $X \rightarrow Y$
-      \EndIf
-      \Comment{Rule 2}
-      \If{there is a directed path from $X$ to $Y$}
-        \State Orient $X \sline Y$ as $X \srightarrow Y$
-      \EndIf
-    \EndFor
-  \Until{no more edges are oriented}
-  \State \Return PDAG
-\EndFunction
-\end{algorithmic}
-\end{algorithm}
 
@@
 
+\subsection{Pseudocode}
+\label{sec:pseudocode}
+The first part of the algorithm finds an undirected graph $G$.
+For each pair of points which are not connected in this graph, it also finds a separating set.
+$\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph $G$.
+This graph changes during the algorithm and so does $\operatorname{Adjacencies} (X)$.
+
+\begin{algorithm}
+\caption{Structure PC}
+\begin{algorithmic}
+\Function{Structure\_PC}{$data$}
+\State connect all pairs of point in undirected graph $G$.
+\State $n \gets 0$
+\While{there are connected pairs $X$, $Y$ such that $n < |\operatorname{Adjecencies}(X)\setminus\{Y\}| $}
+  \ForAll{such connected pairs $X$, $Y$}
+    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
+      \Comment{(Fisher-Z-test on $data$)}
+      \State test whether $X$ and $Y$ are d-separated given $S$
+      \If{$X$ and $Y$ are d-separated given $S$}
+        \State remove the edge $X - Y$ from G
+        \State record $S$ as a separating set of $X$ and $Y$
+      \EndIf
+    \EndFor
+  \EndFor
+  \State $n \gets n+1$
+\EndWhile
+\Return $G$ and the separating sets
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+The second part orients edges in this graph $G$ based on the separating sets.
+\begin{algorithm}
+\caption{Directional PC}
+\begin{algorithmic}
+\Function{Directional\_PC}{$G$, $sepsets$}
+  \State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
+  \Comment{V-structure-rule}
+  \ForAll{unshielded triples $<X, Y, Z>$}
+    \If{$Y$ is not in any separating set of $X$ and $Z \And X$ and $Z$ are dependent given $Y$}
+      \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
+    \EndIf
+  \EndFor
+  \Repeat
+    \ForAll{ordered, connected pairs $X$, $Y$}
+      \Comment{Rule 1}
+      \If{$X \osline Y$ or $X \sleftline Y$, and there is a node $Z$ such that $Z \srightarrow X$}
+        \State Orient $X \sline Y$ as $X \rightarrow Y$
+      \EndIf
+      \Comment{Rule 2}
+      \If{there is a directed path from $X$ to $Y$}
+        \State Orient $X \sline Y$ as $X \srightarrow Y$
+      \EndIf
+    \EndFor
+  \Until{no more edges are oriented}
+  \State \Return PDAG
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
 \subsection*{Data acquisition \& analysis}
 % TODO: vraag Max naar de details van de meetmethode en subject acquisition
-In order 
 We will apply the mentioned methods on structural and functional data gathered from human subjects.
@@
 \section{Results}
+At first, we have applied modified PC on the strucural connectivity data of six subjects.
+The results are presented in figure ~\ref{img:??}.
+In this matrix, indices 1-45 represent the brain regions of the left hemisphere, regions 45-90 represent the right hemisphere, and regions 91-116 represent regions in the cerebellum.
+%TODO left or right hemisphere?
 

### 2013-12-09 docs/paper/paper.tex
Moar methods!

--- 2013-12-09 docs/paper/paper.tex
@@
 %Question: should this be included?
-\subsubsection*{Functional connectivity}
+\subsubsection{Functional connectivity}
 Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
@@
 
-\subsubsection*{Structural connectivity}
+\subsubsection{Structural connectivity}
 Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
@@
 
-\subsubsection*{Effective connectivity}
+\subsubsection{Effective connectivity}
 In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
@@
 
-\section{Methods}
+\section{Material and Methods}
 % Different adaptations of PC
@@
 Two changes have been made in this adaption with respect to modified PC as described by Spirtes et al \cite{Spirtes2000}.
+% TODO: line below is repetition of previous paragraph. Keep?
 In the algorithm as described in section ~\ref{sec:pseudocode}, the search for a separating sets for $X$ and $Y$ stops when such a set is found; instead, EMS continues the search for more separating sets of the same order.
@@
 
-\subsection*{Data acquisition \& analysis}
+\subsection{Experimental validation}
+% Section on how we plan on demonstrating the empirical value of research.
+% Full description from step 1 to final data set
+Our analysis consisted of two parts, corresponding to the two different parts of PC algorithm.
+Firstly we discerned structure by using brain resting-state time series fMRI data with structure PC.
+The generated structure is the same for all variants of PC discussed and proposed, as only the first separating set is used in finding unshielded triples.
+We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging.
+
+\subsection{Data acquisition}
 % TODO: vraag Max naar de details van de meetmethode en subject acquisition
-In order 
+The above analysis was conducted on the same dataset as in \cite{hinne2013structfunc}. %TODO check citation.
+The dataset consists of fMRI
+
 We will apply the mentioned methods on structural and functional data gathered from human subjects.
@@
 We can estimate functional connectivity from fMRI-measurement in combination with structural 
-networks according to a procedure currently submitted by Hinne et al. \cite{hinne2013submitted}.
+networks according to a procedure by Hinne et al. \cite{hinne2013structfunc}.
 To find effective connectivity, both structural networks and functional data is needed.

### 2013-12-09 docs/paper/paper.tex

@@
 
-We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000} can solve some of the difficulties effective connectivity suffers from.
+We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000}, named after \textbf{P}eter Spirtes and \textbf{C}lark Glymour, can solve some of the difficulties effective connectivity suffers from.
 
@@
 
+\subsection{Pseudocode}
+\label{sec:pseudocode}
+The first part of the algorithm finds an undirected graph $G$.
+For each pair of points which are not connected in this graph, it also finds a separating set.
+$\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph $G$.
+This graph changes during the algorithm and so does $\operatorname{Adjacencies} (X)$.
+
+\begin{algorithm}
+\caption{Structure PC}
+\begin{algorithmic}
+\Function{Structure\_PC}{$data$}
+\State connect all pairs of point in undirected graph $G$.
+\State $n \gets 0$
+\While{there are connected pairs $X$, $Y$ such that $n < |\operatorname{Adjecencies}(X)\setminus\{Y\}| $}
+  \ForAll{such connected pairs $X$, $Y$}
+    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
+      \Comment{(Fisher-Z-test on $data$)}
+      \State test whether $X$ and $Y$ are d-separated given $S$
+      \If{$X$ and $Y$ are d-separated given $S$}
+        \State remove the edge $X - Y$ from G
+        \State record $S$ as a separating set of $X$ and $Y$
+      \EndIf
+    \EndFor
+  \EndFor
+  \State $n \gets n+1$
+\EndWhile
+\Return $G$ and the separating sets
+\EndFunction
+\end{algorithmic}
+\label{code:emsstructure}
+\end{algorithm}
+
+The second part orients edges in this graph $G$ based on the separating sets.
+\begin{algorithm}
+\caption{Directional PC}
+\begin{algorithmic}
+\Function{Directional\_PC}{$G$, $sepsets$}
+  \State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
+  \ForAll{unshielded triples $<X, Y, Z>$}
+    \If{$Y$ is not in any separating set of $X$ and $Z \And X$ and $Z$ are dependent given $Y$}
+      \Comment{V-structure-rule}
+      \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
+    \EndIf
+  \EndFor
+  \Repeat
+    \ForAll{ordered, connected pairs $X$, $Y$}
+      \Comment{Rule 1}
+      \If{$X \osline Y$ or $X \sleftline Y$, and there is a node $Z$ such that $Z \srightarrow X$}
+        \State Orient $X \sline Y$ as $X \rightarrow Y$
+      \EndIf
+      \Comment{Rule 2}
+      \If{there is a directed path from $X$ to $Y$}
+        \State Orient $X \sline Y$ as $X \srightarrow Y$
+      \EndIf
+    \EndFor
+  \Until{no more edges are oriented}
+  \State \Return PDAG
+\EndFunction
+\end{algorithmic}
+\label{code:emsdirectional}
+\end{algorithm}
+
 \subsection{Experimental validation}
@@
 \section{Results}
+At first, we have applied modified PC on the strucural connectivity data of six subjects.
+To achieve stability, we have run the algorithm multiple times, in which the data was permuted randomly in each iteration.
+The avarage of all iterations was used as a measure of certainty, i.e.
+how consistent the algorithm is for each point of data.
+
+The results from the altorithms are presented in the form of matrices.
+In such a matrix, indices 1-45 represent the brain regions of the left hemisphere, regions 45-90 represent the right hemisphere, and regions 91-116 represent regions in the cerebellum.
+%TODO left or right hemisphere?
+
+Modified PC and EMS-PC do not differ in finding the structure.
+The results are presented in figure ~\ref{img:??} for all six subjects.
+%TODO aantal iteraties
+
+% Mod-PC
+
+%
+
+%
+
+%
+
+%
 

### 2013-12-09 docs/paper/paper.tex
Added images folder with mean structure image.

--- 2013-12-09 docs/paper/paper.tex
@@
 We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
-Two changes have been made in this adaption with respect to modified PC as described by Spirtes et al \cite{Spirtes2000}.
+Two changes have been made in this adaption with respect to modified PC as described by Spirtes et al \cite{spirtes2000}.
 % TODO: line below is repetition of previous paragraph. Keep?
@@
 
+As a final safe-guard, all unshielded triples are checked for unfaithfulness and marked as such if applicable using found separating sets.
+
 \subsection{Experimental validation}
@@
 % Full description from step 1 to final data set
-Our analysis consisted of two parts, corresponding to the two different parts of PC algorithm.
+Our analysis consisted of two parts, corresponding to the structural and directional parts of PC algorithm.
 Firstly we discerned structure by using brain resting-state time series fMRI data with structure PC.
@@
 We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging.
+Comparison between structures is done visually with an expert on brain anatomy.
+To increase the stability of found structures within subjects, we apply stability selection for different node permutations by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
+
+The second part of our analysis is about discerning orientation within the found structure.
+We applied the directionality rules of original PC, modified PC and EMS-PC.
 

### 2013-12-09 docs/paper/paper.tex
More on experimental validation.

@@
 To increase the stability of found structures within subjects, we apply stability selection for different node permutations by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
+Multi-subject comparison was done by averaging and inspecting standard deviation over subjects.
 
 The second part of our analysis is about discerning orientation within the found structure.
-We applied the directionality rules of original PC, modified PC and EMS-PC.
+We applied the directionality rules of original PC, modified PC and EMS-PC as mentioned above.
+Resulting connectivity patterns are visually compared by an expert for anatomical correctness.
+To compare between methods, we compare the consistency of methods to mark a specific connection as directional over permutations.
+Of specific interest here are non-symmetrical orientations.
+We again applied stability selection for a sufficiently large number of permutations per subject.
 
@@
 The above analysis was conducted on the same dataset as in \cite{hinne2013structfunc}. %TODO check citation.
-The dataset consists of fMRI
-
 We will apply the mentioned methods on structural and functional data gathered from human subjects.

### 2013-12-09 docs/paper/paper.tex

--- 2013-12-09 docs/paper/paper.tex
@@
 If they are found to be conditionally dependent, the triple is oriented as as $X \srightarrow Y \sleftarrow Z$.
+As a final safe-guard, all unshielded triples are checked for unfaithfulness and marked as such if applicable using found separating sets. %TODO wat bedoel je hiermee?
 
@@
 To achieve stability, we have run the algorithm multiple times, in which the data was permuted randomly in each iteration.
-The avarage of all iterations was used as a measure of certainty, i.e.
-how consistent the algorithm is for each point of data.
+The avarage of all iterations was used as a measure of certainty, i.e. how consistent the algorithm is for each point of data.
 
@@
 Modified PC and EMS-PC do not differ in finding the structure.
-The results are presented in figure ~\ref{img:??} for all six subjects.
+The results are presented in figure ~\ref{fig:struct_full_mean} as the avarage over all six subjects.
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/struct_full_mean_gray}
+  \caption{The avarage structure of all subjects}
+  \label{fig:struct_full_mean}
+\end{figure}
 %TODO aantal iteraties
+In this matrix, higher values represent strong certainties of the presence of a connection.
+
+The regions which are most clearly connected are the two hemispheres, and certain regions within the cerebellum.
 

### 2013-12-09 code/experiments/vna.m
functie van max geimporteerd die files maakt om in netdraw te stoppen (labels werken nog niet, die moeten uit een file in een cell van strings)

@@ -1,0 +1,25 @@
+function vna(filename, A, pos, labels)
+
+%load('../../data/pos.mat');
+%labels = fopen('../../data/AALlabels.txt');
+
+n = length(A);
+
+fid = fopen(filename, 'w');
+
+fprintf(fid, '*node data\n');
+fprintf(fid, 'id x y z name\n');
+
+for i=1:n
+     fprintf(fid, '%d %d %d %d %s\n', i, round(pos(i,:)), labels{i}); 
+end
+
+fprintf(fid, '*tie data\n');
+fprintf(fid, 'source target weight\n');
+
+[S, T, w] = find(A);
+
+fprintf(fid, '%d %d %f\n', [S T w]');
+
+fclose(fid);
+

### 2013-12-09 docs/paper/paper.tex
plaatje van structuur max toegevoegd. Ook nog een scriptje toegevoegd om netwerk-filetjes te maken die met netDraw te openen zijn, maar die lijkt daar nog onzin van te maken (ligt naar mijn weten aan netDraw)

@@
 \usepackage{cite}
-\usepackage{anysize}
+%\usepackage{anysize}
 %\usepackage[compact]{titlesec}
@@
 
-\setlength{\parindent}{0.0cm}
-\setlength{\parskip}{0.25cm}
+%\setlength{\parindent}{0.0cm}
+%\setlength{\parskip}{0.25cm}
 
@@
 \label{sec:pseudocode}
-The first part of the algorithm finds an undirected graph $G$.
-For each pair of points which are not connected in this graph, it also finds a separating set.
-$\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph $G$.
+
+The first part of the algorithm finds an undirected graph $G$ and is described in Algorithm ~\ref{code:emsstructure}.
+For each pair of points which are not connected in this graph, it also finds a separating set. $\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph $G$.
 This graph changes during the algorithm and so does $\operatorname{Adjacencies} (X)$.
+The second part orients edges in this graph $G$ based on the separating sets, as described in Algorithm ~\ref{code:emsdirectional}.
 
+%\par\vfill\break % Omdat pseudocode veel hspace nodig heeft :(
+%\advance\hsize by 8cm % Advance page height
+%\advance\hoffset by -4cm % Shift top margin
+%\floatevery{algorithm}{\setlength\hsize{15cm}}
+%\floatevery{algorithmic}{\setlength\hsize{15cm}}
+%TODO: meer hspace geven aan algorithms, alles hierboven werkte niet
 \begin{algorithm}
-\caption{Structure PC}
+\caption{Structure EMS-PC}
 \begin{algorithmic}
@@
     \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
+      \State test whether $X$ and $Y$ are d-separated given $S$
       \Comment{(Fisher-Z-test on $data$)}
-      \State test whether $X$ and $Y$ are d-separated given $S$
       \If{$X$ and $Y$ are d-separated given $S$}
@@
 
-The second part orients edges in this graph $G$ based on the separating sets.
 \begin{algorithm}
-\caption{Directional PC}
+\caption{Directional EMS-PC}
 \begin{algorithmic}
@@
 \end{algorithm}
+
+%\par\vfill\break % Omdat pseudocode veel hspace nodig heeft :(
+%\advance\hsize by -8cm % Retur n old margings and page height
+%\advance\hoffset by 4cm % Return old margings and page height
 
@@
 The results from the altorithms are presented in the form of matrices.
+A connection from region $n$ to region $m$ is represented by point $(n,m)$ or $n,m$ in the matrix.
+For strucural data, this matrix is therefore symmetrical.
+In the directional data, point $(n,m)$ denotes a directed edge from $n$ to $m$.
 In such a matrix, indices 1-45 represent the brain regions of the left hemisphere, regions 45-90 represent the right hemisphere, and regions 91-116 represent regions in the cerebellum.
@@
 Modified PC and EMS-PC do not differ in finding the structure.
-The results are presented in figure ~\ref{fig:struct_full_mean} as the avarage over all six subjects.
+The results are presented in figure ~\ref{fig:struct_full_mean} as the avarage over all six subjects, with 300 iterations per subject.
 \begin{figure}[h!]
@@
 \end{figure}
-%TODO aantal iteraties
 In this matrix, higher values represent strong certainties of the presence of a connection.
 
-The regions which are most clearly connected are the two hemispheres, and certain regions within the cerebellum.
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/structure_max}
+  \caption{The avarage structure of all subjects, according to the DWI-measurements of Hinne et al.}
+  \label{fig:struct_mean_max}
+\end{figure}
+
+One notable difference between the results found in Hinne et al.
+are that connections between the the two hemispheres are clearly visible as two diagonal lines.
+Connections within the hemispheres are less clearly visible, as well as connections within the cerebellum.
 

### 2013-12-09 code/experiments/doodle.m

@@ -1,5 +1,5 @@
-% perm = [1:2:90 2:2:90 91:116];
-perm = 1:116;
+perm = [1:2:90 2:2:90 91:116];
+%perm = 1:116;
 
 G_avg = mean(Gs,4);
 G_avg = G_avg(:,perm,perm);
@@ -48,26 +48,45 @@ imagesc(Subj_avg - Subj_stddev); colorbar; axis square; title('mean-std PDAG');
 % Plot per subject mean
 h = figure; colormap(cmap);
 for i=1:6
-    subplot(2,2,1);
+    subplot(2,3,1);
     imagesc(squeeze(PDAG_avg(i,:,:)));
     title(sprintf('PDAG - Subj %d',i));
     colorbar;
     axis square;
-    subplot(2,2,2);
+    subplot(2,3,2);
     imagesc(squeeze(PDAG_avg(i,:,:)) - squeeze(PDAG_avg(i,:,:))');
     title(sprintf('ASSYM PDAG - Subj %d',i));
     colorbar;
     axis square;
-    subplot(2,2,3);
+    subplot(2,3,4);
     imagesc(squeeze(PDAG_std(i,:,:)));
     title(sprintf('STD PDAG - Subj %d',i));
     colorbar;
     axis square;
-    subplot(2,2,4);
+    subplot(2,3,5);
     imagesc(squeeze(PDAG_avg(i,:,:)) - squeeze(PDAG_std(i,:,:)));
     title(sprintf('mean - std PDAG - Subj %d',i));
     colorbar;
     axis square;
+    
+    
+    subplot(2,3,3);
+    imagesc(squeeze(G_avg(i,:,:)));
+    title(sprintf('Structure - Subj %d',i));
+    axis square;
+    colorbar;
+    
     pause;
 end
 close(h);
+
+% Plot data of Max
+load('data.mat');
+G_max = squeeze(mean(Gs, 1));
+G_max = G_max(perm, perm);
+figure;
+imagesc(G_max);
+title('structure based on DWI');
+colormap 'gray';
+colorbar;
+axis square;

### 2013-12-09 code/experiments/doodle.m
plaatje van PDAGs toegevoegd (nog niet geschikt voor het verslag, maar handig om thuis te kunnen gebruiken als ik matlab daar niet werkende zou krijgen

@@ -48,33 +48,26 @@ imagesc(Subj_avg - Subj_stddev); colorbar; axis square; title('mean-std PDAG');
 % Plot per subject mean
 h = figure; colormap(cmap);
 for i=1:6
-    subplot(2,3,1);
+    subplot(2,2,1);
     imagesc(squeeze(PDAG_avg(i,:,:)));
     title(sprintf('PDAG - Subj %d',i));
     colorbar;
     axis square;
-    subplot(2,3,2);
+    subplot(2,2,2);
     imagesc(squeeze(PDAG_avg(i,:,:)) - squeeze(PDAG_avg(i,:,:))');
     title(sprintf('ASSYM PDAG - Subj %d',i));
     colorbar;
     axis square;
-    subplot(2,3,4);
+    subplot(2,2,3);
     imagesc(squeeze(PDAG_std(i,:,:)));
     title(sprintf('STD PDAG - Subj %d',i));
     colorbar;
     axis square;
-    subplot(2,3,5);
+    subplot(2,2,4);
     imagesc(squeeze(PDAG_avg(i,:,:)) - squeeze(PDAG_std(i,:,:)));
     title(sprintf('mean - std PDAG - Subj %d',i));
     colorbar;
     axis square;
-    
-    
-    subplot(2,3,3);
-    imagesc(squeeze(G_avg(i,:,:)));
-    title(sprintf('Structure - Subj %d',i));
-    axis square;
-    colorbar;
     
     pause;
 end

### 2013-12-09 docs/paper/paper.tex
Wrote paragraph on data acquisition.

@@
 \usepackage[english]{babel}
-%\usepackage[cm]{fullpage}
+\usepackage[cm]{fullpage}
 \usepackage{cite}
-%\usepackage{anysize}
+\usepackage{anysize}
 %\usepackage[compact]{titlesec}
@@
 % Margins & Headers
-%\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
-%\columnsep 0.4in
-%\footskip 0.4in 
-%\usepackage{changepage}
+\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
+\columnsep 0.4in
+\footskip 0.4in 
+\usepackage{changepage}
 
@@
 
-%\setlength{\parindent}{0.0cm}
-%\setlength{\parskip}{0.25cm}
+\setlength{\parindent}{0.0cm}
+\setlength{\parskip}{0.25cm}
 
@@
 A causal discovery method uses data about events to determine which of these events are correlated, and if possible, it determines which event causes which.
-The definition of a causal relation is not trivial \cite{?} and the relevant principles will be covered in this section.
+The definition of a causal relation is not trivial \cite{} and the relevant principles will be covered in this section.
 In our context, the causal discovery methods which have been used aim at inferring a Bayesian network.
@@
 \subsection{Data acquisition}
-% TODO: vraag Max naar de details van de meetmethode en subject acquisition
-The above analysis was conducted on the same dataset as in \cite{hinne2013structfunc}. %TODO check citation.
-We will apply the mentioned methods on structural and functional data gathered from human subjects.
-For this we are largely dependent on data collected by others.
-Hinne et al. has provided fMRI time-series data and structural network data as used and derived in \cite{hinne2013}, of which six subjects were selected at random.
-We can estimate functional connectivity from fMRI-measurement in combination with structural 
-networks according to a procedure by Hinne et al. \cite{hinne2013structfunc}.
-To find effective connectivity, both structural networks and functional data is needed.
+No explicit data acquisition was performed, instead we performed our analysis on the same data as treated by Hinne et al.
+in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
+
+Paraphrasing: ``six healthy volunteers were scanned for resting-state functional data and diffusion-weighted images using a Siemens Magnetom Trio 3 T system at the Donders Centre for Cognitive Neuroimaging at Radboud University in The Netherlands.
+Resting-state fMRI data were acquired at 3 Tesla by using a multi-echo echo-planar imaging (ME-EPI) sequence (voxel size 3.5 mm isotropic, matrix size 64x64, TR=2000 ms, TEs=6.9, 16.2, 25, 35 and 45 ms, 39 slices, GRAPPA factor 3, 6/8 partial Fourier).
+A total of 1030 volumes were obtained.
+Diffusion weighted images were obtained by using the DWI protocol (voxel size 2.0 isotropic, matrix size 110x110, TR=13.00 ms, TE=101 ms, 70 slices, 256 directions at b=1500 s/mm$^2$).''
+
+We did not perform explicit (pre-)processing on the functional and diffusion imaging data.
+Instead, we used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as (pre-)processed by Hinne et al. \cite{hinne2013}.
 
 \section{Results}
-At first, we have applied modified PC on the strucural connectivity data of six subjects.
+At first, we have applied modified PC on the structural connectivity data of six subjects.
 To achieve stability, we have run the algorithm multiple times, in which the data was permuted randomly in each iteration.

### 2013-12-09 docs/paper/paper.tex
Hier en daar wat geschaafd aan tekst.

@@
 
-We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000}, named after \textbf{P}eter Spirtes and \textbf{C}lark Glymour, can solve some of the difficulties effective connectivity suffers from.
+We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000}, named after its authors \textbf{P}eter Spirtes and \textbf{C}lark Glymour, can solve some of the difficulties effective connectivity suffers from.
 
@@
 An approach for finding effective connectivity may be found in the domain of causal discovery.
-A causal discovery method uses data about events to determine which of these events are correlated, and if possible, it determines which event causes which.
-The definition of a causal relation is not trivial \cite{} and the relevant principles will be covered in this section.
-In our context, the causal discovery methods which have been used aim at inferring a Bayesian network.
-This is done by calculating statistical (in)dependencies between stochastic variables.
+A causal discovery method uses data about events to determine which of these events are correlated, and additionally, it attempts to determine which events cause which.
+The definition of a causal relation is not trivial \cite[p.20]{spirtes2000} and the relevant principles will be covered in this section.
+In our context, the causal discovery methods which have been used aim at inferring a Bayesian network or Directed Acyclic Graph (DAG).
+These can be interpreted from a probabilistic and a causal perspective.
+In our case, we consider graphs that also include non-directed edges, a so called Partial DAG (PDAG).
 
-Such a Bayesian network can be modelled as a graph with edges which may be directed or undirected.
-Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
-If there is an edge between two variables $X$ and $Y$, this is denoted as $X - Y$.
+Under the causal interpretation, a PDAG $G$ each variable is represented by a node and correlations between variables are represented by edges between those nodes.
+If such an edge exists between two variables $X$ and $Y$, this is denoted as $X - Y$.
 A causal relation is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
@@
 An ancestor is an (possibly indirect) cause of its descendant.
-A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
-Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
-In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
-As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
-Image {plaatje!} shows an example of a Bayesian network.
+
+% TODO: Add example of a Bayesian network.
+%Image {plaatje!} shows an example of a Bayesian network.
+
 A causal discovery method attempts at finding a real-world Bayesian network.
-But real-world situations like a brain are immensely complex and contain an unimaginable large number of atoms, which could all function as seperate variables.
+The human brain is immensely complex and contains a computationally intractable amount of neurons to model.
 As such, we represent our real-world, underlying Bayesian network with a much simpler model.
-Such a model contains a reduced number of variables to be able to analyse and visualize the data.
+Such a model contains a reduced number of variables in order to remain tractable and to visually interpret model results.
 
@@
 Instead, we used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as (pre-)processed by Hinne et al. \cite{hinne2013}.
+% K is de MAP estimate uit een G-Wishart distributie.
+Dat wil zoveel zeggen als: een distributie over alle precisiematrices die nullen hebben op de plekken waar G een 0 heeft (Hastie, Trevor; Tibshirani, Robert; Friedman, J.
+(2009).
+The Elements of Statistical Learning, voor de ge•nteresseerden).
 
@@
 
-One notable difference between the results found in Hinne et al.
-are that connections between the the two hemispheres are clearly visible as two diagonal lines.
+One notable difference between the results found in Hinne et al. are the strong connections between homological areas, visible as two diagonal lines.
+This is a known shortcoming of diffusion weighted imaging, as these connections are typically too long to measure correctly.
+These connections most likely correspond to the corpus callosum.
 Connections within the hemispheres are less clearly visible, as well as connections within the cerebellum.
@@
 \section{Discussion}
+% Further research
+A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
+The theory used here does not incorporate a temporal component and 
+Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
+%In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
+%As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
+As contemporary measuring techniques are too coarse and slow to model a brain at one point in time, correctly modelling variables over multiple points in time - opposing the acyclic assumption - would provide an interesting direction of further research.
+
+The
 
 \section{Conclusion}
+
+\section{Acknowledgements}
+We would like to thank Max Hinne and Tom Claassen, for without them this research would not have been viable.
+They have provided us with ideas, valuable discussion and the data to perform actual research.
 

### 2013-12-10 docs/paper/paper.tex
small comment addition in discussion.
heel wat plaatjes en toegevoegd, met wat tekst er omheen. Plaatsing van de plaatjes door latex is nog lelijk.

--- 2013-12-09 docs/paper/paper.tex
@@
 \section{Results}
-At first, we have applied modified PC on the strucural connectivity data of six subjects.
+At first, we have applied modified PC on the structural connectivity data of six subjects.
 To achieve stability, we have run the algorithm multiple times, in which the data was permuted randomly in each iteration.
@@
 %TODO left or right hemisphere?
+Modified PC and EMS-PC do not differ in finding the structure.
+The results are presented in figure ~\ref{fig:struct_avg} as the avarage over all six subjects, with 300 iterations per subject.
+The results of Hinne et al. are shown in ~\ref{fig:struct_max} for comparison.
 
-Modified PC and EMS-PC do not differ in finding the structure.
-The results are presented in figure ~\ref{fig:struct_full_mean} as the avarage over all six subjects, with 300 iterations per subject.
 \begin{figure}[h!]
@@
   \includegraphics{images/struct_full_mean_gray}
-  \caption{The avarage structure of all subjects}
-  \label{fig:struct_full_mean}
+  \caption{The avarage structure of all subjects, found with the PC-algorithm}
+  \label{fig:struct_avg}
 \end{figure}
@@
   \caption{The avarage structure of all subjects, according to the DWI-measurements of Hinne et al.}
-  \label{fig:struct_mean_max}
+  \label{fig:struct_max}
 \end{figure}
 
-One notable difference between the results found in Hinne et al.
-are that connections between the the two hemispheres are clearly visible as two diagonal lines.
+One notable difference between the results found in Hinne et al. is the strong connections between homological areas, visible as two diagonal lines.
+This is a known shortcoming of diffusion weighted imaging, as these connections are typically too long to measure correctly.
+These connections most likely correspond to the corpus callosum, as each region is connected with its counterpart in the other hemisphere.
 Connections within the hemispheres are less clearly visible, as well as connections within the cerebellum.
 
-% Mod-PC
+The second step of the PC-algorithm finds directions based on this structure.
+In Figures ~\ref{fig:pdag_avg_mod} and ~\ref{fig:pdag_avg_ems}, the results for both modified PC and EMS-PC are presented.
+The figures look very similar to Figure ~\ref{fig:struct_avg}, and look strongly symmetric.
+In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.
 
-%
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/PDAG_avg_mod}
+  \caption{The PDAG found with modified PC, in which a high value denotes a high certainty of an arrow. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg_mod}
+\end{figure}
 
-%
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/PDAG_avg_expl}
+  \caption{The PDAG found with EMS-PC, in which a high value denotes a high certainty of an arrow. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg_ems}
+\end{figure}
 
-%
+Another representation of this data has been used to emphasize the assymetric parts of Figures ~\ref{fig:pdag_avg_mod} and ~\ref{fig:pdag_avg_ems}.
+To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
+These results are presented in Figures ~\ref{fig:pdag_avg_antisymmetric_mod} and ~\ref{fig:pdag_avg_antisymmetric_ems}.
 
-%
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/PDAG_avg_antisymmetric_mod}
+  \caption{The assymetric parts of the PDAG found with modified PC, in which a high value denotes a high certainty of an one-directional arrow. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg_antisymmetric_mod}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/PDAG_avg_antisymmetric_expl}
+  \caption{The assymetric parts of the PDAG found with EMS-PC, in which a high value denotes a high certainty of an one-directional arrow. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg_antisymmetric_ems}
+\end{figure}
+
+% iets over:
+% hoogte van avg_antisymmetric, hoger bij EMS dus meer richting, maar bij beiden niet zo consistent over proefpersonen
+% verschillen tussen beide hemispheres: niet zo logisch lijkt me, hemisphers zouden intern ongeveer hetzelfde moeten functioneren
+% veel hoge waarden in 'blokjes' rond diagonaal, interessante dingen in de hemispheres?
+% per proefpersoon maxima van ~0.5, maar een veel lager avarage over proefpersonen dus weinig consistentie
 

### 2013-12-10 code/experiments/doodle.m
visualizatie aan doodle toegevoegd om antisymmetrische verbindingen duidelijk in greyscale weer te geven

@@ -73,6 +73,50 @@ for i=1:6
 end
 close(h);
 
+
+% SOME EXTRA SCRIPT-LIKE THINGYS TO GENRATE IMAGES IN BLACK-WHITE
+%Plot PDAGs of all subjects
+h = figure; colormap ('gray');
+for i = 1 : 6
+    subplot(2,3,i);
+    imagesc(squeeze(PDAG_avg(i,:,:)/2));
+    axis square;
+    colorbar;
+end
+
+%Plot avg PDAG of all subjects
+h = figure; colormap('gray');
+imagesc(Subj_avg/2);
+axis square;
+colorbar;
+
+%Plot of antisymmetric of all subjects
+h = figure; colormap('gray');
+for i = 1:6
+    subplot(2,3,i);
+    imagesc((squeeze(PDAG_avg(i,:,:)) - squeeze(PDAG_avg(i,:,:))')/2);
+    axis square;
+    colorbar;
+end
+
+%Plot avg of antisymmetric of all subjects
+h = figure; colormap ('gray');
+imagesc(abs(antisym_subj_avg/2));
+axis square;
+colorbar;
+
+%Plot of a selection of antisymmetric of all subjects
+h = figure; colormap('gray');
+selectionIndices = [46:90];
+for i = 1:6
+    subplot(2,3,i);
+	res = (squeeze(PDAG_avg(i,:,:)) - squeeze(PDAG_avg(i,:,:))')/2;
+	res(res < 0) = 0;
+    imagesc(res(selectionIndices, selectionIndices));
+    axis square;
+    colorbar;
+end
+
 % Plot data of Max
 load('data.mat');
 G_max = squeeze(mean(Gs, 1));

### 2014-01-14 docs/paper/paper final.tex
opzetje voor 'paper final.tex'

--- 2013-12-09 docs/paper/paper.tex
@@
 \usepackage{anysize}
+\usepackage{setspace}
 %\usepackage[compact]{titlesec}
@@
 \def \oleftline {-\hspace{-0.16em}\circ}
-\def \oline {\circ\hspace{-0.38em}-\hspace{-0.4em}\circ}
+\def \oline {\circ\hspace{-0.38em}-\hspace{-0.2em}\circ}
 \def \srightarrow {\text{\textasteriskcentered}\hspace{-0.62em}\rightarrow}
@@
 
+% ### Section Formatting ###
+\usepackage[compact]{titlesec}
+\titlespacing*{\section}{4pt}{*0}{4pt}
+
+% ### Paragraph Formatting ###
+\makeatletter
+\renewcommand{\paragraph}{%
+  \@startsection{paragraph}{4}%
+  {\z@}{0.5ex \@plus 1ex \@minus .2ex}{-1em}%
+  {\normalfont\normalsize\bfseries}%
+}
+\makeatother
+
 % Title
-\title{\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity}}
+\title{\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}}
 \author{
@@
 \setlength{\parindent}{0.0cm}
-\setlength{\parskip}{0.25cm}
+\setlength{\parskip}{3mm plus2mm minus1.5mm}
 
 \begin{abstract}
-% TODO: Add abstract. Look at slides Peter to see what is asked.
+% TODO: first state what you try to do, then how you did it, then what it brought you
+%In this work we applied PC algorithm on resting-state data of healthy human subjects.
+%Additionally we propose a variant of PC algorithm that poses weaker assumptions on the model, better fitting the underlying model.
+%We find that detecting structure homological areas in the brain works better with PC algorithm than most conventional diffusion weighted imaging techniques.
+%In the area of directionality and causality, more research is needed.
 \end{abstract}
 
-\section{Introduction}
+\section{Background}% Types of brain connectivity [optional] [Ramon: lijkt me wel belangijk, geeft ook het belang aan van richting zoeken, hoeft niet lang]
+\subsection{Brain connectivity}
+%Functional
+%Structural
+%Effective
 
-\paragraph{Motivation}
-The brain and particularly the human brain have been studied for hundreds of years.
-Today, the secrets of our brains still are one of the most sought-after.
-The techniques however have changed.
-We have come a long way since the time of Phrenology, the pseudoscience based attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
-Advanced measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview without opening a single skull.
+\section{Causal inference}
+% Causal discovery theory
+%Causal Markov Condition
+%Causal faithfulness
+%d-separation
+%V-structure
+%Causal sufficiency
+%Latent variables
 
-Of particular interest are causal relations between brain regions and cognitive and bodily functions.
-The question if brain region X is functionally or structurally connected to brain region Y and whether these are causally related is still highly relevant; it is the main question of the present-day scientific field of brain connectivity.
-Understanding brain structure implies understanding more about the brain as a whole.
+\section{PC algorithm}
+%Algorithm, assumptions and explanation [Ramon: standaard PC een naam geven en consistent gebruiken, bijv. classical PC?]
+%Performance
+%Stability and faithfulness problems
+%Conservative PC extension
 
-Not surprisingly, brain connectivity also finds applications in medicine.
-Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity.
-%TODO: Citation needed.
-Better methods for analysing connectivity could lead to more insight in these diseases.
-
-The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
-One type of connectivity analysis involves finding how brain region X causes what region Y does. This is called \emph{effective connectivity}.
-More concretely: how does the measured fMRI signal, representing an underlying neuronal process, determine what other neuronal processes are activated.
-Research regarding effective is relatively new and its application still faces many challenges that require further research \cite{ramsey2010}.
-The directed character in effective connectivity seems similar to traditional causal relations.
-In what follows, we suggest to use causal discovery methods for deriving effective connectivity.
-
-\paragraph{Problem statement}
-This project seeks to utilise the strengths of current state-of-the-art causal discovery methods by applying them to resting state brain fMRI data in order to find new applicable methods for determining causal patterns in the human brain.
-
-The work by \cite{ramsey2010} addresses several problems causal analysis in fMRI analysis suffers from.
-Methods in causal discovery have high computational costs, making it nearly impossible to calculate networks with more than hundreds of nodes.
-Another important fact is that fMRI Analysis indirectly measures brain activity.
-Brain models that do not account for possible latent - indirect - sources within the brain or the shortcomings of fMRI may suffer from noise, fail to capture underlying patterns or draw erroneous conclusions.
-Solutions to this particular problem have been proposed that do account for latent sources \cite{ramsey2010, waldorp2011}.
-Another problem is the strong diversity between subjects.
-Every brain is inherently different, making it non-trivial to combine subject data or even draw conclusions across subjects.
-Even within a single brain it changes how different areas react over time.
-
-Several methods for finding effective connectivity already exist \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
-However, using more generally applicable causal discovery methods has not been elaborated on much.
-Also, none of these provide a measure of uncertainty.
-It is a fact that there are errors in any graph produced.
-This is inherent to the methods and the coarseness of the measurements.
-As there is no standard baseline to compare results with, a measure of uncertainty would be highly preferable.
-A framework for such a probabilistic approach is introduced in \cite{claassen2012}.
-
-We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000}, named after \textbf{P}eter Spirtes and \textbf{C}lark Glymour, can solve some of the difficulties effective connectivity suffers from.
-
-% Structure of the rest of proposal
-In the remainder of this proposal we briefly discuss some necessary background knowledge on connectivity and methods, we then discuss our strategy and finally we present a time-plan overview.
-
-\section{Theory}
-
-\subsection{Causal discovery}
-An approach for finding effective connectivity may be found in the domain of causal discovery.
-A causal discovery method uses data about events to determine which of these events are correlated, and if possible, it determines which event causes which.
-The definition of a causal relation is not trivial \cite{} and the relevant principles will be covered in this section.
-In our context, the causal discovery methods which have been used aim at inferring a Bayesian network.
-This is done by calculating statistical (in)dependencies between stochastic variables.
-
-Such a Bayesian network can be modelled as a graph with edges which may be directed or undirected.
-Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
-If there is an edge between two variables $X$ and $Y$, this is denoted as $X - Y$.
-A causal relation is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
-It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$. 
-When there is a directed path $X \rightarrow Z_1 \rightarrow Z_2 \dots \rightarrow Y$, $X$ is said to be an ancestor of $Y$ and $Y$ is a descendant of $X$.
-An ancestor is an (possibly indirect) cause of its descendant.
-A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
-Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
-In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
-As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
-Image {plaatje!} shows an example of a Bayesian network.
-A causal discovery method attempts at finding a real-world Bayesian network.
-But real-world situations like a brain are immensely complex and contain an unimaginable large number of atoms, which could all function as seperate variables.
-As such, we represent our real-world, underlying Bayesian network with a much simpler model.
-Such a model contains a reduced number of variables to be able to analyse and visualize the data.
-
-An important principle for our approach is that of conditional dependence.
-The dependency of two variables might depend on the value of a third variable (or possibly even more variables).
-This principle is best demonstrated with an example.
-When two variables, such as two switches being turned off or on, both cause a light bulb to glow, the two switches might not be dependent.
-Whether one switch is turned on or off does not cause the other to switch on or off.
-However, when the light bulb is glowing, the switches do become dependent.
-If one switch is off, the other one is probably on.
-Or in other words, one switch being turned on decreases the chance of the other switch being turned on.
-In this case, we conclude that the switches are conditionally dependent given that the light bulb is turned on.
-More generally, two variables $X$ and $Y$ can be conditionally dependent on a set of other variables $S$. % het juiste latex-symbooltje vinden :(
-
-A similar principle is conditional independence.
-If two variables are generally dependent, they can still be independent given a third variable.
-An example of this is when a variable $X$ has two children $Y$ and $Z$.
-$Y$ and $Z$ are generally dependent, as they share a common cause or \emph{confounder} or \emph{confounder} $X$;
-When $X$ does not vary, $Y$ and $Z$ lose their confoundernd they lose their dependency.
-As such, $Y$ and $Z$ are independent given \{$X$\}.
-In real-world data, independence can be found with methods such as the Fisher-Z-test.
-
-We define a trail as a set of edges which would make up a path if those edges would be undirected.
-We can now define directional separation, or d-separation.
-Two nodes $X$ and $Y$ are d-separated given a set of nodes $S$ (with $X, Y \notin S$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
-\begin{itemize}
-\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $S$.
-\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $S$.
-\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $S$.
-\end{itemize}
-If $X$ and $Y$ are d-separated given $S$, $S$ is called a separating set of $X$ and $Y$.
-Every two independent variables have a separating set; if they are unconditionally independent, this is the empty  set.
-
-\subsection{Assumptions about the structure}
-In theory, every deterministic real-world situation must be described by some Bayesian model, as all situations conform to causality.
-But the underlying real-world data such as the measured brain activity used in this analysis usually does not conform to a simple model.
-Assumption can be made about the model to make it easier to infer it, but this increases the risk of drawing false conclusions.
-
-One assumption that can be made about the model is that of causal sufficiency. 
-When it is assumed that the model fully describes the real network, so-called V-structures can be concluded. 
-When there is a structure $X - Y - Z$, in which $X$ and $Z$ are not connected, the only possible orientation of the connections is $X \rightarrow Y \leftarrow Z$. % TODO: extra uitleg, d-separation?
-An approach to finding a model is therefore to find all V-structures and orient them in this way.
-
-However, it is very unlikely that all events which are relevant for the causal graph are measured; 
-even with high temporal and spatial resolutions, the entire fysiology of the brain is too complex to capture.
-Latent variables are nodes which are present in the underlying Bayesian graph, but which are not measured.
-The complex reality might be very similar to a simple model based on the measured data, but it might also be very different.
-This correctness of the data is called causal faithfulness.
-When there is a structure $X \leftarrow L \rightarrow Y$ in which $L$ is a latent variable, an edge between $X$ and $Y$ might be found while the these two nodes have no direct ancestor/descendant-relation between them;
-They might be related through a confounder, which is indicated as $X \leftrightarrow Y$.
-As such, for a V-structure for which the underlying Bayesian network has latent variables, a directed edge $X \rightarrow Y$ or $X \leftarrow Y$ would be erroneous. 
-As such, the way of describing V-structures might not draw false conclusions, depending on the causal faithfulness. 
-A weaker conclusion is that a V-structure should be oriented as $X \orightarrow Y \oleftarrow Z$, in which the circles at the end of an edge mean that the form of this end is not yet know; an arrow $\orightarrow$ might be a simple arrow $\rightarrow$ or a double arrow $\leftrightarrow$.
-
-\subsection{The PC-algorithm}
-
-\subsection{Informal description}
-The PC algorithm as used for these experiments is the version which is modified to handle latent variables, as also described by Spirtes, Glymour and Scheines \cite{spirtes2000}, pages 165-167.
-For finding directed edges, the PC-algorithm first finds the undirected structure.
-For each two nodes $X$ and $Y$ which are not neighbours in this structure, it also finds a minimal separating set, which is the smallest set acting as a separated set.
-This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.
-
-The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ (the number of neighbours decreases while edges are removed), using the fisher-Z-test.
-If such an independence is found, this subset is a separating set of $X$ and $Y$ and they are not connected.
-If no such an independence is found, they are connected.
-To find the minimal separating set, all subsets of neighbours are traversed in order of size; the first separating set which is found, is the minimal one.
-
-After the structure and separating sets have been found, the PC-algorithm attempts to orient the edges.
-After Spirtes et al, we use a star-symbol at the end of an edge to denote that any mark may be present: an arrowhead, an "o" or an empty mark.
-When an edge is oriented as an edge with a star, this denotes that this end is not changed.
-A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm.
-Multiple variations of this algorithm are described by Spirtes, Glymour and Scheines \ref{?}.
-For this analysis, we have used the "modified pc algorithm" which does not assume causal sufficiency, and thus is able to handle latent variables.
-
-The first of these rules attempts to find and orient unshielded triples.
-A triple $<X, Y, Z>$ is considered unshielded if $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent.
-As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
-If $Z$ is not in $S$, then the only possible causal relation within this structure can be that of a so-called V-structur; the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.
-
-After this rule which finds V-structures, the other rules are applied in no particular order. These rules are based on arrows that have already been found.
-\begin{itemize}
-\item If there is an edge $X \sline Y$, and if there is a directed path from $X$ to $Y$, then $X \sline Y$ can be oriented as $X \srightarrow Y$.
-\item If there are edges $X \srightarrow Y$ and $Y \sleftline Z$, if the latter is not an edge $Y \sleftarrow Z$, and if $X$ and $Z$ are not connected, orient $Y \sline Z$ as $Y \rightarrow Z$. 
-\end{itemize}
-
-\subsection{Brain Connectivity}
-
-%Question: should this be included?
-\subsubsection{Functional connectivity}
-Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
-This gives an insight in the organisation of the brain.
-Functional connectivity is strongly time-dependent, as activity changes rapidly providing and fMRI has a low temporal resolution.
-This means measured dependencies can be the result of statistical noise.
-Functional connectivity can be deduced by fMRI measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
-Resting state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity.
-
-\subsubsection{Structural connectivity}
-Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
-This is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
-This intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
-Structural connectivity is less time dependent as it involves mappings of anatomical connections opposing temporal activity patterns.
-
-Hinne et al. \cite{hinne2013} proposed a Bayesian method for estimating structural networks based on Diffusion Weighted MRI (DWI) and probabilistic Tractography.
-The location of white-matter tracts does not become immediately clear from imagery alone.
-Other methods rely on thresholds to estimate location of said tracts.
-The proposed method results in a measure of uncertainty about where a hypothesised connection will terminate and so provides a clearly interpretable network structure as result.
-
-\subsubsection{Effective connectivity}
-In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
-It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
-Effective connectivity indicates which brain regions stimulate other regions.
-Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
-It is possible to infer effective connectivity from structural and functional connectivity \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
-
-Research has been done on inferring effective connectivity directly with several methods.
-The most renowned methods are Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modeling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
-These are all methods dedicated to this problem.
-These methods also suffer from similar problems of which perhaps the most important is scaling.
-Most of these methods can only handle tens of nodes in reasonable time, which becomes problematic in realistic situations with up to tens of thousands of nodes, ideally.
-
-\section{Material and Methods}
-% Different adaptations of PC
-The original PC algorithm assumes causal sufficiency, excluding the existence of common confounders between nodes.
-This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
-Also measuring methods are not yet advanced enough to measure the full detail of the brain, thus skipping over details.
-In the perspective of causal inference, these missing areas can also appear as confounders.
-
-% Modified PC
-\paragraph{Modified PC}
-To compensate for this, an adaptation of PC algorithm by Spirtes et al. returns a Markov equivalence class, rather than one specific instance of it.
-In literature it is this algorithm that is usually cited under the name PC Algorithm.
-This also makes PC algorithm more stable as the number of Markov equivalence classes is less than the number of times data can be permuted.
-However the order of vertex removal still plays a key role in the end result.
-This causal-insufficient version is named Modified PC.
-
-% Conservative PC
-\paragraph{Conservative PC}
-Step x in PC algorithm assigns shielded triples based on a single separating set. % TODO: Add step
-Simply put, it is assumed the single found separating set is enough to show conditional independence. 
-Ramsey et al. \cite{ramsey2012} introduce a weaker but still sufficient assumption for directing unshielded triples.
-Through this weaker but still adjacency-faithful assumption, the method is more cautious than PC algorithm in drawing unambiguous conclusions on causal orientations.
-That is why this method is named Conservative PC (CPC).
-The resulting algorithmic changes provide better results and is able to mark incorrectly unshielded triples as unfaithful.
-Sadly, CPC requires a significant amount of additional conditional independence tests, exponential in the largest branching degree of the tested structure.
-
-% Multiple sepsets
-\paragraph{Multiple Separating Sets}
-The idea of CPC, to make a weaker assumption on graph faithfulness, is very potential.
-Instead of doing additional tests, one could attempt to determine causal unfaithfulness with less dependency tests.
-An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found, to test all other vertex subsets of the same order as well.
-This finds all separating sets of the lowest order without increasing complexity.
-Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
-By having more than one separating set, this constraint is strengthened.
-
-% Unfaithfulness test
-One could use these additional separating sets to test oriented unshielded triples for unfaithfulness, similar to CPC.
-Apart from serving as a sanity check, this could serve as the basis of a repair algorithm for errors made by PC.
-
-% Explicit test
-A more explicit way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
-If so, the graph is likely to be more faithful to its modeled distribution.
-
-\paragraph{EMS-PC}
-We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
-Two changes have been made in this adaption with respect to modified PC as described by Spirtes et al \cite{spirtes2000}.
-% TODO: line below is repetition of previous paragraph. Keep?
-In the algorithm as described in section ~\ref{sec:pseudocode}, the search for a separating sets for $X$ and $Y$ stops when such a set is found; instead, EMS continues the search for more separating sets of the same order.
-
-Also, the condition to orient $X - Y - Z$ according to the V-structure-rule has been made more strict.
-As multiple separating sets have been found instead of only one, it needs to be checked whether $Y$ is in not in \emph{any} of those sets before the V-structure can be oriented.
-Also, in addition to the original condition, $X$ and $Z$ are tested for d-separation given $\{Y\}$.
-If they are found to be conditionally dependent, the triple is oriented as as $X \srightarrow Y \sleftarrow Z$.
-As a final safe-guard, all unshielded triples are checked for unfaithfulness and marked as such if applicable using found separating sets. %TODO wat bedoel je hiermee?
-
-\subsection{Pseudocode}
-\label{sec:pseudocode}
-
-The first part of the algorithm finds an undirected graph $G$ and is described in Algorithm ~\ref{code:emsstructure}.
-For each pair of points which are not connected in this graph, it also finds a separating set. $\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph $G$.
-This graph changes during the algorithm and so does $\operatorname{Adjacencies} (X)$.
-The second part orients edges in this graph $G$ based on the separating sets, as described in Algorithm ~\ref{code:emsdirectional}.
-
-%\par\vfill\break % Omdat pseudocode veel hspace nodig heeft :(
-%\advance\hsize by 8cm % Advance page height
-%\advance\hoffset by -4cm % Shift top margin
-%\floatevery{algorithm}{\setlength\hsize{15cm}}
-%\floatevery{algorithmic}{\setlength\hsize{15cm}}
-%TODO: meer hspace geven aan algorithms, alles hierboven werkte niet
-\begin{algorithm}
-\caption{Structure EMS-PC}
-\begin{algorithmic}
-\Function{Structure\_PC}{$data$}
-\State connect all pairs of point in undirected graph $G$.
-\State $n \gets 0$
-\While{there are connected pairs $X$, $Y$ such that $n < |\operatorname{Adjecencies}(X)\setminus\{Y\}| $}
-  \ForAll{such connected pairs $X$, $Y$}
-    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
-      \State test whether $X$ and $Y$ are d-separated given $S$
-      \Comment{(Fisher-Z-test on $data$)}
-      \If{$X$ and $Y$ are d-separated given $S$}
-        \State remove the edge $X - Y$ from G
-        \State record $S$ as a separating set of $X$ and $Y$
-      \EndIf
-    \EndFor
-  \EndFor
-  \State $n \gets n+1$
-\EndWhile
-\Return $G$ and the separating sets
-\EndFunction
-\end{algorithmic}
-\label{code:emsstructure}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{Directional EMS-PC}
-\begin{algorithmic}
-\Function{Directional\_PC}{$G$, $sepsets$}
-  \State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
-  \ForAll{unshielded triples $<X, Y, Z>$}
-    \If{$Y$ is not in any separating set of $X$ and $Z \And X$ and $Z$ are dependent given $Y$}
-      \Comment{V-structure-rule}
-      \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
-    \EndIf
-  \EndFor
-  \Repeat
-    \ForAll{ordered, connected pairs $X$, $Y$}
-      \Comment{Rule 1}
-      \If{$X \osline Y$ or $X \sleftline Y$, and there is a node $Z$ such that $Z \srightarrow X$}
-        \State Orient $X \sline Y$ as $X \rightarrow Y$
-      \EndIf
-      \Comment{Rule 2}
-      \If{there is a directed path from $X$ to $Y$}
-        \State Orient $X \sline Y$ as $X \srightarrow Y$
-      \EndIf
-    \EndFor
-  \Until{no more edges are oriented}
-  \State \Return PDAG
-\EndFunction
-\end{algorithmic}
-\label{code:emsdirectional}
-\end{algorithm}
-
-%\par\vfill\break % Omdat pseudocode veel hspace nodig heeft :(
-%\advance\hsize by -8cm % Retur n old margings and page height
-%\advance\hoffset by 4cm % Return old margings and page height
+\section{Methods}
+%= Methods = (our contributions) [Tom: I'm not completely satisfied with this structure. It's too split and diverse to put everything under 'methods']
+% Improvements to PC
+%Multiple separating sets
+%Partial unfaithfulness check
+%EMS-PC
 
 \subsection{Experimental validation}
-% Section on how we plan on demonstrating the empirical value of research.
-% Full description from step 1 to final data set
-Our analysis consisted of two parts, corresponding to the structural and directional parts of PC algorithm.
-Firstly we discerned structure by using brain resting-state time series fMRI data with structure PC.
-The generated structure is the same for all variants of PC discussed and proposed, as only the first separating set is used in finding unshielded triples.
-We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging.
-Comparison between structures is done visually with an expert on brain anatomy.
-To increase the stability of found structures within subjects, we apply stability selection for different node permutations by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
-Multi-subject comparison was done by averaging and inspecting standard deviation over subjects.
+%Application of methods on data
+%Comparison with previous research (due to lack of baseline)
 
-The second part of our analysis is about discerning orientation within the found structure.
-We applied the directionality rules of original PC, modified PC and EMS-PC as mentioned above.
-Resulting connectivity patterns are visually compared by an expert for anatomical correctness.
-To compare between methods, we compare the consistency of methods to mark a specific connection as directional over permutations.
-Of specific interest here are non-symmetrical orientations.
-We again applied stability selection for a sufficiently large number of permutations per subject.
-
-\subsection{Data acquisition}
-No explicit data acquisition was performed, instead we performed our analysis on the same data as treated by Hinne et al.
-in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
-
-Paraphrasing: ``six healthy volunteers were scanned for resting-state functional data and diffusion-weighted images using a Siemens Magnetom Trio 3 T system at the Donders Centre for Cognitive Neuroimaging at Radboud University in The Netherlands.
-Resting-state fMRI data were acquired at 3 Tesla by using a multi-echo echo-planar imaging (ME-EPI) sequence (voxel size 3.5 mm isotropic, matrix size 64x64, TR=2000 ms, TEs=6.9, 16.2, 25, 35 and 45 ms, 39 slices, GRAPPA factor 3, 6/8 partial Fourier).
-A total of 1030 volumes were obtained.
-Diffusion weighted images were obtained by using the DWI protocol (voxel size 2.0 isotropic, matrix size 110x110, TR=13.00 ms, TE=101 ms, 70 slices, 256 directions at b=1500 s/mm$^2$).''
-
-We did not perform explicit (pre-)processing on the functional and diffusion imaging data.
-Instead, we used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as (pre-)processed by Hinne et al. \cite{hinne2013}.
+\section{Data acquisition}
+%Subject acquisition and instruction
+%Used equipment and specs
+%Experimental setup
+%Scanner settings
+%Preprocessing
+%- Software
+%- Scripts
+%Resulting data specs
+%[Ramon: tot hoeverre is dit in detail nodig?
+Enerzijds moeten we niet doen alsof het kant en klaar uit dat apparaat komt, maar anderzijds wordt het in Max' artikel beschreven.
+Zijn dingen als "settings" dan niet overkill?]
 
 \section{Results}
-At first, we have applied modified PC on the structural connectivity data of six subjects.
-To achieve stability, we have run the algorithm multiple times, in which the data was permuted randomly in each iteration.
-The avarage of all iterations was used as a measure of certainty, i.e. how consistent the algorithm is for each point of data.
+\subsection{Structure}
+%structure per brain region
+%comparison with Max's results
+%certainty (plot of the distribution of certainties?)
 
-The results from the altorithms are presented in the form of matrices.
-A connection from region $n$ to region $m$ is represented by point $(n,m)$ or $n,m$ in the matrix.
-For strucural data, this matrix is therefore symmetrical.
-In the directional data, point $(n,m)$ denotes a directed edge from $n$ to $m$.
-In such a matrix, indices 1-45 represent the brain regions of the left hemisphere, regions 45-90 represent the right hemisphere, and regions 91-116 represent regions in the cerebellum.
-%TODO left or right hemisphere?
+\subsection{Causal relations}
+%direction
+%direction per subject? To show that there is little consistency
+%certainty (plot)
 
-Modified PC and EMS-PC do not differ in finding the structure.
-The results are presented in figure ~\ref{fig:struct_full_mean} as the avarage over all six subjects, with 300 iterations per subject.
-\begin{figure}[h!]
-  \centering
-  \includegraphics{images/struct_full_mean_gray}
-  \caption{The avarage structure of all subjects}
-  \label{fig:struct_full_mean}
-\end{figure}
-In this matrix, higher values represent strong certainties of the presence of a connection.
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics{images/structure_max}
-  \caption{The avarage structure of all subjects, according to the DWI-measurements of Hinne et al.}
-  \label{fig:struct_mean_max}
-\end{figure}
-
-One notable difference between the results found in Hinne et al.
-are that connections between the the two hemispheres are clearly visible as two diagonal lines.
-Connections within the hemispheres are less clearly visible, as well as connections within the cerebellum.
-
-% Mod-PC
-
-%
-
-%
-
-%
-
-%
+\subsection{Performance}
+%runtime
 
 \section{Discussion}
+%structure is nice (inter-hemisphere connections)
+%direction is less nice, little consistency
+%better methods for direction (Tom's new algorithm)
+%performs well
 
 \section{Conclusion}
-
-\bibliography{references}{}
-\addcontentsline{toc}{section}{References}
-\bibliographystyle{plain}
+%repeat everything in 1 paragraph, or leave away
 

### 2014-01-14 docs/paper/paper final.tex
besproken en uitgewerkte skeleton en de .tex staat klaar

@@
 \end{abstract}
+\section{Introduction}
+\subsection{Subject introduction}
+%Effective connectivity resting state analysis in human brains
+%Resting state
+%Brains are networks; apply graphical models and causal discovery
 
-\section{Background}% Types of brain connectivity [optional] [Ramon: lijkt me wel belangijk, geeft ook het belang aan van richting zoeken, hoeft niet lang]
-\subsection{Brain connectivity}
+\subsection{ Motivation for research}
+%Better understanding of brains
+%Relatively new field - not yet an established analysis method
+%Existing problems with current methods and applications, specifically [Tom: mention all, or just the relevant problems?]
+%- Model selection
+%- Indirect measurements
+%- Modelling causal structure across individuals (varying responses to same stimuli)
+%- Distinct but overlapping ROIs
+%- Varying Haemodynamic response (HRF)
+%Find shortcomings of PC and improve it (EMS-PC)
+
+\subsection{Problem statement}
+%See whether applying causal methods (PC and EMS-PC) can help bring new insights and whether we can find evidence for the existence of causal patterns on a whole brain level.
+
+\subsection{Previous research}
+%[Tom: we did not explicitly have this paragraph, so this is additional effort]
+%[Ramon]: Lijkt me vergelijkbaar met de verwijzingen naar literatuur die we al hebben? Met wat extra referenties naar wat Max zei: temporele methodes zijn onbetrouwbaar, dat hebben wij niet
+%[Ramon]: even een goede balans zoeken tussen dit en "Existing problems" in motivation
+
+\subsection{Research goal}
+%How does the standard causal discovery method PC-Algorithm behave regarding the aforementioned problems when applied to resting-state functional time-series data?
+%- Can we overcome the consistency problems of PC-algorithm applied on resting-state functional data?
+%- Does the PC algorithm applied on resting-state functional data provide consistent, specific and anatomically plausible connectivity?
+%- How does functional connectivity found by the first part of PC relate to other structural / functional methods applied on the same resting-state functional data regarding the aforementioned problems?
+%
+%How does the performance of EMS-PC relate to PC regarding the aforementioned problems when applied to the same resting-state functional time-series data?
+%- When would EMS-PC perform better than PC-algorithm regarding directionality results?
+%- Is EMS-PC more consistent in its results within a single subject than PC-algorithm on resting-state functional data, regarding the multiple separating set addition and the explicit v-structure test?
+%- Can we find unfaithfulness within our model without additional independence tests?
+
+\subsection{Relevance of research}
+%If HRF can be avoided, causal patterns can be established with greater reliability
+%Insight and diagnostics in neuro-degenerative diseases
+%Any found evidence of causal patterns can give an impulse to further research
+%[Ramon: is dit niet al duidelijk uit motivation? Lijkt me dubbelop]
+
+\section{Background}
+\subsection{Types of brain connectivity}% [optional] [Ramon: lijkt me wel belangijk, geeft ook het belang aan van richting zoeken, hoeft niet lang]
 %Functional
@@
 
-\section{Causal inference}
-% Causal discovery theory
+\subsection{Causal discovery theory}
 %Causal Markov Condition
@@
 
-\section{PC algorithm}
-%Algorithm, assumptions and explanation [Ramon: standaard PC een naam geven en consistent gebruiken, bijv. classical PC?]
+\subsection{PC algorithm}
+%Algorithm, assumptions and explanation [Ramon: standaard PC een naam geven en consistent gebruiken, bijv. standard PC?]
+%Pseudocode
 %Performance
 %Stability and faithfulness problems
+%Soundness and non-completeness
 %Conservative PC extension
 
-\section{Methods}
-%= Methods = (our contributions) [Tom: I'm not completely satisfied with this structure. It's too split and diverse to put everything under 'methods']
-% Improvements to PC
-%Multiple separating sets
-%Partial unfaithfulness check
-%EMS-PC
+\section{Methods}% (our contributions) [Tom: I'm not completely satisfied with this structure. It's too split and diverse to put everything under 'methods']
+\subsection{Improvements to PC}
+%Motivation (refer to results-section)
+%Multiple separating sets and why it should be better / not worse (correctness, runtime)
+%Partial unfaithfulness check and why it should be better / not worse (correctness, runtime)
+%EMS-PC extension (no pseudo-code, but additions regarding standard PC)
 
 \subsection{Experimental validation}
-%Application of methods on data
-%Comparison with previous research (due to lack of baseline)
+\paragraph{Experimental pipeline}
+%- What data is needed
+%- Apply structure PC on resting-state functional time-series
+%- Apply directional PC
+%- Parameter optimisation (model selection)
+%
+\paragraph{Standard PC}
+%- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
+%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
+%- Markov class Asymmetry and certain asymmetry within subjects
+%- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
+%- Consistency of directionality across subjects when aggregating results
+%
+\paragraph{EMS-PC}
+%- Measuring whole-brain structural and directional consistency of EMS-PC
+%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
+%- Markov class Asymmetry and certain asymmetry within subjects
+%- Unfaithfulness check with and without additional separating sets
+%
+\paragraph{Comparison}
+%- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
+%- Comparison of ... of results across subjects
+%- Run-time of methods and comparison between run-times
+%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al.
 
-\section{Data acquisition}
+\subsection{Data acquisition}
 %Subject acquisition and instruction
@@
 \section{Results}
-\subsection{Structure}
-%structure per brain region
-%comparison with Max's results
-%certainty (plot of the distribution of certainties?)
-
-\subsection{Causal relations}
-%direction
-%direction per subject? To show that there is little consistency
-%certainty (plot)
-
-\subsection{Performance}
-%runtime
+%Answer all the questions as mentioned in methods, in the same order
+%No aggregation over direction, motivate this
 
 \section{Discussion}
-%structure is nice (inter-hemisphere connections)
-%direction is less nice, little consistency
-%better methods for direction (Tom's new algorithm)
-%performs well
+%Poor consistency between subjects
+%Good consistency of structure within and between subjects
+%inter-hemisphere connection
+%cause of poor consistency: inherent to our method or viable for further research?
+%data aquisition and noise, improvement in data -> improvement in causal inference
+%better methods for direction (Tom's new algorithm), and why that might solve causal inference problems
+%run-time is quick
+%HRF
 
-\section{Conclusion}
-%repeat everything in 1 paragraph, or leave away
+\subsection{Conclusion}
+%repeat everything in 1 paragraph, or leave away, whatever seems most fitting in the end. HRF makes us awesome
 

### 2014-01-14 docs/paper/paper final.tex
stukje over PC

@@
 %Algorithm, assumptions and explanation [Ramon: standaard PC een naam geven en consistent gebruiken, bijv. standard PC?]
-%Pseudocode
-%Performance
-%Stability and faithfulness problems
-%Soundness and non-completeness
-%Conservative PC extension
+\paragraph{Description}
+The PC algorithm is a so-called constraint based method.
+Starting from a fully connected undirected graph, the PC algorithm first forms structure by cutting away edges of conditionally independent variables.
+In the second step, orientation rules are applied based on found separating sets and structure.
+The original PC algorithm assumes causal sufficiency, excluding the existence of common confounders between nodes.
+This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
+Also measuring methods are not yet advanced enough to measure the full detail of the brain, thus skipping over details.
+In the perspective of causal inference, these missing areas can also appear as confounders.
+For the following experiments we have used the causal insufficient version of PC algorithm, modified to handle latent variable as described by Spirtes, Glymour and Scheines \cite[p.165-167]{spirtes2000}.
+From now on, we will refer to this algorithm as ``standard PC''.
+Pseudocode for this method is shown in algorithm \ref{alg:pc_alg}.
+
+\begin{algorithm}
+\caption{Standard PC algorithm}
+\begin{spacing}{1.1}
+\begin{algorithmic}[1]
+\Function{PC}{$data$}
+\State connect all pairs of point in undirected graph $G$.
+\Comment{Structural Part} 
+\State $n \gets 0$
+\While{there are connected pairs ($X$, $Y$) such that $n < |\operatorname{Adjacencies}(X)\setminus\{Y\}| $}
+  \ForAll{such connected pairs ($X$, $Y$)}
+    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
+      \If{$X$ and $Y$ are independent given $S$}
+        \State remove edge $X - Y$ from G
+        \State record $S$ as separating set of ($X, Y$)
+        \State continue with the next pair ($X$, $Y$)
+      \EndIf
+    \EndFor
+  \EndFor
+  \State $n \gets n+1$
+\EndWhile
+\State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
+\Comment{Directional Part}
+\ForAll{unshielded triples $\left <X, Y,Z \right> $ in $G$ }
+  \If{$Y$ is not in the separating set of $X$ and $Z$}
+    \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
+  \EndIf
+\EndFor
+\While{Edges can still be oriented}
+  \If{there is an edge $X \sline Y$, and there is a directed path from $X$ to $Y$}
+    \State orient $X \sline Y$ as $X \rightarrow Y$
+  \EndIf
+  \If{there are edges $X \srightarrow Y$ and $Y \sleftline Z$, the latter is not an edge $Y \sleftarrow Z$, and $X$ and $Z$ are not connected}
+    \State orient $Y \sline Z$ as $Y \rightarrow Z$
+  \EndIf
+\EndWhile
+\State Apply other directionality rules from Spirtes et al. \cite[p.165-167]{spirtes2000} %TODO uitschrijven?
+\State \Return $G$ and the separating sets
+\EndFunction
+\end{algorithmic}
+\end{spacing}
+\label{alg:pc_alg}
+\end{algorithm}
+
+For finding directed edges, the PC-algorithm first finds undirected structure.
+For each two non-adjacent nodes $X$ and $Y$ it also finds a minimal separating set, which is the smallest set acting as a separating set.
+This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.
+
+The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ using a statistical independence test such as the Chi-square or Fisher-Z independence tests.
+If such an independence is found, this subset is a separating set of $X$ and $Y$ and they are not connected.
+If no such an independence is found, they are connected.
+To find the minimal separating set, all subsets of neighbours are traversed in order of size; the first separating set which is found, is the minimal one.
+
+Next, PC algorithm attempts to orient these edges.
+The 'o' mark represents the entire Markov class in that edge of the graph.
+Like Spirtes et al, we use a star-symbol at the end of an edge to denote that any mark may be present: an arrowhead, an "o" or an empty mark.
+When an edge is oriented as an edge with a star, this denotes that this end is not changed.
+
+The first of the directionality rules attempts to orient unshielded triples.
+For a triple $\left < X,Y,Z, \right>$ to be oriented, $X$ and $Z$ should have a separating set $S$ not containing $Y$.
+If so, the triple is marked a V-structure.
+
+After all V-structures have been found, the other rules are applied in no particular order.
+These rules are based on arrows that have already been found.
+
+\paragraph{Run-time performance}
+The PC-algorithm run-time is dependent on the number of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
+However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test d-separation of every pair of nodes.
+For a sparse graph, many edges might be removed for small separating sets, and the standard PC algorithm therefore often runs fast in practice and is able to perform well on hundreds of variables.
+
+\paragraph{Conservative PC extension}
+The V-structure rule in PC algorithm orients unshielded triples based on a single separating set. 
+Simply put, it is assumed the single found separating set is enough to show conditional independence. 
+Ramsey et al. \cite{ramsey2012} introduce a weaker but still sufficient assumption for finding V-structures.
+Through this weaker but still adjacency-faithful assumption, the method is more cautious than PC algorithm in drawing unambiguous conclusions on causal orientations.
+That is why this method is named Conservative PC (CPC).
+The resulting algorithmic changes provide better results and is able to mark incorrectly unshielded triples as unfaithful.
+Sadly, CPC requires a significant amount of additional conditional independence tests, exponential in the largest branching degree of the tested structure.
+Although the weaker assumptions made with conservative PC might be useful for this research, the run-time performance is not good enough to apply it on our dataset.
 

### 2014-01-14 docs/paper/paper final.tex
ems PC

@@
 \ForAll{unshielded triples $\left <X, Y,Z \right> $ in $G$ }
+                                                                                                                                    \Comment{Find V-structures}
   \If{$Y$ is not in the separating set of $X$ and $Z$}
@@
 \While{Edges can still be oriented}
+                                                                                                                                    \Comment{Additional orientations}
   \If{there is an edge $X \sline Y$, and there is a directed path from $X$ to $Y$}
@@
   \EndIf
-  \If{there are edges $X \srightarrow Y$ and $Y \sleftline Z$, the latter is not an edge $Y \sleftarrow Z$, and $X$ and $Z$ are not connected}
+  \If{there are edges $X \srightarrow Y$ and $Y \sleftline Z$, the latter is not an edge $Y \sleftarrow Z$}
+    \If{$X$ and $Z$ are not connected}
       \State orient $Y \sline Z$ as $Y \rightarrow Z$
     \EndIf
+  \EndIf
 \EndWhile
-\State Apply other directionality rules from Spirtes et al. \cite[p.165-167]{spirtes2000} %TODO uitschrijven?
 \State \Return $G$ and the separating sets
@@
 
+\paragraph{Multiple Separating Sets}
+As will be explained in section~\ref{sec:results}, standard PC does not seem to handle unfaithfulness well enough.
+The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC, is potential as less erroneous arrows are found.
+Instead of doing additional tests as described in CPC, one could attempt to determine causal unfaithfulness with less dependency tests.
+An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found, to test all other vertex subsets of the same order as well.
+This finds all separating sets of the lowest order without increasing complexity.
+Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
+By having more than one separating set, this constraint is strengthened is the data is not faithful.
+If it is, the results are not changed.
+This also does not change the run-time complexity.
+
+% Unfaithfulness test
+One could use these additional separating sets to test oriented unshielded triples for unfaithfulness, similar to CPC.
+Apart from serving as a sanity check, this could serve as the basis of a repair algorithm for errors made by PC.
+
+% Explicit test
+A more explicit way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
+If so, the graph might become more faithful to its modeled distribution; If standard PC might orient a V-structure, whereas this condition does not hold, then the concluded V-structure might be erroneous.
+For faithful graphs, this extra condition should not change the result, as it will always hold.
+
+\paragraph{EMS-PC}
+We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
+The two changes described in the previous section have been made in this adaption with respect to standard PC as described in section ~\ref{sec:pseudocode}.
+At line 10, the search for a separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
+All of these separating sets are recorded.
+
+Also, the condition on line 14 to orient $X - Y - Z$ according to the V-structure-rule has been made more strict.
+As multiple separating sets have been found instead of only one, it needs to be checked whether $Y$ is in not in \emph{any} of those sets before the V-structure can be oriented.
+In addition to the original condition, $X$ and $Z$ are tested for d-separation given $\{Y\}$.
+If they are found to be conditionally dependent, the triple is oriented as as $X \srightarrow Y \sleftarrow Z$.
+%As a final safe-guard, all unshielded triples are checked for unfaithfulness and marked as such if applicable using found separating sets. %TODO wat bedoel je hiermee?
+%TODO Willen we nog iets over unfaithful dingen zeggen? Zo ja, dan wel een antwoord op deze vragen...
+% TODO: How do you mark unfaithful triples? can you spot the impact in the output?
+
 \subsection{Experimental validation}
@@
 
-\section{Results}
+\section{Results}\label{sec:results}
 %Answer all the questions as mentioned in methods, in the same order
@@
 
+\bibliography{references}{}
+\addcontentsline{toc}{section}{References}
+\bibliographystyle{plain}
+
 \end{document}

### 2014-01-14 docs/paper/paper final.tex
Did a grammar and style pass on current contents.

@@
 \paragraph{Description}
-The PC algorithm is a so-called constraint based method.
-Starting from a fully connected undirected graph, the PC algorithm first forms structure by cutting away edges of conditionally independent variables.
+The PC algorithm is a so-called constraint based method, where solutions are found by incrementally leaving out or adding parts that do not violate a set of constraints.
+Starting from a fully connected undirected graph, the PC algorithm forms structure by cutting away edges between conditionally independent variables.
 In the second step, orientation rules are applied based on found separating sets and structure.
@@
 This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
-Also measuring methods are not yet advanced enough to measure the full detail of the brain, thus skipping over details.
+Also measuring methods are not yet advanced enough to measure detailed neuronal processing of the brain, thus skipping over details.
 In the perspective of causal inference, these missing areas can also appear as confounders.
 For the following experiments we have used the causal insufficient version of PC algorithm, modified to handle latent variable as described by Spirtes, Glymour and Scheines \cite[p.165-167]{spirtes2000}.
-From now on, we will refer to this algorithm as ``standard PC''.
+From now on, we will refer to this algorithm as \textit{standard PC}.
 Pseudocode for this method is shown in algorithm \ref{alg:pc_alg}.
@@
 \Function{PC}{$data$}
-\State connect all pairs of point in undirected graph $G$.
+\State connect all pairs of point in undirected graph $G$. \label{alg:pc_alg:p1_start}
                                                                                                                                     \Comment{Structural Part} 
@@
         \State record $S$ as separating set of ($X, Y$)
-        \State continue with the next pair ($X$, $Y$)
+        \State continue with the next pair ($X$, $Y$) \label{alg:pc_alg:next_pair}
       \EndIf
@@
   \EndFor
-  \State $n \gets n+1$
+  \State $n \gets n+1$ \label{alg:pc_alg:p1_end}
 \EndWhile
-\State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
+\\
+\State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$ \label{alg:pc_alg:p2_start}
                                                                                                                                     \Comment{Directional Part}
@@
                                                                                                                                     \Comment{Find V-structures}
-  \If{$Y$ is not in the separating set of $X$ and $Z$}
+  \If{$Y$ is not in the separating set of $X$ and $Z$} \label{alg:pc_alg:v_struct}
     \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
@@
 \EndWhile
-\State \Return $G$ and the separating sets
+\State \Return $G$ and the separating sets \label{alg:pc_alg:p2_end}
 \EndFunction
@@
 
-For finding directed edges, the PC-algorithm first finds undirected structure.
-For each two non-adjacent nodes $X$ and $Y$ it also finds a minimal separating set, which is the smallest set acting as a separating set.
-This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.
+%Before finding directed edges, the PC-algorithm first finds undirected structure.
+%For each two non-adjacent nodes $X$ and $Y$ standard PC finds a minimal separating set, which is the smallest set acting as a separating set.
+%This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.
 
+Starting from a fully connected graph, standard PC finds a minimal separating set for every pair of non-adjacent nodes $X$ and $Y$.
+These are lines \ref{alg:pc_alg:p1_start} to \ref{alg:pc_alg:p1_end} in the pseudocode.
 The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ using a statistical independence test such as the Chi-square or Fisher-Z independence tests.
+%TODO: Add citation to these methods.
 If such an independence is found, this subset is a separating set of $X$ and $Y$ and they are not connected.
-If no such an independence is found, they are connected.
-To find the minimal separating set, all subsets of neighbours are traversed in order of size; the first separating set which is found, is the minimal one.
+If no such set can be found, the variables are assumed to be dependent and thus remain connected.
+To find a minimal separating set, all subsets of neighbours are traversed in order of size; the first separating set which is found, is the minimal one.
 
 Next, PC algorithm attempts to orient these edges.
-The 'o' mark represents the entire Markov class in that edge of the graph.
+These are lines \ref{alg:pc_alg:p2_start} to \ref{alg:pc_alg:p2_end} in the pseudocode.
+The 'o' mark represents that it is unknown what orientation the specific edge should have thus representing an entire Markov class of solutions rather than a specific instance.
 Like Spirtes et al, we use a star-symbol at the end of an edge to denote that any mark may be present: an arrowhead, an "o" or an empty mark.
 When an edge is oriented as an edge with a star, this denotes that this end is not changed.
-
 The first of the directionality rules attempts to orient unshielded triples.
@@
 If so, the triple is marked a V-structure.
-
 After all V-structures have been found, the other rules are applied in no particular order.
-These rules are based on arrows that have already been found.
+These rules are based on already found directionality.
 
 \paragraph{Run-time performance}
-The PC-algorithm run-time is dependent on the number of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
+The PC-algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
 However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test d-separation of every pair of nodes.
-For a sparse graph, many edges might be removed for small separating sets, and the standard PC algorithm therefore often runs fast in practice and is able to perform well on hundreds of variables.
+For a sparse graph, many edges may be removed for small separating sets, and this enables the standard PC algorithm to often run fast in practice perform well on hundreds of variables.
+% TODO: add citation to example publication.
+Peter Lucas maybe or examples in the sprites book?
+
+\paragraph{Shortcomings}
+Although PC is proven correct, the version provided above as standard PC is not.
+It is also not complete.
+There is additional the number of possible graphs in the returned Markov class increasing exponentially with the amount of variables in the provided data, thus not always providing ample information.
+Also, standard PC is dependent on the order in which these variables are presented, as this is the same order in which edges between pairs of these variables are considered for removal\cite[p.88]{spirtes2000}.
+
+In practice, the lack of full correctness and completeness are of little concern as it involves only very specific and little occurring connectivity patterns\cite[p.127-130]{spirtes2000}.
+% TODO: Paragraph seems incomplete.
 
@@
 \paragraph{Multiple Separating Sets}
-As will be explained in section~\ref{sec:results}, standard PC does not seem to handle unfaithfulness well enough.
-The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC, is potential as less erroneous arrows are found.
+As will be explained in the results section, standard PC is a bit hasty in assigning v-structures, resulting in many unnecessary and perhaps unfaithful directionality.
+The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC is potential as less erroneous arrows are found.
 Instead of doing additional tests as described in CPC, one could attempt to determine causal unfaithfulness with less dependency tests.
-An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found, to test all other vertex subsets of the same order as well.
+An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found to test all other vertex subsets of the same order as well.
 This finds all separating sets of the lowest order without increasing complexity.
@@
 \paragraph{EMS-PC}
-We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
-The two changes described in the previous section have been made in this adaption with respect to standard PC as described in section ~\ref{sec:pseudocode}.
-At line 10, the search for a separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
+We have combined the above ideas - multiple separating sets, unfaithfulness check and explicit orientation check - into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
+These three changes have been made in this adaption with respect to standard PC as described in algorithm \ref{alg:pc_alg}.
+At line \ref{alg:pc_alg:next_pair}, the search for separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
 All of these separating sets are recorded.
 
-Also, the condition on line 14 to orient $X - Y - Z$ according to the V-structure-rule has been made more strict.
+Also, the condition on line \ref{alg:pc_alg:v_struct} to orient $X - Y - Z$ according to the V-structure-rule has been made more strict.
 As multiple separating sets have been found instead of only one, it needs to be checked whether $Y$ is in not in \emph{any} of those sets before the V-structure can be oriented.
-In addition to the original condition, $X$ and $Z$ are tested for d-separation given $\{Y\}$.
+In addition to the original condition, $X$ and $Z$ are tested for \textit{d}-separation given $\{Y\}$.
 If they are found to be conditionally dependent, the triple is oriented as as $X \srightarrow Y \sleftarrow Z$.
@@
 %- Parameter optimisation (model selection)
-%
+
 \paragraph{Standard PC}
@@
 %- Consistency of directionality across subjects when aggregating results
-%
+
 \paragraph{EMS-PC}
@@
 %- Markov class Asymmetry and certain asymmetry within subjects
+%- Amount of separating sets found
 %- Unfaithfulness check with and without additional separating sets
-%
+
 \paragraph{Comparison}

### 2014-01-14 docs/paper/paper final.tex
Added section on unfaithfulness check.

@@
 As multiple separating sets have been found instead of only one, it needs to be checked whether $Y$ is in not in \emph{any} of those sets before the V-structure can be oriented.
-In addition to the original condition, $X$ and $Z$ are tested for \textit{d}-separation given $\{Y\}$.
-If they are found to be conditionally dependent, the triple is oriented as as $X \srightarrow Y \sleftarrow Z$.
-%As a final safe-guard, all unshielded triples are checked for unfaithfulness and marked as such if applicable using found separating sets. %TODO wat bedoel je hiermee?
-%TODO Willen we nog iets over unfaithful dingen zeggen? Zo ja, dan wel een antwoord op deze vragen...
-% TODO: How do you mark unfaithful triples? can you spot the impact in the output?
+In addition to this original condition, $X$ and $Z$ are \textit{explicitly} tested for \textit{d}-separation given $\{Y\}$.
+If they are found to be conditionally \textit{dependent}, the triple is oriented as as $X \srightarrow Y \sleftarrow Z$.
+Otherwise it is not oriented.
+
+Given the found separating sets, it is possible to test whether these sets provide a consistent answer to the question whether the unshielded triple $\left<X,Y,Z\right>$ is \textit{d}-separated by $Y$.
+It should be the case that $Y$ occurs in all found separating sets.
+If this is not the case, the triple is marked as unfaithful.
+This can be used as a diagnostic to judge initial reliability of results without almost any additional computational cost.
+The marked triples could be coloured differently in the resulting 
+The downside of this additional check is results are merely used as a diagnostic and thus the unshielded triple is oriented in a v-structure nevertheless.
+It is also not a complete check as this would require as many tests as in CPC and thus no positive conclusions can be drawn from it.
+Also if an unfaithful triple is detected, the impact on the final output is unclear as errors might have cascaded into further orientation steps.
 

### 2014-01-15 docs/paper/paper final.tex
Wrote first part of discussion.

@@
 \section{Discussion}
-%Poor consistency between subjects
 %Good consistency of structure within and between subjects
-%inter-hemisphere connection
 %cause of poor consistency: inherent to our method or viable for further research?
@@
 %HRF
+
+%Comparison with DWI inter-hemisphere connection
+Standard PC and particularly the first, structural part, definitely has similarity with respect to existing methods based on DWI.
+This is not completely surprising as standard PC yields a functional network and DWI based methods anatomical structural networks, two network types which are tightly related.
+A less apparent observation is the clear presence of correlation between homologous area's in the functional network, which lacks in the structural network.
+An explanation to this difference might be that these correlations respond to the long white-matter tracts connecting the two hemispheres through the corpus-callosum.
+It is known that DWI has difficulty in measuring such long tracts, while the presence of correlation between two connected ends cannot be argued against.
+As such, statistical baed causal discovery methods contribute additional information to finding structural connectivity.
+
+% Comparison standard PC and EMS-PC
+Both standard PC and EMS-PC find many bi-directional arrows, with some one-directional arrows.
+
+%Poor consistency of directionality between subjects
+There is some significant and consistent directionality within each of the subjects.
+However the variance of this directionality between subjects is so high, there is no significant consistent one-directionality between them.
+We were unable to find existing literature on connectivity methods regarding significant intra-subject one-directionality.
+
+%Lack of symmetry
+There is little to none symmetry in directionality between the two hemispheres within and across subjects.
+One would at least expect some symmetry to be found.
+This demonstrates that the gap between analytical methods for effective connectivity and anatomical correctness is still to wide.
+
+% Further research
+Further research may be done to investigating whether any of these arrows have a physiological meaning.
+This can be done by repeating these experiment with more subjects to get a better view on the significance of the directional edges.
+
+A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
+However, exclusion of cycles in brain networks has not been proven and also seems unlikely.
+We believe an interesting direction of further research would thus be to apply cyclical causal discovery.
+% TODO: Add citation to Joris Mooij.
+
+% TODO: What do you mean with 'it can also do this'? What does 'this' refer to?
+% Computational advantage
+%It can also do this quite efficiently, as a few hundred iterations can be run in a matter of hours per subject.
+%We have not performed a statistical analysis on the number of iterations needed to stabilise the results.
+%It may be possible that a significant result can already be found with much less iterations.
+
+% TODO: Suggereren task based analysis ipv resting state voor beter causaal beeld. 
+% TODO: Maak duidelijk welke aannames gedaan worden in welke PC variant en welke aannames verzwakt zouden kunnen worden.
 

### 2014-01-15 docs/paper/paper final.tex
Wrote some more paragraphs for the discussion.

@@
 %Good consistency of structure within and between subjects
-%cause of poor consistency: inherent to our method or viable for further research?
-%data aquisition and noise, improvement in data -> improvement in causal inference
-%better methods for direction (Tom's new algorithm), and why that might solve causal inference problems
-%run-time is quick
+
 %HRF
+
+% TODO: Maak duidelijk welke aannames gedaan worden in welke PC variant en welke aannames verzwakt zouden kunnen worden.
+
+% Quick run-time
+For PC as well as standard PC we were able to run hundreds of iterations on 116 variables in a matter of hours per subject.
+Iterations can also be run in parallel.
+Further speed increments can be achieved by coding the algorithm in a low overhead language.
 
@@
 
-% Comparison standard PC and EMS-PC
+% Comparison Markov class standard PC and EMS-PC
 Both standard PC and EMS-PC find many bi-directional arrows, with some one-directional arrows.
+% TODO: find out how large these groups are.
 
@@
 
-A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
+A causal network or Bayesian network cannot contain cycles by definition.
 However, exclusion of cycles in brain networks has not been proven and also seems unlikely.
-We believe an interesting direction of further research would thus be to apply cyclical causal discovery.
-% TODO: Add citation to Joris Mooij.
+We believe an interesting direction of further research would be to apply cyclical causal methods, such as the method proposed by Mooij and Heskes\cite{Mooij20013}.
 
-% TODO: What do you mean with 'it can also do this'? What does 'this' refer to?
-% Computational advantage
-%It can also do this quite efficiently, as a few hundred iterations can be run in a matter of hours per subject.
-%We have not performed a statistical analysis on the number of iterations needed to stabilise the results.
-%It may be possible that a significant result can already be found with much less iterations.
+We believe application of a Bayesian approach will provide a better insight.
+Standard PC and EMS-PC lack a robust measure of uncertainty.
+There is no baseline or correctness measure to compare results or data with
+A Bayesian approach such as proposed by Claassen and Heskes will provide more slack and insight in results \cite{claassen2012}.
+This in turn might leverage development of better and faster data driven methods.
 
 % TODO: Suggereren task based analysis ipv resting state voor beter causaal beeld.
-% TODO: Maak duidelijk welke aannames gedaan worden in welke PC variant en welke aannames verzwakt zouden kunnen worden.
+
+%cause of poor consistency: inherent to our method or viable for further research?
 

### 2014-01-16 docs/paper/paper final.tex
stuk over validation toegevoegd. Nog niet tevreden over data acquisition, dat is lastig te verwoorden

--- 2014-01-15 docs/paper/paper final.tex
@@
 \subsection{Research goal}
-%How does the standard causal discovery method PC-Algorithm behave regarding the aforementioned problems when applied to resting-state functional time-series data?
-%- Can we overcome the consistency problems of PC-algorithm applied on resting-state functional data?
+%How does the standard causal discovery method PC Algorithm behave regarding the aforementioned problems when applied to resting-state functional time-series data?
+%- Can we overcome the consistency problems of PC algorithm applied on resting-state functional data?
 %- Does the PC algorithm applied on resting-state functional data provide consistent, specific and anatomically plausible connectivity?
@@
 %How does the performance of EMS-PC relate to PC regarding the aforementioned problems when applied to the same resting-state functional time-series data?
-%- When would EMS-PC perform better than PC-algorithm regarding directionality results?
-%- Is EMS-PC more consistent in its results within a single subject than PC-algorithm on resting-state functional data, regarding the multiple separating set addition and the explicit v-structure test?
+%- When would EMS-PC perform better than PC algorithm regarding directionality results?
+%- Is EMS-PC more consistent in its results within a single subject than PC algorithm on resting-state functional data, regarding the multiple separating set addition and the explicit v-structure test?
 %- Can we find unfaithfulness within our model without additional independence tests?
@@
 
-%Before finding directed edges, the PC-algorithm first finds undirected structure.
+%Before finding directed edges, the PC algorithm first finds undirected structure.
 %For each two non-adjacent nodes $X$ and $Y$ standard PC finds a minimal separating set, which is the smallest set acting as a separating set.
@@
 \paragraph{Run-time performance}
-The PC-algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
+The PC algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
 However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test d-separation of every pair of nodes.
@@
 \paragraph{Multiple Separating Sets}
-As will be explained in the results section, standard PC is a bit hasty in assigning v-structures, resulting in many unnecessary and perhaps unfaithful directionality.
+As will be explained in the results section, standard PC is a bit hasty in assigning v-structures, resulting in much unnecessary and perhaps unfaithful directionality.
 The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC is potential as less erroneous arrows are found.
@@
 Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
-By having more than one separating set, this constraint is strengthened is the data is not faithful.
-If it is, the results are not changed.
+By having more than one separating set, this constraint is strengthened if the data is not faithful.
+If it is, then the results are not changed.
 This also does not change the run-time complexity.
@@
 
+%stability and iterations TODO: right place for this? I think it is, as it can be seen as a simple extension to naive PC
+\paragraph{Stability}
+The graph resulting from the PC algorithm depends on the order of the nodes\cite[p.120-122]{spirtes2000}.
+It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
+As such, it seems useful to calculate a chance of a connection appearing.
+When running the experiments, we have applied the algorithms multiple times using random permutations of the nodes to obtain stability selection as described by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
+These results have been avaraged to conclude the certainty of the found edges.
+
 \subsection{Experimental validation}
-\paragraph{Experimental pipeline}
+\paragraph{Experimental pipeline}{sec:pipeline}
 %- What data is needed
@@
 %- Apply directional PC
+Both standard PC and EMS-pc have been used to find brain connectivity.
+The data which is used is functional time-series data, and it is the same as the data used in Hinne et al.\cite{hinne2013}, as described in Section~\ref{sec:data}.
+This data is obtained from six subject.
+From this data, covariance between regions can be determined, so that regions can be tested for \emph{d}-separation with the Fisher's conditional independence test.
+This is required to use the PC algorithm.
+First, the strucural part of the PC algorithm will be applied on the data.
+This will result in a graph describing undirected connections between regions.
+There is no difference between standard PC and EMS-PC in this result.
+After that, the second part of the PC algorithm is run to orient edges in this graph.
+This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.
 %- Parameter optimisation (model selection)
+The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject.
 
-\paragraph{Standard PC}
+\paragraph{Concluding connectivity from PC}
 %- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
+Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
+This is done both per subject, and as an avarage over all subjects.
+We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
+
+In the second stage of PC, a directed graph is found.
+Again, a certainty of each individual arrowhead may be found by applying stability selection.
 %- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
+Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
+Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imlpy a causal relation.
+Edges of two arrowhead imply a common confounder.
 %- Markov class Asymmetry and certain asymmetry within subjects
 %- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
+The results will again be analyses per subject, and as an avarage over all subjects; Causal relations appearing in specific subjects will, or in general.
 %- Consistency of directionality across subjects when aggregating results
+The functioning of the brain will not completely differ between different subjects and therefore some consistency across subjects is expected.
+Consistency between the two hemispheres is also expected.
 
-\paragraph{EMS-PC}
+%\paragraph{EMS-PC}   Er werd veel herhaald t.o.v.
+standard PC, dus ik heb ze samen genomen.
+amount of sepsets en faithfulness is niet dubbel, maar hebben we deze resultaten opgeslagen en kunnen we er iets nuttigs over zeggen?
 %- Measuring whole-brain structural and directional consistency of EMS-PC
@@
 
-\paragraph{Comparison}
+\paragraph{Comparison between standard PC and EMS-PC}
 %- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
+As EMS-PC is more constrained in concluding directionality, it may find less erroneous arrows than standard PC.
+To verify this, the number and consistency of one-directional arrows is compared.
 %- Comparison of ... of results across subjects
+Again, this is both done per-subject and on the average across subjects.
 %- Run-time of methods and comparison between run-times
-%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al.
+The run-time of both methods will be compared.
+%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC
 
-\subsection{Data acquisition}
+\subsection{Data acquisition}\label{sec:data}
 %Subject acquisition and instruction
@@
 %Resulting data specs
-%[Ramon: tot hoeverre is dit in detail nodig?
-Enerzijds moeten we niet doen alsof het kant en klaar uit dat apparaat komt, maar anderzijds wordt het in Max' artikel beschreven.
-Zijn dingen als "settings" dan niet overkill?]
+%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen?
+"This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
+We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
+This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
+This functional data is divided in 119 nodes and 1029 time-series per subject.
+Additionaly, diffusion-weighted images of these subjects have been obtained.
+Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
+We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.
+
+No additional pre-processing has been done in our research in order to apply the PC algorithm on this data.
 

### 2014-01-16 docs/paper/paper final.tex
foutje gefixt
Spelling and style corrections. 'Avarage' is not a word.

@@
 Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
-By having more than one separating set, this constraint is strengthened if the data is not faithful.
-If it is, then the results are not changed.
-This also does not change the run-time complexity.
+By having more than one separating set, this constraint is strengthened in case of noisy data.
 
@@
 
-%stability and iterations TODO: right place for this? I think it is, as it can be seen as a simple extension to naive PC
-\paragraph{Stability}
-The graph resulting from the PC algorithm depends on the order of the nodes\cite[p.120-122]{spirtes2000}.
-It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
-As such, it seems useful to calculate a chance of a connection appearing.
-When running the experiments, we have applied the algorithms multiple times using random permutations of the nodes to obtain stability selection as described by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
-These results have been avaraged to conclude the certainty of the found edges.
-
 \subsection{Experimental validation}
@@
 This data is obtained from six subject.
-From this data, covariance between regions can be determined, so that regions can be tested for \emph{d}-separation with the Fisher's conditional independence test.
+From this data, covariance between regions can be determined, so that regions can be tested for \textit{d}-separation with the Fisher's conditional independence test.
 This is required to use the PC algorithm.
@@
 This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.
+
+\paragraph{Stability}
+Standard PC depends on the ordering of variables in the provided data\cite[p.120-122]{spirtes2000}.
+It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
+To correct for this in results, we have randomly permuted our data for a large amount of iterations and counted the frequencies of occurring edges.
+This is in accordance with the stability selection method for graphs as described by Meinshausen and B{\"u}hlmann\cite{meinshausen2010}.
+
 %- Parameter optimisation (model selection)
 The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject.
+% TODO: Onvoldoende.
 
@@
 Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
-This is done both per subject, and as an avarage over all subjects.
+This is done both per subject, and as an average over all subjects.
 We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
@@
 Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
-Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imlpy a causal relation.
+Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imply a causal relation.
 Edges of two arrowhead imply a common confounder.
@@
 This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
-This functional data is divided in 119 nodes and 1029 time-series per subject.
-Additionaly, diffusion-weighted images of these subjects have been obtained.
+This functional data is divided in 116 nodes and 1029 time-series per subject.
+Additionally, diffusion-weighted images of these subjects have been obtained.
 Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.

### 2014-01-16 docs/paper/paper final.tex
Wrote part on 'what data is needed'

@@
 \subsection{PC algorithm}
-%Algorithm, assumptions and explanation [Ramon: standaard PC een naam geven en consistent gebruiken, bijv. standard PC?]
+%Algorithm, assumptions and explanation
 \paragraph{Description}
@@
 
-%Before finding directed edges, the PC algorithm first finds undirected structure.
-%For each two non-adjacent nodes $X$ and $Y$ standard PC finds a minimal separating set, which is the smallest set acting as a separating set.
-%This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.
-
 Starting from a fully connected graph, standard PC finds a minimal separating set for every pair of non-adjacent nodes $X$ and $Y$.
@@
 \paragraph{Multiple Separating Sets}
-As will be explained in the results section, standard PC is a bit hasty in assigning v-structures, resulting in much unnecessary and perhaps unfaithful directionality.
-The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC is potential as less erroneous arrows are found.
+As can be seen in the result in the next section, standard PC is a bit hasty in assigning v-structures, resulting in much unnecessary and perhaps unfaithful directionality.
+The idea of CPC, to rely less on graph faithfulness than standard PC is potential as less erroneous arrows are assigned in theory.
 Instead of doing additional tests as described in CPC, one could attempt to determine causal unfaithfulness with less dependency tests.
@@
 By having more than one separating set, this constraint is strengthened in case of noisy data.
-
 % Unfaithfulness test
@@
 % Explicit test
-A more explicit way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
+A more \textit{explicit} way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
 If so, the graph might become more faithful to its modeled distribution; If standard PC might orient a V-structure, whereas this condition does not hold, then the concluded V-structure might be erroneous.
@@
 %- What data is needed
+PC algorithm requires observational data where each instance is a vector of single observations per variable.
+We can model functional resting-state time-series as coming from some stable underlying mechanism that is largely time-independent.
+Resting-state data can be intuitively thought of being more time independent than task-influenced data due to the emergent character of activity pattens that occur when in resting-state.
+Nevertheless it is most likely not a fully correct modelling assumption, though enables straightforward application of standard and EMS-PC to functional data without the problem of HRF.
+
 %- Apply structure PC on resting-state functional time-series
@@
 
+% TODO: Move to appendix?
 \subsection{Data acquisition}\label{sec:data}

### 2014-01-16 docs/paper/paper final.tex
Moved data acquisition to appendix A

@@
 %- Apply directional PC
-Both standard PC and EMS-pc have been used to find brain connectivity.
-The data which is used is functional time-series data, and it is the same as the data used in Hinne et al.\cite{hinne2013}, as described in Section~\ref{sec:data}.
-This data is obtained from six subject.
-From this data, covariance between regions can be determined, so that regions can be tested for \textit{d}-separation with the Fisher's conditional independence test.
-This is required to use the PC algorithm.
-First, the strucural part of the PC algorithm will be applied on the data.
+We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al\cite{hinne2013}.
+1030 volumes were obtained per subject.
+After preprocessing the fMRI data was parcellated according to the Automated Anatomical Labeling (AAL) atlas in 116 regions of interest.
+The exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing is provided in appendix \ref{sec:data}.
+We use Fisher's z-test for conditional independence to match the continuous nature of the data.
+
+For all experiments we first apply part 1 of standard PC and EMS-PC on the data.
 This will result in a graph describing undirected connections between regions.
-There is no difference between standard PC and EMS-PC in this result.
-After that, the second part of the PC algorithm is run to orient edges in this graph.
-This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.
+For structure only the first found separating set is relevant, thus EMS-PC and standard PC return identical graph structure.
+This structure is then oriented with part 2 of standard PC and EMS-PC given earlier found separating sets.
+This procedure yields different orientations in theory.
 
@@
 
-%- Parameter optimisation (model selection)
+\paragraph{Parameter optimisation}
 The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject.
@@
 %- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC
-
-% TODO: Move to appendix?
-\subsection{Data acquisition}\label{sec:data}
-%Subject acquisition and instruction
-%Used equipment and specs
-%Experimental setup
-%Scanner settings
-%Preprocessing
-%- Software
-%- Scripts
-%Resulting data specs
-%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen?
-"This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
-We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
-This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
-This functional data is divided in 116 nodes and 1029 time-series per subject.
-Additionally, diffusion-weighted images of these subjects have been obtained.
-Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
-We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.
-
-No additional pre-processing has been done in our research in order to apply the PC algorithm on this data.
 
@@
 
+\appendix
+\section{Data acquisition}\label{sec:data}
+%Subject acquisition and instruction
+%Used equipment and specs
+%Experimental setup
+%Scanner settings
+%Preprocessing
+%- Software
+%- Scripts
+%Resulting data specs
+%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen?
+"This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
+We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
+This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
+This functional data is divided in 116 nodes and 1029 time-series per subject.
+Additionally, diffusion-weighted images of these subjects have been obtained.
+Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
+We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.
+
+No additional pre-processing has been done in our research in order to apply the PC algorithm on this data.
+
 \end{document}

### 2014-01-16 docs/paper/paper final.tex
stukje over resultaten toegevoegd, enkele plaatjes moeten nog toegevoegd worden

--- 2014-01-16 docs/paper/paper final.tex
@@
 In practice, the lack of full correctness and completeness are of little concern as it involves only very specific and little occurring connectivity patterns\cite[p.127-130]{spirtes2000}.
-% TODO: Paragraph seems incomplete.
+% TODO: Paragraph seems incomplete and some sentences are weird.
 
@@
 Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
-By having more than one separating set, this constraint is strengthened in case of noisy data.
+By having more than one separating set, this constraint is strengthened if the data is not faithful.
+If it is, then the results are not changed.
+This also does not change the run-time complexity.
 
@@
 
+\paragraph{Stability}
+The graph resulting from the PC algorithm depends on the order of the nodes\cite[p.120-122]{spirtes2000}.
+It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
+As such, it seems useful to calculate a chance of a connection appearing.
+When running the experiments, we have applied the algorithms multiple times using random permutations of the nodes to obtain stability selection as described by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
+These results have been avaraged to conclude the certainty of the found edges.
+
 \subsection{Experimental validation}
@@
 This data is obtained from six subject.
-From this data, covariance between regions can be determined, so that regions can be tested for \textit{d}-separation with the Fisher's conditional independence test.
+From this data, covariance between regions can be determined, so that regions can be tested for \emph{d}-separation with the Fisher's conditional independence test.
 This is required to use the PC algorithm.
@@
 This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.
-
-\paragraph{Stability}
-Standard PC depends on the ordering of variables in the provided data\cite[p.120-122]{spirtes2000}.
-It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
-To correct for this in results, we have randomly permuted our data for a large amount of iterations and counted the frequencies of occurring edges.
-This is in accordance with the stability selection method for graphs as described by Meinshausen and B{\"u}hlmann\cite{meinshausen2010}.
-
 %- Parameter optimisation (model selection)
-The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject.
-% TODO: Onvoldoende.
+The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject for stability selection.
 
@@
 Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
-This is done both per subject, and as an average over all subjects.
+This is done both per subject, and as an avarage over all subjects.
 We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
@@
 Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
-Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imply a causal relation.
+Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imlpy a causal relation.
 Edges of two arrowhead imply a common confounder.
@@
 The functioning of the brain will not completely differ between different subjects and therefore some consistency across subjects is expected.
-Consistency between the two hemispheres is also expected.
+Consistency between the two hemispheres is also expected, as they are anatomically symmetrical.
 
@@
 This functional data is divided in 116 nodes and 1029 time-series per subject.
-Additionally, diffusion-weighted images of these subjects have been obtained.
+Additionaly, diffusion-weighted images of these subjects have been obtained.
 Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
@@
 \section{Results}\label{sec:results}
+
+The results from the agorithms are presented in the form of matrices.
+A connection from region $n$ to region $m$ is represented by point $(n,m)$ in the matrix.
+Higher values represent strong certainties of the presence of a connection.
+In such a matrix, indices 1-45 represent the brain regions of the left hemisphere, regions 45-90 represent the right hemisphere, and regions 91-116 represent regions in the cerebellum.
+
+\subsection{The structural graph}
+%- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
+%Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
+%This is done both per subject, and as an avarage over all subjects.
+%We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
+Modified PC and EMS-PC do not differ in finding the structure.
+As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
+The results are presented per subject in Figure~\ref{} and as an avarage across all subjects iin Figure ~\ref{fig:struct_avg}. %TODO figure fixen per subjects
+For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=0.5\textwidth]{images/struct_full}
+  \includegraphics[width=0.5\textwidth]{images/structure_max}
+  \caption{The avarage structure of all subjects.
+The structure on the left is found with the PC-algorithm, and the structure on the right is found with DWI-measurements by Hinne et al.
+Note that these matrices are subdivided into blocks corresponding with the subdivision of indices as described in Section~\ref{sec:results}}.
+The two diagonal lines represent connections between the two hemispheres.
+\label{fig:struct_avg}
+\end{figure}
+
+\subsection{The directed graph}
+This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
+The results for standard PC are shown per subject in Figure~\ref{} and in as an avarage across subjects in Figure~\ref{fig:pdag_avg_mod}.
+Additionally, a histogram showing the frequencies of the values in the avarage matrix is presented in Figure~\ref{}.
+Although many arrowheads are found with probabilities above 0.9, the graph looks strongly symmetric.
+In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.
+
+The results for EMS-PC are presented in a similar way.
+The results per subjects are shown in Figure~\ref{} and the avarage across subjects in Figure~\ref{fig:pdag_avg_ems}.
+A histogram of the avarage is presented in Figure~\ref{}.
+Again, the image looks strongly symmetric.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/pdag_avg_colored_mod}
+  \caption{The PDAG found with standard PC, in which a high value denotes a high certainty of an arrowhead. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg_mod}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/PDAG_avg_colored_expl}
+  \caption{The PDAG found with EMS-PC, in which a high value denotes a high certainty of an arrowhead. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg_ems}
+\end{figure}
+
+Another representation of this data has been used to emphasise the asymmetric parts of Figures ~\ref{fig:pdag_avg_mod} and ~\ref{fig:pdag_avg_ems}.
+To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
+These resulting antisymmetric matrices are presented in Figures ~\ref{fig:pdag_avg_antisymmetric_mod} and ~\ref{fig:pdag_avg_antisymmetric_ems}.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/arrowheads_avg_mod}
+  \caption{The asymmetric arrowheads of the PDAG found with standard PC, in which a high value denotes a high certainty of an one-directional arrow. The average of all six subjects is taken.}
+  \label{fig:pdag_avg_antisymmetric_mod}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics{images/PDAG_avg_antisymmetric_expl}
+  \caption{The assymetric parts of the PDAG found with EMS-PC, in which a high value denotes a high certainty of an one-directional arrow. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg_antisymmetric_ems}
+\end{figure}
+
+%Again, a certainty of each individual arrowhead may be found by applying stability selection.
+%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
+%Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
+%Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imlpy a causal relation.
+%Edges of two arrowhead imply a common confounder.
+%- Markov class Asymmetry and certain asymmetry within subjects
+%- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
+%The results will again be analyses per subject, and as an avarage over all subjects; Causal relations appearing in specific subjects will, or in general.
+%- Consistency of directionality across subjects when aggregating results
+%The functioning of the brain will not completely differ between different subjects and therefore some consistency across subjects is expected.
+%Consistency between the two hemispheres is also expected, as they are anatomically symmetrical.
+
+%\paragraph{EMS-PC}   Er werd veel herhaald t.o.v.
+standard PC, dus ik heb ze samen genomen.
+amount of sepsets en faithfulness is niet dubbel, maar hebben we deze resultaten opgeslagen en kunnen we er iets nuttigs over zeggen?
+%- Measuring whole-brain structural and directional consistency of EMS-PC
+%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
+%- Markov class Asymmetry and certain asymmetry within subjects
+%- Amount of separating sets found
+%- Unfaithfulness check with and without additional separating sets
+
+%\paragraph{Comparison between standard PC and EMS-PC}
+%- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
+%As EMS-PC is more constrained in concluding directionality, it may find less erroneous arrows than standard PC.
+%To verify this, the number and consistency of one-directional arrows is compared.
+%- Comparison of ... of results across subjects
+%Again, this is both done per-subject and on the average across subjects.
+%- Run-time of methods and comparison between run-times
+%The run-time of both methods will be compared.
+%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC
+
 %Answer all the questions as mentioned in methods, in the same order
@@
 Standard PC and particularly the first, structural part, definitely has similarity with respect to existing methods based on DWI.
-This is not completely surprising as standard PC yields a functional network and DWI based methods anatomical structural networks, two network types which are tightly related.
-A less apparent observation is the clear presence of correlation between homologous area's in the functional network, which lacks in the structural network.
+This is not completely surprising as standard PC yields a functional network and DWI based methods yield anatomical structural networks, two network types which are tightly related.
+A less apparent observation is the clear presence of correlation between homologous area's in the functional network, which lacks in the structural network as found with DWI.
 An explanation to this difference might be that these correlations respond to the long white-matter tracts connecting the two hemispheres through the corpus-callosum.

### 2014-01-16 docs/paper/paper final.tex
Elaborated on exprimental validation.

--- 2014-01-16 docs/paper/paper final.tex
@@
 \subsection{Experimental validation}
-\paragraph{Experimental pipeline}{sec:pipeline}
-%- What data is needed
+\paragraph{Experimental pipeline}\label{sec:pipeline}
+%- What type of data is needed
 PC algorithm requires observational data where each instance is a vector of single observations per variable.
@@
 
-%- Apply structure PC on resting-state functional time-series
-%- Apply directional PC
+% Application to data
 We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al\cite{hinne2013}.
@@
 
+% Implementation
+We have implemented standard PC and EMS-PC in Matlab using the Bayesian Network Toolbox for Matlab.
+% TODO: Finish this.
+
+% Pipeline
 For all experiments we first apply part 1 of standard PC and EMS-PC on the data.
-This will result in a graph describing undirected connections between regions.
+This results in a graph describing undirected functional connections between regions.
 For structure only the first found separating set is relevant, thus EMS-PC and standard PC return identical graph structure.
@@
 
-\paragraph{Stability}
+% Stability selection
 Standard PC depends on the ordering of variables in the provided data\cite[p.120-122]{spirtes2000}.
@@
 
-\paragraph{Parameter optimisation}
-The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject.
-% TODO: Onvoldoende.
+% Parameter selection
+Due to the lack of a baseline to compare results with, dependency thresholds were chosen by hand.
+All experiments listed here were run with an alpha value of 0.01 for the independence test.
+The number of random permutations was set on 1000 per subject, which was found to be enough to determine statistical significance for 99\% of the found edges with a binomial statistical test.
 
-\paragraph{Concluding connectivity from PC}
 %- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
-Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
-This is done both per subject, and as an average over all subjects.
-We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
+We obtained structure and causal graphs for each subject with the previous procedure.
+For each graph, we express the level of consistency of an edge in the frequency of occurrence of that type of edge over all permutations.
+The frequency ratio per is then used to create the whole brain image for structure and directionality.
+We compared the generated functional graph with the Bayesian inference approach on finding structural connectivity from Diffusion Weighted Imaging (DWI) proposed by Hinne et al.\cite{hinne2013} on the same data using edge consistency.
 
-In the second stage of PC, a directed graph is found.
-Again, a certainty of each individual arrowhead may be found by applying stability selection.
-%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
-Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
-Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imply a causal relation.
-Edges of two arrowhead imply a common confounder.
-%- Markov class Asymmetry and certain asymmetry within subjects
-%- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
-The results will again be analyses per subject, and as an avarage over all subjects; Causal relations appearing in specific subjects will, or in general.
-%- Consistency of directionality across subjects when aggregating results
-The functioning of the brain will not completely differ between different subjects and therefore some consistency across subjects is expected.
-Consistency between the two hemispheres is also expected.
+%Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects and Markov class Asymmetry
+Since standard PC and EMS-PC return Markov equivalence classes, we record the size of the solution space per subject.
+Different permutations of nodes may yield different directionality for the same edge, including bi-directional relationships.
+Of specific interest are consistent non-symmetrical orientations, as one-directional arrows imply a high-confidence causal relation.
+Bi-directional edges imply an unseen common confounder.
+We obtain uni-directional edges by subtracting the symmetrical bi-directional components from the causal graph.
 
-%\paragraph{EMS-PC}   Er werd veel herhaald t.o.v.
-standard PC, dus ik heb ze samen genomen.
-amount of sepsets en faithfulness is niet dubbel, maar hebben we deze resultaten opgeslagen en kunnen we er iets nuttigs over zeggen?
-%- Measuring whole-brain structural and directional consistency of EMS-PC
-%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
-%- Markov class Asymmetry and certain asymmetry within subjects
-%- Amount of separating sets found
-%- Unfaithfulness check with and without additional separating sets
+% Multi-subject experiments
+To aggregate results over multiple subjects, we sum frequencies per found edge type for structure and directionality to obtain consistency of edges and directionality.
+We expected brain functioning to not completely differ between subjects.
 
-\paragraph{Comparison between standard PC and EMS-PC}
-%- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
-As EMS-PC is more constrained in concluding directionality, it may find less erroneous arrows than standard PC.
-To verify this, the number and consistency of one-directional arrows is compared.
-%- Comparison of ... of results across subjects
-Again, this is both done per-subject and on the average across subjects.
+% EMS-PC
+For results on EMS-PC, we record the size of the separating sets found per edge as well as the amount of unfaithful marked v-structures.
+
+% Comparison standard PC and EMS-PC
+We compare standard PC and EMS-PC on their causal graph.
+As EMS-PC is more constrained in concluding directionality, it may find less erroneous edges than standard PC.
+As we cannot verify this directly, we compare the total count, consistency and standard-deviation of one-directional arrows.
+We also compare run-time per iteration between both
+
 %- Run-time of methods and comparison between run-times
 The run-time of both methods will be compared.
+
 %- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC

### 2014-01-16 docs/paper/paper final.tex

--- 2014-01-16 docs/paper/paper final.tex
@@
 Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
-By having more than one separating set, this constraint is strengthened if the data is not faithful.
-If it is, then the results are not changed.
-This also does not change the run-time complexity.
-
+By having more than one separating set, this constraint is strengthened in case of noisy data.
 % Unfaithfulness test
@@
 %- What data is needed
+PC algorithm requires observational data where each instance is a vector of single observations per variable.
+We can model functional resting-state time-series as coming from some stable underlying mechanism that is largely time-independent.
+Resting-state data can be intuitively thought of being more time independent than task-influenced data due to the emergent character of activity pattens that occur when in resting-state.
+Nevertheless it is most likely not a fully correct modelling assumption, though enables straightforward application of standard and EMS-PC to functional data without the problem of HRF.
+
 %- Apply structure PC on resting-state functional time-series
 %- Apply directional PC
-Both standard PC and EMS-pc have been used to find brain connectivity.
-The data which is used is functional time-series data, and it is the same as the data used in Hinne et al.\cite{hinne2013}, as described in Section~\ref{sec:data}.
-This data is obtained from six subject.
-From this data, covariance between regions can be determined, so that regions can be tested for \emph{d}-separation with the Fisher's conditional independence test.
-This is required to use the PC algorithm.
-First, the strucural part of the PC algorithm will be applied on the data.
+We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al\cite{hinne2013}.
+1030 volumes were obtained per subject.
+After preprocessing the fMRI data was parcellated according to the Automated Anatomical Labeling (AAL) atlas in 116 regions of interest.
+The exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing is provided in appendix \ref{sec:data}.
+We use Fisher's z-test for conditional independence to match the continuous nature of the data.
+
+For all experiments we first apply part 1 of standard PC and EMS-PC on the data.
 This will result in a graph describing undirected connections between regions.
-There is no difference between standard PC and EMS-PC in this result.
-After that, the second part of the PC algorithm is run to orient edges in this graph.
-This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.
-%- Parameter optimisation (model selection)
+For structure only the first found separating set is relevant, thus EMS-PC and standard PC return identical graph structure.
+This structure is then oriented with part 2 of standard PC and EMS-PC given earlier found separating sets.
+This procedure yields different orientations in theory.
+
+\paragraph{Stability}
+Standard PC depends on the ordering of variables in the provided data\cite[p.120-122]{spirtes2000}.
+It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
+To correct for this in results, we have randomly permuted our data for a large amount of iterations and counted the frequencies of occurring edges.
+This is in accordance with the stability selection method for graphs as described by Meinshausen and B{\"u}hlmann\cite{meinshausen2010}.
+
+\paragraph{Parameter optimisation}
 The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject for stability selection.
+% TODO: Onvoldoende.
 
@@
 %- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC
-
-\subsection{Data acquisition}\label{sec:data}
-%Subject acquisition and instruction
-%Used equipment and specs
-%Experimental setup
-%Scanner settings
-%Preprocessing
-%- Software
-%- Scripts
-%Resulting data specs
-%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen?
-"This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
-We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
-This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
-This functional data is divided in 116 nodes and 1029 time-series per subject.
-Additionaly, diffusion-weighted images of these subjects have been obtained.
-Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
-We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.
-
-No additional pre-processing has been done in our research in order to apply the PC algorithm on this data.
 

### 2014-01-16 docs/paper/paper final.tex

@@
 
-\paragraph{Parameter optimisation}
-The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject for stability selection.
-% TODO: Onvoldoende.
+% Parameter selection
+Due to the lack of a baseline to compare results with, dependency thresholds were chosen by hand.
+All experiments listed here were run with an alpha value of 0.01 for the independence test.
+The number of random permutations was set on 1000 per subject, which was found to be enough to determine statistical significance for 99\% of the found edges with a binomial statistical test.
 
-\paragraph{Concluding connectivity from PC}
 %- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
-Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
-This is done both per subject, and as an average over all subjects.
-We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
 
-In the second stage of PC, a directed graph is found.
-Again, a certainty of each individual arrowhead may be found by applying stability selection.
-%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
-Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
-Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imply a causal relation.
-Edges of two arrowhead imply a common confounder.
-%- Markov class Asymmetry and certain asymmetry within subjects
-%- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
-The results will again be analyses per subject, and as an avarage over all subjects; Causal relations appearing in specific subjects will, or in general.
-%- Consistency of directionality across subjects when aggregating results
-The functioning of the brain will not completely differ between different subjects and therefore some consistency across subjects is expected.
-Consistency between the two hemispheres is also expected, as they are anatomically symmetrical.
+We obtained structure and causal graphs for each subject with the previous procedure.
+For each graph, we express the level of consistency of an edge in the frequency of occurrence of that type of edge over all permutations.
+The frequency ratio per is then used to create the whole brain image for structure and directionality.
+We compared the generated functional graph with the Bayesian inference approach on finding structural connectivity from Diffusion Weighted Imaging (DWI) proposed by Hinne et al.\cite{hinne2013} on the same data using edge consistency.
 
-%\paragraph{EMS-PC}   Er werd veel herhaald t.o.v.
-standard PC, dus ik heb ze samen genomen.
-amount of sepsets en faithfulness is niet dubbel, maar hebben we deze resultaten opgeslagen en kunnen we er iets nuttigs over zeggen?
-%- Measuring whole-brain structural and directional consistency of EMS-PC
-%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
-%- Markov class Asymmetry and certain asymmetry within subjects
-%- Amount of separating sets found
-%- Unfaithfulness check with and without additional separating sets
+%Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects and Markov class Asymmetry
+Since standard PC and EMS-PC return Markov equivalence classes, we record the size of the solution space per subject.
+Different permutations of nodes may yield different directionality for the same edge, including bi-directional relationships.
+Of specific interest are consistent non-symmetrical orientations, as one-directional arrows imply a high-confidence causal relation.
+Bi-directional edges imply an unseen common confounder.
+We obtain uni-directional edges by subtracting the symmetrical bi-directional components from the causal graph.
 
-\paragraph{Comparison between standard PC and EMS-PC}
-%- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
-As EMS-PC is more constrained in concluding directionality, it may find less erroneous arrows than standard PC.
-To verify this, the number and consistency of one-directional arrows is compared.
-%- Comparison of ... of results across subjects
-Again, this is both done per-subject and on the average across subjects.
+% Multi-subject experiments
+To aggregate results over multiple subjects, we sum frequencies per found edge type for structure and directionality to obtain consistency of edges and directionality.
+We expected brain functioning to not completely differ between subjects.
+
+% EMS-PC
+For results on EMS-PC, we record the size of the separating sets found per edge as well as the amount of unfaithful marked v-structures.
+
+% Comparison standard PC and EMS-PC
+We compare standard PC and EMS-PC on their causal graph.
+As EMS-PC is more constrained in concluding directionality, it may find less erroneous edges than standard PC.
+As we cannot verify this directly, we compare the total count, consistency and standard-deviation of one-directional arrows.
 %- Run-time of methods and comparison between run-times
-The run-time of both methods will be compared.
+We also compare run-time per iteration between both.
+
 %- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC

### 2014-01-16 docs/paper/paper final.tex
Cleanup in 'Method'.

@@
 
-\paragraph{Stability}
-The graph resulting from the PC algorithm depends on the order of the nodes\cite[p.120-122]{spirtes2000}.
-It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
-As such, it seems useful to calculate a chance of a connection appearing.
-When running the experiments, we have applied the algorithms multiple times using random permutations of the nodes to obtain stability selection as described by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
-These results have been avaraged to conclude the certainty of the found edges.
-
 \subsection{Experimental validation}
@@
 % Implementation
-We have implemented standard PC and EMS-PC in Matlab using the Bayesian Network Toolbox for Matlab.
-% TODO: Finish this.
+We have implemented standard PC and EMS-PC in Matlab using statistical tests and ideas from the Bayesian Network Toolbox for Matlab.\footnote{Bayesian Network Toolbox for Matlab: https://github.com/bayesnet/bnt}
+This toolbox also includes an implementation for standard PC, however to gain better insight in the workings and to have greater flexibility in application, we have implemented standard PC separately.
+The Matlab scripts used for the results are made freely available online.\footnote{Matlab implementation of EMS-PC: https://github.com/tomderuijter/EMS-PC}
 
@@
 
-%- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
-
+% Structural and Directional Consistency
+% DWI Comparison
 We obtained structure and causal graphs for each subject with the previous procedure.
@@
 
-%Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects and Markov class Asymmetry
+%Markov equivalence class size and asymmetry
 Since standard PC and EMS-PC return Markov equivalence classes, we record the size of the solution space per subject.
@@
 
-%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC
-
 \section{Results}\label{sec:results}
-
-The results from the agorithms are presented in the form of matrices.
+The results from the algorithms are presented in the form of matrices.
 A connection from region $n$ to region $m$ is represented by point $(n,m)$ in the matrix.

### 2014-01-16 docs/paper/paper final.tex
Made references consistent.

@@
 From now on, we will refer to this algorithm as \textit{standard PC}.
-Pseudocode for this method is shown in algorithm \ref{alg:pc_alg}.
+Pseudocode for this method is shown in algorithm~\ref{alg:pc_alg}.
 
@@
 Starting from a fully connected graph, standard PC finds a minimal separating set for every pair of non-adjacent nodes $X$ and $Y$.
-These are lines \ref{alg:pc_alg:p1_start} to \ref{alg:pc_alg:p1_end} in the pseudocode.
+These are lines~\ref{alg:pc_alg:p1_start} to \ref{alg:pc_alg:p1_end} in the pseudocode.
 The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ using a statistical independence test such as the Chi-square or Fisher-Z independence tests.
@@
 Next, PC algorithm attempts to orient these edges.
-These are lines \ref{alg:pc_alg:p2_start} to \ref{alg:pc_alg:p2_end} in the pseudocode.
+These are lines~\ref{alg:pc_alg:p2_start} to \ref{alg:pc_alg:p2_end} in the pseudocode.
 The 'o' mark represents that it is unknown what orientation the specific edge should have thus representing an entire Markov class of solutions rather than a specific instance.
@@
 We have combined the above ideas - multiple separating sets, unfaithfulness check and explicit orientation check - into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
-These three changes have been made in this adaption with respect to standard PC as described in algorithm \ref{alg:pc_alg}.
-At line \ref{alg:pc_alg:next_pair}, the search for separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
+These three changes have been made in this adaption with respect to standard PC as described in algorithm~\ref{alg:pc_alg}.
+At line~\ref{alg:pc_alg:next_pair}, the search for separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
 All of these separating sets are recorded.
 
-Also, the condition on line \ref{alg:pc_alg:v_struct} to orient $X - Y - Z$ according to the V-structure-rule has been made more strict.
+Also, the condition on line~\ref{alg:pc_alg:v_struct} to orient $X - Y - Z$ according to the V-structure-rule has been made more strict.
 As multiple separating sets have been found instead of only one, it needs to be checked whether $Y$ is in not in \emph{any} of those sets before the V-structure can be oriented.
@@
 After preprocessing the fMRI data was parcellated according to the Automated Anatomical Labeling (AAL) atlas in 116 regions of interest.
-The exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing is provided in appendix \ref{sec:data}.
+The exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing is provided in appendix~\ref{sec:data}.
 We use Fisher's z-test for conditional independence to match the continuous nature of the data.
@@
 \section{Results}\label{sec:results}
-The results from the algorithms are presented in the form of matrices.
+Due to the size of the resulting graphs, we represent them in the form of connectivity matrices for convenience.
 A connection from region $n$ to region $m$ is represented by point $(n,m)$ in the matrix.
-Higher values represent strong certainties of the presence of a connection.
-In such a matrix, indices 1-45 represent the brain regions of the left hemisphere, regions 45-90 represent the right hemisphere, and regions 91-116 represent regions in the cerebellum.
+Higher values represent higher frequencies of a connection over all permutations.
+In all matrices, indices 1-45 represent the brain regions of the left hemisphere, indices 45-90 represent the right hemisphere, and indices 91-116 represent regions in the cerebellum.
+% For clarification, these are also plotted separately.
 
-\subsection{The structural graph}
+\subsection{Structural graph}
 %- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
@@
 As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
-The results are presented per subject in Figure~\ref{} and as an avarage across all subjects iin Figure ~\ref{fig:struct_avg}. %TODO figure fixen per subjects
+The results are presented per subject in figure~\ref{} and as an avarage across all subjects in figure~\ref{fig:struct_avg}. %TODO: figure fixen per subjects
 For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.
@@
 The structure on the left is found with the PC-algorithm, and the structure on the right is found with DWI-measurements by Hinne et al.
-Note that these matrices are subdivided into blocks corresponding with the subdivision of indices as described in Section~\ref{sec:results}}.
-The two diagonal lines represent connections between the two hemispheres.
+Note that these matrices are subdivided into blocks corresponding with the subdivision of indices as described in section~\ref{sec:results}.
+The two diagonal lines represent connections between the two hemispheres.}
 \label{fig:struct_avg}
@@
 
-\subsection{The directed graph}
+\subsection{Causal graph}
 This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
-The results for standard PC are shown per subject in Figure~\ref{} and in as an avarage across subjects in Figure~\ref{fig:pdag_avg_mod}.
-Additionally, a histogram showing the frequencies of the values in the avarage matrix is presented in Figure~\ref{}.
+The results for standard PC are shown per subject in Figure~\ref{} and in as an avarage across subjects in figure~\ref{fig:pdag_avg_mod}.
+Additionally, a histogram showing the frequencies of the values in the avarage matrix is presented in figure~\ref{}.
 Although many arrowheads are found with probabilities above 0.9, the graph looks strongly symmetric.
@@
 The results for EMS-PC are presented in a similar way.
-The results per subjects are shown in Figure~\ref{} and the avarage across subjects in Figure~\ref{fig:pdag_avg_ems}.
-A histogram of the avarage is presented in Figure~\ref{}.
+The results per subjects are shown in figure~\ref{} and the avarage across subjects in figure~\ref{fig:pdag_avg_ems}.
+A histogram of the average is presented in figure~\ref{}.
 Again, the image looks strongly symmetric.

### 2014-01-16 docs/paper/paper final.tex
Small details in structure.

@@
 
-In practice, the lack of full correctness and completeness are of little concern as it involves only very specific and little occurring connectivity patterns\cite[p.127-130]{spirtes2000}.
+In practice, the lack of full correctness and completeness are of little concern as these involve only very specific and little occurring connectivity patterns\cite[p.127-130]{spirtes2000}.
 % TODO: Paragraph seems incomplete and some sentences are weird.
@@
 
-\section{Methods}% (our contributions) [Tom: I'm not completely satisfied with this structure. It's too split and diverse to put everything under 'methods']
+\section{Methods}% (our contributions)
+Standard PC provides a good starting point for an exploratory point of view in causal discovery.
+Already with simple and small adjustments to standard PC with little increment in computational cost, standard PC can be made more reliable.
+In the rest of this section we describe our additions to this method, as well as an experimental setup
+
 \subsection{Improvements to PC}
-%Motivation (refer to results-section)
 %Multiple separating sets and why it should be better / not worse (correctness, runtime)
-%Partial unfaithfulness check and why it should be better / not worse (correctness, runtime)
-%EMS-PC extension (no pseudo-code, but additions regarding standard PC)
-
 \paragraph{Multiple Separating Sets}
@@
 % Explicit test
+\paragraph{Explicit V-structure test}
 A more \textit{explicit} way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
@@
 
+%EMS-PC extension (no pseudo-code, but additions regarding standard PC)
 \paragraph{EMS-PC}

### 2014-01-16 docs/paper/paper final.tex
Added section on causal discovery methods under background.

@@
 %Latent variables
+An approach for finding analysing effective connectivity may be found in the domain of causal discovery.
+A causal discovery method uses data of observed events to determine which of these events are correlated and additionally attempts to determine which events cause which.
+We can define a causal relation as follows.
+
+% Causal relation
+\paragraph{Causal relation}
+For a specific system $S$, a variable $Q$ causes variable $R$ in $S$ if there is at least one value $q$ and at least one value $r$ that during a specific event in which $Q=q$ causes a \textit{different} event such that $R = r$ in $S$\cite[p.21]{spirtes2000}.
+
+In the context of brain connectivity, variables correspond to regions of interest, events correspond to moments in time after a specific stimulus or in resting state, and values correspond to the measured BOLD response related to this event in that specific region.
+By modelling these variables and the causal relationships between them in a Directed Acyclic Graph (DAG), we can make use of the existing theory of Bayesian Networks to infer underlying causal models.
+
+Under the causal interpretation of a DAG $G$, a causal relation is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
+In terms of the graph, $X$ is the causal parent of $Y$, and $Y$ is the child of $X$. 
+When there is a directed path $X \rightarrow Z_1 \rightarrow Z_2 \dots \rightarrow Y$, $X$ is said to be an ancestor of $Y$ and $Y$ is a descendant of $X$.
+An ancestor is a (possibly indirect) cause of its descendant.
+The graph $G$ is also referred to as a causal graph.
+
+Under the probabilistic interpretation a DAG $G$, also called a Bayesian network, represents a joint probability distribution over a set of variables $V$.
+Conditional dependencies are shown by arrows between vertices.
+Proportionality is shown by an undirected edge between vertices.
+A provable property of these networks is the \textit{Markov Property}: every variable in $G$ is independent of its non-descendants conditioned on its parents.
+This property enables us to link the field of Bayesian networks, for which useful mathematical properties exist, to the causal interpretation.
+
+\paragraph{Causal Markov Condition}
+A set of variables represented by a causal DAG $G$ is probabilistically independent of its non-descendants (non-effects), conditioned on its parents (direct causes).
+
+This allows reasoning of causality based on probability.
+The Causal Markov Condition also has a converse principle allowing us to reason with graph structure and directionality.
+
+\paragraph{Causal Faithfulness Condition}
+A set of variables represented by a causal DAG $G$ does not have any conditional independencies unless entailed by the Causal Markov Condition.
+
+This second condition implies the graph correctly portrays the underlying probability distribution.
+In a Bayesian network, edges between vertices denote a conditional dependency.
+By instantiating certain variables, other variables can become independent or dependent.
+With the above two properties on underlying causal mechanisms and their graph, causal mechanisms and causal dependencies can graphically be determined through the notion of \textit{d}-separation.
+
+\paragraph{\textit{d}-separation}
+Two nodes $X$ and $Y$ are \textit{d}-separated given a node set $S$ (with $X, Y \notin S$) if there is no path between $X,Y$ or a path from $X$ to $Y$ through $S$ for which at least one of the following holds:
+\begin{itemize}
+\item The path contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $S$.
+\item The path contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $S$.
+\item The path contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are \textbf{not} in $S$.
+\end{itemize}
+If $X$ and $Y$ are \textit{d}-separated, $S$ is called a separating set of $X$ and $Y$.
+Every two independent variables have a separating set; if they are unconditionally independent, this is the empty set.
+
+In the context of brain connectivity, it is doubtable that measurements and underlying mechanisms are not cyclical or are fully true to the Causal Markov condition or to Causal faithfulness.
+However as there is no true baseline to compare with, this is difficult to determine analytically.
+If it is the case that graphs generated with functional data are not true to the above properties in specific parts of the graph, these parts are said to be \textit{unfaithful} to the underlying causal mechanism.
+
+Call a triple of variables $\left < X,Y,Z \right >$ an \textit{unshielded triple} if $X$ and $Z$ are adjacent to Y, but not to each other.
+
+\paragraph{V-structure}
+An unshielded triple can be oriented $X \rightarrow Y \leftarrow Z$ if and only all sets that \textit{d}-separate $X$ from $Z$ do not contain $Y$.
+This structure is named \textit{V-structure}.
+
+It is a certainty we cannot observe the full causal process in the human brain.
+We may additionally assume that observed variables are a subset of those in the underlying causal mechanism.
+Models having this Open World property are called \textit{Causal Insufficient}.
+When sampling from the observed distribution, we would find a relation $X \leftrightarrow Z$.
+Since we assumed Causal Markov Conditionality, no feedback loops exist and all bi-directional arrows found must indicate existence of a latent confounder $Y$ where $X \leftarrow Y \rightarrow Z$.
 
 \subsection{PC algorithm}
-%Algorithm, assumptions and explanation
-\paragraph{Description}
+% Algorithm description
 The PC algorithm is a so-called constraint based method, where solutions are found by incrementally leaving out or adding parts that do not violate a set of constraints.

### 2014-01-16 docs/paper/paper final.tex
Added part of brain connectivity types to introduction.

@@
 %[Ramon]: even een goede balans zoeken tussen dit en "Existing problems" in motivation
+% Previous work
+Previous work on effective connectivity include the aforementioned Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modelling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
+All of these methods fail to cope with the problems mentioned in the introduction.
+Most of these methods can only handle tens of variables in reasonable time, which becomes problematic in realistic situations with up to tens of thousands of variables.
 
@@
 \section{Background}
-\subsection{Types of brain connectivity}% [optional] [Ramon: lijkt me wel belangijk, geeft ook het belang aan van richting zoeken, hoeft niet lang]
-%Functional
-%Structural
-%Effective
+\subsection{Types of brain connectivity}
+Work on brain connectivity theory discerns three types of brain connectivity.
+
+\paragraph{Functional connectivity}
+Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
+Functional connectivity can be deduced through correlation of Functional Magnetic Resonance Imaging (fMRI) measurements of oxygen saturation  in blood - the BOLD response - while performing a certain cognitive task, or in so called resting-state.
+
+\paragraph{Resting-state}
+Resting-state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity; neural activity does not stop when not doing anything in particular.
+Resting-state provides a typical whole brain behaviour with a low frequency ( < .1 Hz) that is reliable, robust, unique to a subject and task-independent.
+Measuring resting-state thus provides an advantage when analysing brain connectivity on a global, whole brain level.
+Resting-state measurements have the disadvantage that it is difficult to check whether a subject truly is in a resting-state; it is often quite difficult not to think of anything in particular.
+Resting-state analysis has been shown to work well in practice \cite{Lowe2000, doria2010, Bullmore2009}.
+
+\paragraph{Structural connectivity}
+Structural connectivity is defined as the mapping of white-matter tracts in the brain between different regions \cite{friston1994}.
+It is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
+This intuitive relation can also be demonstrated empirically \cite{vandenheuvel2009}.
+MRI scanners can be used to measure white-matter tracts directly.
+As water is most concentrated in these regions, these tracts can be detected by measuring the diffusion of water in the brain.
+This process is named Diffusion Weighted MRI (DWI).
+For all practical purposes structural connectivity is time-independent - it changes over years, and still marginally during adulthood.
+
+\paragraph{Effective connectivity}
+In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of connectivity relations.
+It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
+Functional time series data needs to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which.
+Granger causality is a method relying on this temporal principle.
+In brain data, this principle might not necessarily hold due to the fluctuations in haemodynamic response, the lag between neuronal activity and the measured BOLD response.
+Nevertheless, in theory it is possible to infer effective connectivity from structural connectivity and functional data \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.
+An established relation of effective connectivity to actual anatomical brain relations has yet to be established.
 
 \subsection{Causal discovery theory}
-%Causal Markov Condition
-%Causal faithfulness
-%d-separation
-%V-structure
-%Causal sufficiency
-%Latent variables
 An approach for finding analysing effective connectivity may be found in the domain of causal discovery.

### 2014-01-16 docs/paper/paper final.tex
Made a start to the introduction.

@@
 \end{abstract}
+
 \section{Introduction}
@@
 %Brains are networks; apply graphical models and causal discovery
+
+% Petty introductory talk
+The brain and particularly the human brain have been studied for hundreds of years.
+Today, brain research is still one of the thriving fields of research.
+A lot has changed since the time of Phrenology, the pseudoscience attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
+Advanced and non-intrusive measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview with widespread clinical and research purposes.
 
@@
 
-\subsection{Conclusion}
+% \section{Conclusion}
 %repeat everything in 1 paragraph, or leave away, whatever seems most fitting in the end. HRF makes us awesome

### 2014-01-16 docs/paper/paper final.tex
plaatjes toegevoegd; de structuur per subject is nog wel de verkeerde zwart-wit, dus die fix ik morgen even (staat al een scriptje voor klaar)

@@
 In all matrices, indices 1-45 represent the brain regions of the left hemisphere, indices 45-90 represent the right hemisphere, and indices 91-116 represent regions in the cerebellum.
+As such, the matrices are divided into blocks: the top left corner represents the connections within the left hemisphere, the bottom left corner represents the connections between the left hemisphere and the cerebellum, etc.
 % For clarification, these are also plotted separately.
 
-\subsection{Structural graph}
+\subsection{Structural graphs}
 %- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
@@
 As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
-The results are presented per subject in figure~\ref{} and as an avarage across all subjects in figure~\ref{fig:struct_avg}. %TODO: figure fixen per subjects
+The results are presented per subject in figure~\ref{fig:struct_subjects} and as an avarage across all subjects in figure~\ref{fig:struct_avg}.
 For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{images/struct_full}
-  \includegraphics[width=0.5\textwidth]{images/structure_max}
+  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop} % TODO remove colorbar, why is it even there
+  \includegraphics[height=0.28\textwidth]{images/new/struct_subj2-crop}
+  \includegraphics[height=0.28\textwidth]{images/new/struct_subj3_colorbar-crop} % one colorbar at the top right of the figure
+  \includegraphics[height=0.28\textwidth]{images/new/struct_subj4-crop}
+  \includegraphics[height=0.28\textwidth]{images/new/struct_subj5-crop}
+  \includegraphics[height=0.28\textwidth]{images/new/struct_subj6-crop}
+  \caption{The structure of all six subjects as found with the PC algorithm. The two diagonal lines represent connections between the two hemispheres.}
+\label{fig:struct_subjects}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=0.49\textwidth]{images/struct_full}
+  \includegraphics[width=0.49\textwidth]{images/structure_max}
   \caption{The avarage structure of all subjects.
 The structure on the left is found with the PC-algorithm, and the structure on the right is found with DWI-measurements by Hinne et al.
-Note that these matrices are subdivided into blocks corresponding with the subdivision of indices as described in section~\ref{sec:results}.
 The two diagonal lines represent connections between the two hemispheres.}
@@
 
-\subsection{Causal graph}
+\subsection{Directed graphs}
 This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
-The results for standard PC are shown per subject in Figure~\ref{} and in as an avarage across subjects in figure~\ref{fig:pdag_avg_mod}.
-Additionally, a histogram showing the frequencies of the values in the avarage matrix is presented in figure~\ref{}.
-Although many arrowheads are found with probabilities above 0.9, the graph looks strongly symmetric.
+As there were no notable differences between different subjects, only the avarage PDAG across all subjects is presented.
+The results for standard PC and EMS-PC are shown in figure~\ref{fig:pdag_avg}.
+Although many arrowheads are found with probabilities above 0.9 for both variants of the algorithm, the graph looks strongly symmetric.
 In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.
-
-The results for EMS-PC are presented in a similar way.
-The results per subjects are shown in figure~\ref{} and the avarage across subjects in figure~\ref{fig:pdag_avg_ems}.
-A histogram of the average is presented in figure~\ref{}.
-Again, the image looks strongly symmetric.
 
@@
   \centering
-  \includegraphics{images/pdag_avg_colored_mod}
-  \caption{The PDAG found with standard PC, in which a high value denotes a high certainty of an arrowhead. The avarage of all six subjects is taken.}
-  \label{fig:pdag_avg_mod}
+  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_mod-crop}
+  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_expl-crop}
+  \caption{The PDAG found standard PC at the left, and EMS-PC at the right. A high value denotes a high certainty of an arrowhead. The avarage of all six subjects is taken.}
+  \label{fig:pdag_avg}
+\end{figure}
+
+Another representation of this data has been used to emphasise the asymmetric parts of the directed graphs.
+To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
+These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:assym_subjects}.
+The avarage across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:assym_avg}.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[height=0.25\textwidth]{images/new/assym_subj1_expl-crop}
+  \includegraphics[height=0.25\textwidth]{images/new/assym_subj2_expl-crop}
+  \includegraphics[height=0.25\textwidth]{images/new/assym_subj3_expl-crop}
+  \includegraphics[height=0.25\textwidth]{images/new/assym_subj4_expl-crop}
+  \includegraphics[height=0.25\textwidth]{images/new/assym_subj5_expl-crop}
+  \includegraphics[height=0.25\textwidth]{images/new/assym_subj6_expl-crop}
+  \caption{The assymetric parts of the PDAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
+\label{fig:assym_subjects}
 \end{figure}
@@
   \centering
-  \includegraphics{images/PDAG_avg_colored_expl}
-  \caption{The PDAG found with EMS-PC, in which a high value denotes a high certainty of an arrowhead. The avarage of all six subjects is taken.}
-  \label{fig:pdag_avg_ems}
+  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_mod-crop}
+  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_expl-crop}
+  \caption{The assymetric parts of the PDAG found with standard PC on the left, and EMS-PC on the right.
+A high value denotes a high certainty of an one-directional arrow.
+The avarage of all six subjects is taken.}
+  \label{fig:assym_avg}
 \end{figure}
 
-Another representation of this data has been used to emphasise the asymmetric parts of Figures ~\ref{fig:pdag_avg_mod} and ~\ref{fig:pdag_avg_ems}.
-To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
-These resulting antisymmetric matrices are presented in Figures ~\ref{fig:pdag_avg_antisymmetric_mod} and ~\ref{fig:pdag_avg_antisymmetric_ems}.
-
+As can be seen in figures~\ref{assym_subjects} and~\ref{assym_avg}, the certainties of one-directional arrows are very low compared to the certainties of the bi-directional arrows.
+Also, the certainties of one-directional arrows are somewhat higher in EMS-PC than in standard PC.
+To illustrate this, the distribution of these values are presented in figure~\ref{fig:distributions}
 \begin{figure}[h!]
   \centering
-  \includegraphics{images/arrowheads_avg_mod}
-  \caption{The asymmetric arrowheads of the PDAG found with standard PC, in which a high value denotes a high certainty of an one-directional arrow. The average of all six subjects is taken.}
-  \label{fig:pdag_avg_antisymmetric_mod}
+  \includegraphics[width=0.49\textwidth]{images/new/dist_all_mod_blackstruct_reddir_blueantisym-crop}
+  \includegraphics[width=0.49\textwidth]{images/new/dist_all_expl_blackstruct_reddir_blueantisym-crop}
+  \caption{The distribution of values, for standard PC at the left, and EMS-PC at the right.
+The horizontal axis shows the relative number of occurences in the graph.
+The black line represent the distribution of results of the structural part of PC, and is equal for both variants.
+The red and the blue line represent the distribution of values of arrowheads and one-directional arrows, respectively.}
+  \label{fig:assym_avg}
 \end{figure}
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics{images/PDAG_avg_antisymmetric_expl}
-  \caption{The assymetric parts of the PDAG found with EMS-PC, in which a high value denotes a high certainty of an one-directional arrow. The avarage of all six subjects is taken.}
-  \label{fig:pdag_avg_antisymmetric_ems}
-\end{figure}
-
-%Again, a certainty of each individual arrowhead may be found by applying stability selection.
-%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
-%Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
-%Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imlpy a causal relation.
-%Edges of two arrowhead imply a common confounder.
-%- Markov class Asymmetry and certain asymmetry within subjects
-%- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
-%The results will again be analyses per subject, and as an avarage over all subjects; Causal relations appearing in specific subjects will, or in general.
-%- Consistency of directionality across subjects when aggregating results
-%The functioning of the brain will not completely differ between different subjects and therefore some consistency across subjects is expected.
-%Consistency between the two hemispheres is also expected, as they are anatomically symmetrical.
-
-%\paragraph{EMS-PC}   Er werd veel herhaald t.o.v.
-standard PC, dus ik heb ze samen genomen.
-amount of sepsets en faithfulness is niet dubbel, maar hebben we deze resultaten opgeslagen en kunnen we er iets nuttigs over zeggen?
-%- Measuring whole-brain structural and directional consistency of EMS-PC
-%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
-%- Markov class Asymmetry and certain asymmetry within subjects
-%- Amount of separating sets found
-%- Unfaithfulness check with and without additional separating sets
-
-%\paragraph{Comparison between standard PC and EMS-PC}
-%- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
-%As EMS-PC is more constrained in concluding directionality, it may find less erroneous arrows than standard PC.
-%To verify this, the number and consistency of one-directional arrows is compared.
-%- Comparison of ... of results across subjects
-%Again, this is both done per-subject and on the average across subjects.
-%- Run-time of methods and comparison between run-times
-%The run-time of both methods will be compared.
-%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC
 

### 2014-01-17 docs/paper/paper final.tex
stukje aan de discussie toegevoegd

@@
 The horizontal axis shows the relative number of occurences in the graph.
+Zero-values make up more than 84\% of the graph, these are not shown in this distribution.
 The black line represent the distribution of results of the structural part of PC, and is equal for both variants.
 The red and the blue line represent the distribution of values of arrowheads and one-directional arrows, respectively.}
-  \label{fig:assym_avg}
+  \label{fig:distributions}
 \end{figure}
@@
 %Good consistency of structure within and between subjects
+Many structural connections found by the PC algorithm are found consistently, both within one subject and also across multiple subjects.
+The connections between the two hemispheres can clearly be seen in all subjects, as well as many connections in the corpus collosum.
+The consistency can also be seen in the distribution of values in figure~\ref{distributions}.
+Non-zero values make up approximately 16 percentage points of the graph, of which 3 percentage points show a certainty larger than 0.5.
+Approximately 1 percentage point even shows a certainty of 1.
 
@@
 Both standard PC and EMS-PC find many bi-directional arrows, with some one-directional arrows.
-% TODO: find out how large these groups are.
+As can be seen in the distribution of values in figure~\ref{distributions}, nearly all found edges are oriented as bi-directional arrows.
+For EMS-PC, this number is somewhat lower but there are still few one-directional arrows, and they do not have certainties higher than 0.15.
 

### 2014-01-17 docs/paper/paper final.tex
veel toegevoegd, met name het stuk introduction-previous work-motivation etc.

@@
 \begin{abstract}
+Many different approaches have been used to infer brain connectivity.
+But few general causal discovery methods have yet been used to find effective connectivity.
+In this article, we describe how the PC algorithm is used on functional brain-data.
+This approach results in connectivity matrices which clearly show some well-known brain connections.
+Some disadvantages of the PC algorithm for this use are discussed, and an attempt is made to decrease these disadvantages by making adjustments to the algorithm.
+Causal relations have only be found with very low certainty. 
 % TODO: first state what you try to do, then how you did it, then what it brought you
@@
 \section{Introduction}
-\subsection{Subject introduction}
 %Effective connectivity resting state analysis in human brains
@@
 %Relatively new field - not yet an established analysis method
-%Existing problems with current methods and applications, specifically [Tom: mention all, or just the relevant problems?]
+
+% Motivation for brain connectomics
+The question how our brains are wired - brain connectomics - drives one of the leading fields of research in neuroscience.
+Our brains are in essence networks which can be analysed similarly to very large graphs.
+Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity \cite{Bullmore2009}.
+An application of connectivity analysis would be to provide markers to identify the disease and its severity, as well as another means of studying these diseases.
+
+% Causal discovery and brain networks
+The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
+Research regarding effective connectivity - causal relations in brain networks - is relatively new and its application still faces many challenges that require further research \cite{ramsey2010}.
+
+%Existing problems with current methods and applications, specifically [Tom: mention all, or just the relevant problems?][Ramon: de relevante nu benoemd]
 %- Model selection
+
+Ramsey et al. lists several problems in inferring brain connectivity.
 %- Indirect measurements
-%- Modelling causal structure across individuals (varying responses to same stimuli)
+One of these problems is that many interesting brain-data is measured indirectly; a measured variable is an effect of the cause we want to know.
+This makes inference hard in practical models: many latent variables may be present, influencing this cause-effect relation.
 %- Distinct but overlapping ROIs
 %- Varying Haemodynamic response (HRF)
+Related is the problem of haemodynamic response fluctuations (HRF).
+In functional MRI-measurements, the delay between brain activity and the indirect measurement of that activity is unknown. %TODO moest MRI nou helemaal uitgeschreven of niet? Ik vind zelf van niet.
+Any form of causal inference based on sequence in time might therefore be incorrect; A cause may be measured after its effects, if it has a larger delay.
+%- Modelling causal structure across individuals (varying responses to same stimuli)
+Also, brain-data is not always consistent, both over time and across different persons. This makes it hard to get significant results.
 %Find shortcomings of PC and improve it (EMS-PC)
 
-\subsection{Problem statement}
-%See whether applying causal methods (PC and EMS-PC) can help bring new insights and whether we can find evidence for the existence of causal patterns on a whole brain level.
+Similarly to how graph theory has been applied in functional and anatomical brain networks, it seems natural to apply methods from causal discovery to deduce effective brain networks.
+Such an approach can deduce causal relations without depending on sequence in time, negating the problem of HRF.
 
 \subsection{Previous research}
-%[Tom: we did not explicitly have this paragraph, so this is additional effort]
-%[Ramon]: Lijkt me vergelijkbaar met de verwijzingen naar literatuur die we al hebben? Met wat extra referenties naar wat Max zei: temporele methodes zijn onbetrouwbaar, dat hebben wij niet
-%[Ramon]: even een goede balans zoeken tussen dit en "Existing problems" in motivation
-% Previous work
 Previous work on effective connectivity include the aforementioned Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modelling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
@@
 
+In the field of causality, Spirtes et al.\cite{spirtes2000} have described several methods of inferring causal relations.
+Notable is the PC algorithm, named after its authors \textbf{P}eter Spirtes and \textbf{C}lark Glymour.
+Several adaptions of this algorithm have been described, for different assumptions about the input data.
+The PC algorithm runs in two stages; the first stage finds undirected connections between nodes, and the second stage uses these connections to infer causal relations.
+%TODO nog meer previous research over causal discovery? Bayesian approaches?
+
+\subsection{Problem statement}
+%See whether applying causal methods (PC and EMS-PC) can help bring new insights and whether we can find evidence for the existence of causal patterns on a whole brain level.
+This work seeks to utilise the strengths of a current state-of-the-art causal discovery method, namely the PC algorithm, by applying it to resting state brain fMRI data.
+As such methods negate the HRF, and generally have a good run-time performance, this may lead to new applicable methods for determining causal patterns in the human brain.
+
 \subsection{Research goal}
 %How does the standard causal discovery method PC Algorithm behave regarding the aforementioned problems when applied to resting-state functional time-series data?
-%- Can we overcome the consistency problems of PC algorithm applied on resting-state functional data?
+With this research, we want to find out to what extend the PC algorithm can solve the mentioned problems, when applied on resting-state functional time-series data. 
 %- Does the PC algorithm applied on resting-state functional data provide consistent, specific and anatomically plausible connectivity?
+After the algorithm has been applied on this data, the resulting model can be analysed.
+From this it can be concluded whether the PC algorithm provides a way of finding consistent, specific and anatomically plausible connectivity.
 %- How does functional connectivity found by the first part of PC relate to other structural / functional methods applied on the same resting-state functional data regarding the aforementioned problems?
-%
+As the PC algorithm runs in two stages, two results will also be found.
+The first result is a network of correlated brain regions, comparable to a functional connectivity network.
+The second stage finds causal relations, comparable with effective connectivity.
+
 %How does the performance of EMS-PC relate to PC regarding the aforementioned problems when applied to the same resting-state functional time-series data?
-%- When would EMS-PC perform better than PC algorithm regarding directionality results?
+The PC algorithm makes some strong assumptions about the data.
+Brain data generally contains much noise and latent variables, and these assumptions might not hold.
+We have therefore attempted to make the PC algorithm more robust, and we will analyse whether these changes have an impact on the results.
+
+Connectivity may differ per person, but many connections are also present in every person.
+These analyses will therefore be applied both on measurements of individual brains, and on cross-subject aggregations.
+%- When would EMS-PC perform better than PC algorithm regarding directionality results? [Ramon: lijkt me meer iets voor validation]
 %- Is EMS-PC more consistent in its results within a single subject than PC algorithm on resting-state functional data, regarding the multiple separating set addition and the explicit v-structure test?
@@
 
-\subsection{Relevance of research}
+%\subsection{Relevance of research}
 %If HRF can be avoided, causal patterns can be established with greater reliability
@@
 
-% Causal relation
 \paragraph{Causal relation}
@@
 \subsection{PC algorithm}
-% Algorithm description
 The PC algorithm is a so-called constraint based method, where solutions are found by incrementally leaving out or adding parts that do not violate a set of constraints.
@@
 As such, the matrices are divided into blocks: the top left corner represents the connections within the left hemisphere, the bottom left corner represents the connections between the left hemisphere and the cerebellum, etc.
-% For clarification, these are also plotted separately.
 
@@
 Also, the certainties of one-directional arrows are somewhat higher in EMS-PC than in standard PC.
-To illustrate this, the distribution of these values are presented in figure~\ref{fig:distributions}
+For EMS-PC, the highest certainties found for one-directional edges per subject are in the range of 0.45 to 0.9, whereas the avarage across subjects has a maximum value of 0.3.
+The few one-directional arrows found in idividual subjects therefore do not seem to be consistent across different subjects.
+
+To illustrate the difference in certainties between bi-directional and one-directional edges, the distributions of these values are presented in figure~\ref{fig:distributions}.
 \begin{figure}[h!]
@@
 
-%Answer all the questions as mentioned in methods, in the same order
-%No aggregation over direction, motivate this
+\subsection{Execution time of the algorithm}
+To apply the algorithms on our data, we have used a Xeon E5-2670, running on 2,6GHz.
+To run one iteration of the PC algorithm for one subject, modified PC needed approximately one minute and EMS-PC needed approximately two minutes to find both structure and direction.
+Although PC is not designed to be multithreaded, running multiple iterations lends itself for using multiple threads, as the iterations are fully independent, and only need to be averaged out afterwards.
 
@@
 The connections between the two hemispheres can clearly be seen in all subjects, as well as many connections in the corpus collosum.
-The consistency can also be seen in the distribution of values in figure~\ref{distributions}.
+The consistency can also be seen in the distribution of values in figure~\ref{fig:distributions}.
 Non-zero values make up approximately 16 percentage points of the graph, of which 3 percentage points show a certainty larger than 0.5.
 Approximately 1 percentage point even shows a certainty of 1.
-
-%HRF
 
@@
 Both standard PC and EMS-PC find many bi-directional arrows, with some one-directional arrows.
-As can be seen in the distribution of values in figure~\ref{distributions}, nearly all found edges are oriented as bi-directional arrows.
+As can be seen in the distribution of values in figure~\ref{fig:distributions}, nearly all found edges are oriented as bi-directional arrows.
 For EMS-PC, this number is somewhat lower but there are still few one-directional arrows, and they do not have certainties higher than 0.15.
+These certainties are not high enough to conclude any causal relations.
 
@@
 We were unable to find existing literature on connectivity methods regarding significant intra-subject one-directionality.
-
 %Lack of symmetry
-There is little to none symmetry in directionality between the two hemispheres within and across subjects.
+There is also little to none symmetry in directionality between the two hemispheres within and across subjects.
 One would at least expect some symmetry to be found.
-This demonstrates that the gap between analytical methods for effective connectivity and anatomical correctness is still to wide.
+This demonstrates that the gap between analytical methods for effective connectivity and anatomical correctness is still too wide.
 
@@
 This can be done by repeating these experiment with more subjects to get a better view on the significance of the directional edges.
+
+%HRF
+One main strength of a general causal discovery method is that fluctuations in haemodynamic respons do not pose a problem.
+To continue this type of research, other causal discovery methods might be also be applied, and attempts can be made to negate the disadvantages of the PC algorithm.
 
@@
 A Bayesian approach such as proposed by Claassen and Heskes will provide more slack and insight in results \cite{claassen2012}.
-This in turn might leverage development of better and faster data driven methods.
+This in turn will leverage development of better and faster data driven methods.
 

### 2014-01-17 docs/paper/paper final.tex
wat TODOs afgewerkt en data acquisition verplaatst

@@
 Causal relations have only be found with very low certainty. 
-% TODO: first state what you try to do, then how you did it, then what it brought you
 %In this work we applied PC algorithm on resting-state data of healthy human subjects.
@@
 Related is the problem of haemodynamic response fluctuations (HRF).
-In functional MRI-measurements, the delay between brain activity and the indirect measurement of that activity is unknown. %TODO moest MRI nou helemaal uitgeschreven of niet? Ik vind zelf van niet.
+In functional MRI-measurements, the delay between brain activity and the indirect measurement of that activity is unknown.
 Any form of causal inference based on sequence in time might therefore be incorrect; A cause may be measured after its effects, if it has a larger delay.
@@
 It is also not complete.
-There is additional the number of possible graphs in the returned Markov class increasing exponentially with the amount of variables in the provided data, thus not always providing ample information.
+The number of possible graphs in the returned Markov class increases exponentially with the amount of variables in the provided data, thus the network resulting from PC might not be sufficient to describe the underlying structure.
 Also, standard PC is dependent on the order in which these variables are presented, as this is the same order in which edges between pairs of these variables are considered for removal\cite[p.88]{spirtes2000}.
@@
 In practice, the lack of full correctness and completeness are of little concern as these involve only very specific and little occurring connectivity patterns\cite[p.127-130]{spirtes2000}.
-% TODO: Paragraph seems incomplete and some sentences are weird.
 
@@
 
+\section{Data acquisition}\label{sec:data}
+%Subject acquisition and instruction
+%Used equipment and specs
+%Experimental setup
+%Scanner settings
+%Preprocessing
+%- Software
+%- Scripts
+%Resulting data specs
+We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
+This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
+This functional data is divided in 116 nodes and 1029 time-series per subject.
+Additionally, diffusion-weighted images of these subjects have been obtained.
+Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
+We used the six resulting resting-state functional time series data and connectivity matrices from this Bayesian analysis.
+
+All measurements and pre-processing as described in this section has been done during previous research.
+Noadditional pre-processing was needed in our research in order to apply the PC algorithm on this data.
+
 \section{Results}\label{sec:results}
@@
   \centering
-  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop} % TODO remove colorbar, why is it even there
+  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop}
   \includegraphics[height=0.28\textwidth]{images/new/struct_subj2-crop}
@@
 
-As can be seen in figures~\ref{assym_subjects} and~\ref{assym_avg}, the certainties of one-directional arrows are very low compared to the certainties of the bi-directional arrows.
+As can be seen in figures~\ref{fig:assym_subjects} and~\ref{fig:assym_avg}, the certainties of one-directional arrows are very low compared to the certainties of the bi-directional arrows.
 Also, the certainties of one-directional arrows are somewhat higher in EMS-PC than in standard PC.
@@
 
-\appendix
-\section{Data acquisition}\label{sec:data}
-%Subject acquisition and instruction
-%Used equipment and specs
-%Experimental setup
-%Scanner settings
-%Preprocessing
-%- Software
-%- Scripts
-%Resulting data specs
-%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen?
-"This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
-We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
-This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
-This functional data is divided in 116 nodes and 1029 time-series per subject.
-Additionally, diffusion-weighted images of these subjects have been obtained.
-Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
-We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.
-
-No additional pre-processing has been done in our research in order to apply the PC algorithm on this data.
-
 \end{document}

### 2014-01-17 docs/paper/paper final.tex
Finalized introduction.

--- 2014-01-17 docs/paper/paper final.tex
@@
 \section{Introduction}
-%Effective connectivity resting state analysis in human brains
-%Resting state
-%Brains are networks; apply graphical models and causal discovery
-
-% Petty introductory talk
+% Subject introduction
 The brain and particularly the human brain have been studied for hundreds of years.
@@
 A lot has changed since the time of Phrenology, the pseudoscience attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
-Advanced and non-intrusive measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview with widespread clinical and research purposes.
-
-\subsection{ Motivation for research}
-%Better understanding of brains
-%Relatively new field - not yet an established analysis method
+Advanced and non-intrusive measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview, with widespread established clinical and research purposes.
 
 % Motivation for brain connectomics
-The question how our brains are wired - brain connectomics - drives one of the leading fields of research in neuroscience.
+The question how our brains are wired - brain connectomics - drives one of the main fields of research in neuroscience.
 Our brains are in essence networks which can be analysed similarly to very large graphs.
+That is why it is almost natural to apply graphical modelling methods to brain data.
+The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
 Neuro-degenerative diseases such as Alzheimer's disease, Parkinson's disease, dementia, Amyotrophic lateral sclerosis (ALS) have all been shown to severely alter brain connectivity \cite{Bullmore2009}.
 An application of connectivity analysis would be to provide markers to identify the disease and its severity, as well as another means of studying these diseases.
+A helpful analysis tool for finding such markers would be to find which brain areas exert influence on other areas leading  to the unwanted effects.
 
-% Causal discovery and brain networks
-The field of brain connectivity finds its roots in the early 1990s \cite{friston1993functional, friston1994}, though because of more recent developments in the field of Artificial Intelligence and Machine Learning it is possible to do more thorough analysis \cite{vandenheuvel2010}.
-Research regarding effective connectivity - causal relations in brain networks - is relatively new and its application still faces many challenges that require further research \cite{ramsey2010}.
-
-%Existing problems with current methods and applications, specifically [Tom: mention all, or just the relevant problems?][Ramon: de relevante nu benoemd]
-%- Model selection
-
-Ramsey et al. lists several problems in inferring brain connectivity.
-%- Indirect measurements
-One of these problems is that many interesting brain-data is measured indirectly; a measured variable is an effect of the cause we want to know.
+% Problems with current methods
+Research regarding effective connectivity - causal relations in brain networks - is still new and its application still faces many challenges that require further research \cite{ramsey2010}.
+% Model selection
+Methods are unable to fine-tune their results for more than tens of variables, while in practice systems have thousands of variables.
+% Indirect measurements
+Brain data on which these methods are applied are indirect measurements of underlying neuronal activity through blood oxygen levels, water diffusion or magnetic and electrical fields.
 This makes inference hard in practical models: many latent variables may be present, influencing this cause-effect relation.
-%- Distinct but overlapping ROIs
-%- Varying Haemodynamic response (HRF)
-Related is the problem of haemodynamic response fluctuations (HRF).
-In functional MRI-measurements, the delay between brain activity and the indirect measurement of that activity is unknown. %TODO moest MRI nou helemaal uitgeschreven of niet? Ik vind zelf van niet.
-Any form of causal inference based on sequence in time might therefore be incorrect; A cause may be measured after its effects, if it has a larger delay.
-%- Modelling causal structure across individuals (varying responses to same stimuli)
-Also, brain-data is not always consistent, both over time and across different persons. This makes it hard to get significant results.
-%Find shortcomings of PC and improve it (EMS-PC)
-
-Similarly to how graph theory has been applied in functional and anatomical brain networks, it seems natural to apply methods from causal discovery to deduce effective brain networks.
-Such an approach can deduce causal relations without depending on sequence in time, negating the problem of HRF.
-
-\subsection{Previous research}
-Previous work on effective connectivity include the aforementioned Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modelling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
-All of these methods fail to cope with the problems mentioned in the introduction.
-Most of these methods can only handle tens of variables in reasonable time, which becomes problematic in realistic situations with up to tens of thousands of variables.
-
-In the field of causality, Spirtes et al.\cite{spirtes2000} have described several methods of inferring causal relations.
-Notable is the PC algorithm, named after its authors \textbf{P}eter Spirtes and \textbf{C}lark Glymour.
-Several adaptions of this algorithm have been described, for different assumptions about the input data.
-The PC algorithm runs in two stages; the first stage finds undirected connections between nodes, and the second stage uses these connections to infer causal relations.
-%TODO nog meer previous research over causal discovery? Bayesian approaches?
+% Modelling causal structure across individuals
+It is also hard to model connectivity across individuals.
+Up to some extent brain anatomy is similar for multiple subjects, but in modelling tasks the differences are too large to accurately aggregate data.
+% Varying Haemodynamic Response (HRF)
+Even a single brain reacts different over time.
+In functional Magnetic Resonance Imaging (fMRI) the delay between brain activity and the Haemodynamic Response Fluctuates (HRF) over time, blurring temporal relations between cause and effect; a cause may be measured after its effects.
 
@@
 %See whether applying causal methods (PC and EMS-PC) can help bring new insights and whether we can find evidence for the existence of causal patterns on a whole brain level.
-This work seeks to utilise the strengths of a current state-of-the-art causal discovery method, namely the PC algorithm, by applying it to resting state brain fMRI data.
-As such methods negate the HRF, and generally have a good run-time performance, this may lead to new applicable methods for determining causal patterns in the human brain.
+This work seeks to utilise the strengths of an established causal discovery method, namely the PC algorithm, by applying it to resting-state brain fMRI data of the human brain.
+The PC algorithm can cope with hundreds of variables while remaining tractable.
+We evade the problem of HRF by assuming our data comes from a distribution in a state of equilibrium, where temporal components are less strong.
 
-\subsection{Research goal}
-%How does the standard causal discovery method PC Algorithm behave regarding the aforementioned problems when applied to resting-state functional time-series data?
-With this research, we want to find out to what extend the PC algorithm can solve the mentioned problems, when applied on resting-state functional time-series data. 
-%- Does the PC algorithm applied on resting-state functional data provide consistent, specific and anatomically plausible connectivity?
-After the algorithm has been applied on this data, the resulting model can be analysed.
-From this it can be concluded whether the PC algorithm provides a way of finding consistent, specific and anatomically plausible connectivity.
-%- How does functional connectivity found by the first part of PC relate to other structural / functional methods applied on the same resting-state functional data regarding the aforementioned problems?
-As the PC algorithm runs in two stages, two results will also be found.
-The first result is a network of correlated brain regions, comparable to a functional connectivity network.
-The second stage finds causal relations, comparable with effective connectivity.
+We introduce an adaptation to the PC algorithm named EMS-PC that is also theoretically sound and performs at least as good as PC without additional run-time complexity.
+We wonder whether application of both methods provides a means of finding consistent, specific and anatomically plausible connectivity.
+As there is no baseline to compare results with, we compare these methods with another method for finding structural connectivity applied on the same resting-state data.
 
-%How does the performance of EMS-PC relate to PC regarding the aforementioned problems when applied to the same resting-state functional time-series data?
-The PC algorithm makes some strong assumptions about the data.
-Brain data generally contains much noise and latent variables, and these assumptions might not hold.
-We have therefore attempted to make the PC algorithm more robust, and we will analyse whether these changes have an impact on the results.
+% Resting-state
+Resting-state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity; neural activity does not stop when not doing anything in particular.
+Resting-state provides a typical whole brain behaviour that is reliable, robust, unique to a subject and task-independent.
+Measuring resting-state thus provides an advantage when analysing brain connectivity on a global, whole brain level.
+Resting-state analysis has been shown to work well in practice \cite{Lowe2000, doria2010, Bullmore2009}.
 
-Connectivity may differ per person, but many connections are also present in every person.
-These analyses will therefore be applied both on measurements of individual brains, and on cross-subject aggregations.
-%- When would EMS-PC perform better than PC algorithm regarding directionality results? [Ramon: lijkt me meer iets voor validation]
-%- Is EMS-PC more consistent in its results within a single subject than PC algorithm on resting-state functional data, regarding the multiple separating set addition and the explicit v-structure test?
-%- Can we find unfaithfulness within our model without additional independence tests?
+% Previous research
+Previous work on causal methods that have been applied on brain data include Granger causality \cite{roebroeck2005}, Structural Equation Modelling \cite{mclntosh1994}, Multivariate Autoregressive Modelling \cite{harrison2003} and Dynamic Causal Modelling \cite{friston2003}.
+All of these methods fail to cope with the problems mentioned in the introduction.
+Most of these methods can only handle tens of variables in reasonable time even though they are data driven methods.
+This is problematic in realistic situations with up to tens of thousands of variables.
+Some of these methods rely too much on the temporal component, so that application on brain data may not be very reliable.
 
-%\subsection{Relevance of research}
-%If HRF can be avoided, causal patterns can be established with greater reliability
-%Insight and diagnostics in neuro-degenerative diseases
-%Any found evidence of causal patterns can give an impulse to further research
-%[Ramon: is dit niet al duidelijk uit motivation? Lijkt me dubbelop]
+% Structure of the rest of the paper
+In the next section we cover necessary background details on brain connectivity, causal discovery and the PC algorithm.
+In section~\ref{sec:methods} we introduce the EMS-PC method and our experimental approach.
+We then present our results.
+We close with a discussion on the covered material.
 
@@
 Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
-Functional connectivity can be deduced through correlation of Functional Magnetic Resonance Imaging (fMRI) measurements of oxygen saturation  in blood - the BOLD response - while performing a certain cognitive task, or in so called resting-state.
-
-\paragraph{Resting-state}
-Resting-state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity; neural activity does not stop when not doing anything in particular.
-Resting-state provides a typical whole brain behaviour with a low frequency ( < .1 Hz) that is reliable, robust, unique to a subject and task-independent.
-Measuring resting-state thus provides an advantage when analysing brain connectivity on a global, whole brain level.
-Resting-state measurements have the disadvantage that it is difficult to check whether a subject truly is in a resting-state; it is often quite difficult not to think of anything in particular.
-Resting-state analysis has been shown to work well in practice \cite{Lowe2000, doria2010, Bullmore2009}.
+Functional connectivity can be deduced through correlation of fMRI measurements of oxygen saturation  in blood - the BOLD response - while performing a certain cognitive task in resting-state.
 
@@
 An approach for finding analysing effective connectivity may be found in the domain of causal discovery.
+Pearl\cite{pearl2000causality} and Spirtes, Glymour and Scheines\cite{spirtes2000} have described the theoretical basis of causal inference methods.
 A causal discovery method uses data of observed events to determine which of these events are correlated and additionally attempts to determine which events cause which.
@@
 The PC algorithm is a so-called constraint based method, where solutions are found by incrementally leaving out or adding parts that do not violate a set of constraints.
+It is named after its authors \textbf{P}eter Spirtes and \textbf{C}lark Glymour.
 Starting from a fully connected undirected graph, the PC algorithm forms structure by cutting away edges between conditionally independent variables.
@@
 The independence of $X$ and $Y$ is tested given a subset of neighbours of $X$ or $Y$ using a statistical independence test such as the Chi-square or Fisher-Z independence tests.
-%TODO: Add citation to these methods.
 If such an independence is found, this subset is a separating set of $X$ and $Y$ and they are not connected.
@@
 The 'o' mark represents that it is unknown what orientation the specific edge should have thus representing an entire Markov class of solutions rather than a specific instance.
-Like Spirtes et al, we use a star-symbol at the end of an edge to denote that any mark may be present: an arrowhead, an "o" or an empty mark.
+Like Spirtes, Glymour and Scheines, we use a star-symbol at the end of an edge to denote that any mark may be present: an arrowhead, an "o" or an empty mark.
 When an edge is oriented as an edge with a star, this denotes that this end is not changed.
@@
 For a sparse graph, many edges may be removed for small separating sets, and this enables the standard PC algorithm to often run fast in practice perform well on hundreds of variables.
-% TODO: add citation to example publication.
-Peter Lucas maybe or examples in the sprites book?
 
@@
 
-\section{Methods}% (our contributions)
+\section{Methods}\label{sec:methods}% (our contributions)
 Standard PC provides a good starting point for an exploratory point of view in causal discovery.
@@
 % Application to data
-We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al\cite{hinne2013}.
+We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al.\cite{hinne2013}.
 1030 volumes were obtained per subject.
@@
 To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
-These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:assym_subjects}.
-The avarage across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:assym_avg}.
+These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}.
+The avarage across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.
 
@@
   \caption{The assymetric parts of the PDAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
-\label{fig:assym_subjects}
+\label{fig:asym_subjects}
 \end{figure}
@@
 The avarage of all six subjects is taken.}
-  \label{fig:assym_avg}
+  \label{fig:asym_avg}
 \end{figure}
 
-As can be seen in figures~\ref{assym_subjects} and~\ref{assym_avg}, the certainties of one-directional arrows are very low compared to the certainties of the bi-directional arrows.
+As can be seen in figures~\ref{fig:asym_subjects} and~\ref{fig:asym_avg}, the certainties of one-directional arrows are very low compared to the certainties of the bi-directional arrows.
 Also, the certainties of one-directional arrows are somewhat higher in EMS-PC than in standard PC.

### 2014-01-17 docs/paper/paper final.tex
Removed some comment statements.

@@
 
-%EMS-PC extension (no pseudo-code, but additions regarding standard PC)
+%EMS-PC extension
 \paragraph{EMS-PC}
@@
 \paragraph{Experimental pipeline}\label{sec:pipeline}
-%- What type of data is needed
+% What type of data is needed
 PC algorithm requires observational data where each instance is a vector of single observations per variable.
@@
 As we cannot verify this directly, we compare the total count, consistency and standard-deviation of one-directional arrows.
-%- Run-time of methods and comparison between run-times
+% Run-time of methods and comparison between run-times
 We also compare run-time per iteration between both.
+
+\section{Data acquisition}\label{sec:data}
+%Subject acquisition and instruction
+%Used equipment and specs
+%Experimental setup
+%Scanner settings
+%Preprocessing
+%- Software
+%- Scripts
+%Resulting data specs
+We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
+This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
+This functional data is divided in 116 nodes and 1029 time-series per subject.
+Additionally, diffusion-weighted images of these subjects have been obtained.
+Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
+We used the six resulting resting-state functional time series data and connectivity matrices from this Bayesian analysis.
+
+All measurements and pre-processing as described in this section has been done during previous research.
+No additional pre-processing was needed in our research in order to apply the PC algorithm on this data.
 
@@
 \subsection{Structural graphs}
-%- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
-%Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
-%This is done both per subject, and as an avarage over all subjects.
-%We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
+% Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
+% Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
+% This is done both per subject, and as an avarage over all subjects.
+% We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
 Modified PC and EMS-PC do not differ in finding the structure.
 As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
-The results are presented per subject in figure~\ref{fig:struct_subjects} and as an avarage across all subjects in figure~\ref{fig:struct_avg}.
+The results are presented per subject in figure~\ref{fig:struct_subjects} and as an average across all subjects in figure~\ref{fig:struct_avg}.
 For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.
@@
 This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
-As there were no notable differences between different subjects, only the avarage PDAG across all subjects is presented.
+As there were no notable differences between different subjects, only the average PDAG across all subjects is presented.
 The results for standard PC and EMS-PC are shown in figure~\ref{fig:pdag_avg}.
@@
   \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_expl-crop}
-  \caption{The PDAG found standard PC at the left, and EMS-PC at the right. A high value denotes a high certainty of an arrowhead. The avarage of all six subjects is taken.}
+  \caption{The PDAG found standard PC at the left, and EMS-PC at the right. A high value denotes a high certainty of an arrowhead. The average of all six subjects is taken.}
   \label{fig:pdag_avg}
@@
 These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}.
-The avarage across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.
+The average across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.
 
@@
   \includegraphics[width=0.49\textwidth]{images/new/assym_avg_expl-crop}
-  \caption{The assymetric parts of the PDAG found with standard PC on the left, and EMS-PC on the right.
+  \caption{The asymmetric parts of the PDAG found with standard PC on the left, and EMS-PC on the right.
 A high value denotes a high certainty of an one-directional arrow.
-The avarage of all six subjects is taken.}
+The average of all six subjects is taken.}
   \label{fig:asym_avg}
@@
 Many structural connections found by the PC algorithm are found consistently, both within one subject and also across multiple subjects.
-The connections between the two hemispheres can clearly be seen in all subjects, as well as many connections in the corpus collosum.
+The connections between the two hemispheres can clearly be seen in all subjects, as well as many connections in the corpus callosum.
 The consistency can also be seen in the distribution of values in figure~\ref{fig:distributions}.

### 2014-01-17 docs/paper/paper final.tex
margins weggehaald, scheelt een halve pagina
Fixed bibliography typos. Addition on unfaithfulness check.

@@
 % Margins & Headers
-\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
-\columnsep 0.4in
-\footskip 0.4in 
+%\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
+%\columnsep 0.4in
+%\footskip 0.4in 
 \usepackage{changepage}
@@
 The PC algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
-However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test d-separation of every pair of nodes.
+However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test \textit{d}-separation of every pair of nodes.
 For a sparse graph, many edges may be removed for small separating sets, and this enables the standard PC algorithm to often run fast in practice perform well on hundreds of variables.
@@
 %EMS-PC extension
-\paragraph{EMS-PC}
+\subsection{EMS-PC}
 We have combined the above ideas - multiple separating sets, unfaithfulness check and explicit orientation check - into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
@@
 \subsection{Experimental validation}
-\paragraph{Experimental pipeline}\label{sec:pipeline}
 % What type of data is needed
@@
 
-% Application to data
+\paragraph{Data acquisition}
 We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al.\cite{hinne2013}.
-1030 volumes were obtained per subject.
 After preprocessing the fMRI data was parcellated according to the Automated Anatomical Labeling (AAL) atlas in 116 regions of interest.
-The exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing is provided in appendix~\ref{sec:data}.
+This provides 1029 samples with 116 standard-normally distributed variables.
+For the exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing we refer to the paper by Hinne et al. 
+Additionally, diffusion-weighted images of these subjects have been obtained from the same study.
+These were provided as six $116 \times 116$ binary connectivity matrices.
+No additional preprocessing was needed for application of our methods.
 We use Fisher's z-test for conditional independence to match the continuous nature of the data.
 
-% Implementation
+\paragraph{Implementation}
 We have implemented standard PC and EMS-PC in Matlab using statistical tests and ideas from the Bayesian Network Toolbox for Matlab.\footnote{Bayesian Network Toolbox for Matlab: https://github.com/bayesnet/bnt}
@@
 
-% Parameter selection
+\paragraph{Parameter selection}
 Due to the lack of a baseline to compare results with, dependency thresholds were chosen by hand.
@@
 
-\section{Data acquisition}\label{sec:data}
-%Subject acquisition and instruction
-%Used equipment and specs
-%Experimental setup
-%Scanner settings
-%Preprocessing
-%- Software
-%- Scripts
-%Resulting data specs
-We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
-This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
-This functional data is divided in 116 nodes and 1029 time-series per subject.
-Additionally, diffusion-weighted images of these subjects have been obtained.
-Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
-We used the six resulting resting-state functional time series data and connectivity matrices from this Bayesian analysis.
-
-All measurements and pre-processing as described in this section has been done during previous research.
-No additional pre-processing was needed in our research in order to apply the PC algorithm on this data.
-
 \section{Results}\label{sec:results}
@@
 In all matrices, indices 1-45 represent the brain regions of the left hemisphere, indices 45-90 represent the right hemisphere, and indices 91-116 represent regions in the cerebellum.
-As such, the matrices are divided into blocks: the top left corner represents the connections within the left hemisphere, the bottom left corner represents the connections between the left hemisphere and the cerebellum, etc.
+As such, the matrices are divided into blocks: the top left corner represents the connections within the left hemisphere, the bottom left corner represents the connections between the left hemisphere and the cerebellum, and so on.
 
@@
 % We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
-Modified PC and EMS-PC do not differ in finding the structure.
+Modified PC and EMS-PC do not differ in finding functional connectivity.
 As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
@@
 This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
-As there were no notable differences between different subjects, only the average PDAG across all subjects is presented.
+As there were little to none notable differences between different subjects, only the average DAG across all subjects is presented.
 The results for standard PC and EMS-PC are shown in figure~\ref{fig:pdag_avg}.
-Although many arrowheads are found with probabilities above 0.9 for both variants of the algorithm, the graph looks strongly symmetric.
+Although many arrowheads are found with frequency ratio above 0.9 for both variants of the algorithm, the graph looks strongly symmetric.
 In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.
@@
   \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_expl-crop}
-  \caption{The PDAG found standard PC at the left, and EMS-PC at the right. A high value denotes a high certainty of an arrowhead. The average of all six subjects is taken.}
+  \caption{The partially DAG found by standard PC on the left, and EMS-PC on the right.
+A high value denotes a high certainty of an arrowhead.
+The sum of edge frequencies of all six subjects is taken.}
   \label{fig:pdag_avg}
@@
 To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
-These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}.
+The resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}.
 The average across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.
@@
   \includegraphics[height=0.25\textwidth]{images/new/assym_subj6_expl-crop}
-  \caption{The assymetric parts of the PDAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
+  \caption{The asymetric parts of the partial DAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
 \label{fig:asym_subjects}
@@
   \includegraphics[width=0.49\textwidth]{images/new/assym_avg_expl-crop}
-  \caption{The asymmetric parts of the PDAG found with standard PC on the left, and EMS-PC on the right.
+  \caption{The asymmetric parts of the partial DAG found with standard PC on the left, and EMS-PC on the right.
 A high value denotes a high certainty of an one-directional arrow.
@@
 \end{figure}
+
+Our additional unfaithfulness check without additional independence tests were unable to detect any unfaithful v-structures.
 
@@
 These certainties are not high enough to conclude any causal relations.
+The additional unfaithfulness test proved inconclusive, as no unfaithful v-structures were found.
+This is not bad news, but it does suggest there are methodological details being skipped over.
 
@@
 
+Instead of targeting the brain as a whole in resting-state, one could focus more on specific parts to increase the modelling detail.
+In such a case task-based data would work well.
+An area corresponding to a specific task, for instance the motor cortex, could be inspected in relationship with itself and other areas.
+
 A causal network or Bayesian network cannot contain cycles by definition.
@@
 
-% TODO: Suggereren task based analysis ipv resting state voor beter causaal beeld.
-
-%cause of poor consistency: inherent to our method or viable for further research?
-
-% \section{Conclusion}
-%repeat everything in 1 paragraph, or leave away, whatever seems most fitting in the end. HRF makes us awesome
-
 \bibliography{references}{}

### 2014-01-17 docs/paper/paper final.tex
plaatjes in 2x3 opgedeeld ipv 3x2 zodat ze wat beter zichtbaar zijn O:-)

--- 2014-01-17 docs/paper/paper final.tex
@@
   \centering
-  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop}
-  \includegraphics[height=0.28\textwidth]{images/new/struct_subj2-crop}
-  \includegraphics[height=0.28\textwidth]{images/new/struct_subj3_colorbar-crop} % one colorbar at the top right of the figure
-  \includegraphics[height=0.28\textwidth]{images/new/struct_subj4-crop}
-  \includegraphics[height=0.28\textwidth]{images/new/struct_subj5-crop}
-  \includegraphics[height=0.28\textwidth]{images/new/struct_subj6-crop}
+  \includegraphics[height=0.29\textheight]{images/new/struct_subj1-crop}
+  \includegraphics[height=0.29\textheight]{images/new/struct_subj2_colorbar-crop} % one colorbar at the top right of the figure
+  \includegraphics[height=0.29\textheight]{images/new/struct_subj3-crop}
+  \includegraphics[height=0.29\textheight]{images/new/struct_subj4-crop}
+  \includegraphics[height=0.29\textheight]{images/new/struct_subj5-crop}
+  \includegraphics[height=0.29\textheight]{images/new/struct_subj6-crop}
   \caption{The structure of all six subjects as found with the PC algorithm. The two diagonal lines represent connections between the two hemispheres.}
@@
   \centering
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj1_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj2_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj3_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj4_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj5_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj6_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj1_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj2_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj3_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj4_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj5_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj6_expl-crop}
   \caption{The assymetric parts of the PDAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}

### 2014-01-17 docs/paper/paper final.tex

--- 2014-01-17 docs/paper/paper final.tex
@@
   \centering
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj1_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj2_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj3_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj4_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj5_expl-crop}
-  \includegraphics[height=0.25\textwidth]{images/new/assym_subj6_expl-crop}
-  \caption{The asymetric parts of the partial DAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
-\label{fig:asym_subjects}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj1_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj2_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj3_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj4_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj5_expl-crop}
+  \includegraphics[height=0.26\textheight]{images/new/assym_subj6_expl-crop}
+  \caption{The assymetric parts of the PDAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
+label{fig:asym_subjects}
 \end{figure}

### 2014-01-17 docs/paper/paper final.tex
legenda en labels bij assen toegevoegd in de verdelingsgrafieken

@@
 % Needed packages
-\documentclass[a4paper, 10pt, english, onecolumn]{article}
+\documentclass[a4paper, english]{article}
 \usepackage[english]{babel}
-\usepackage[cm]{fullpage}
+%\usepackage[cm]{fullpage}
 \usepackage{cite}
-\usepackage{anysize}
+%\usepackage{anysize} <- this makes the article much wider
 \usepackage{setspace}
@@
 \usepackage{hyperref}
-
 \usepackage{amssymb,amsmath}
@@
 \usepackage[noend]{algpseudocode}
-
 % for coloring individual cells in a table
@@
 \begin{figure}[h!]
+\hspace{-0.5in}
+\begin{minipage}[t]{1.2\textwidth}
   \centering
@@
   \includegraphics[height=0.29\textheight]{images/new/struct_subj6-crop}
+  \end{minipage}
 \caption{The structure of all six subjects as found with the PC algorithm. The two diagonal lines represent connections between the two hemispheres.}
+
 \label{fig:struct_subjects}
@@
   \caption{The avarage structure of all subjects.
-The structure on the left is found with the PC-algorithm, and the structure on the right is found with DWI-measurements by Hinne et al.
+The structure on the left is found with the PC algorithm, and the structure on the right is found with DWI-measurements by Hinne et al.
 The two diagonal lines represent connections between the two hemispheres.}
@@
 \begin{figure}[h!]
+\hspace{-0.4in}
+\begin{minipage}[t]{1.2\textwidth}
   \centering
@@
   \includegraphics[height=0.26\textheight]{images/new/assym_subj6_expl-crop}
-  \caption{The assymetric parts of the PDAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
-label{fig:asym_subjects}
+\end{minipage}
+  \caption{The asymetric parts of the partial DAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
+\label{fig:asym_subjects}
 \end{figure}
@@
   \centering
-  \includegraphics[width=0.49\textwidth]{images/new/dist_all_mod_blackstruct_reddir_blueantisym-crop}
-  \includegraphics[width=0.49\textwidth]{images/new/dist_all_expl_blackstruct_reddir_blueantisym-crop}
+  \includegraphics[width=0.49\textwidth]{images/new/dist_all_mod_legend-crop}
+  \includegraphics[width=0.49\textwidth]{images/new/dist_all_expl_legend-crop}
   \caption{The distribution of values, for standard PC at the left, and EMS-PC at the right.
-The horizontal axis shows the relative number of occurences in the graph.
 Zero-values make up more than 84\% of the graph, these are not shown in this distribution.

### 2014-01-17 docs/paper/paper final.tex
veel kleine verbeteringetjes aan de hand van feedback van een proeflezer (credit to Petra)

@@
   \textbf{T. de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}\\
-%  \textbf{T. Claassen} - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
-%  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
+  \small Supervised by\\
+  \small T. Claassen - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
+  \small M. Hinne - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
 }
@@
 But few general causal discovery methods have yet been used to find effective connectivity.
-In this article, we describe how the PC algorithm is used on functional brain-data.
+In this article, we describe how the PC (Peter-Clark) algorithm is used on functional brain-data.
 This approach results in connectivity matrices which clearly show some well-known brain connections.
 Some disadvantages of the PC algorithm for this use are discussed, and an attempt is made to decrease these disadvantages by making adjustments to the algorithm.
-Causal relations have only be found with very low certainty. 
+Causal relations have only been found with very low certainty. 
 %In this work we applied PC algorithm on resting-state data of healthy human subjects.
@@
 A lot has changed since the time of Phrenology, the pseudoscience attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
-Advanced and non-intrusive measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview, with widespread established clinical and research purposes.
+Advanced and non-intrusive measurement methods exist that allow scientists to peek inside brains, giving a coarse but broad overview, with widespread established clinical and research purposes.
 
@@
 %See whether applying causal methods (PC and EMS-PC) can help bring new insights and whether we can find evidence for the existence of causal patterns on a whole brain level.
-This work seeks to utilise the strengths of an established causal discovery method, namely the PC algorithm, by applying it to resting-state brain fMRI data of the human brain.
+This work seeks to utilise the strengths of an established causal discovery method, by applying it to resting-state brain fMRI data of the human brain.
+The algorithm to be used is the the PC algorithm, named after \textbf{P}eter Spirtes and \textbf{C}larck Glymour, who developed it.
 The PC algorithm can cope with hundreds of variables while remaining tractable.
@@
 
-We introduce an adaptation to the PC algorithm named EMS-PC that is also theoretically sound and performs at least as good as PC without additional run-time complexity.
+We introduce an adaptation of the PC algorithm named EMS-PC that is also theoretically sound and performs at least as good as PC without additional run-time complexity.
 We wonder whether application of both methods provides a means of finding consistent, specific and anatomically plausible connectivity.
-As there is no baseline to compare results with, we compare these methods with another method for finding structural connectivity applied on the same resting-state data.
+As there is no baseline to compare results with, we compare these methods with DWI-measurements found by Hinne et al.\cite{hinne2013} for finding structural connectivity applied on the same resting-state data.
 
@@
 \subsection{Execution time of the algorithm}
-To apply the algorithms on our data, we have used a Xeon E5-2670, running on 2,6GHz.
+To apply the algorithms on our data, we have used a Xeon E5-2670, running on 2,6\emph{GHz}.
 To run one iteration of the PC algorithm for one subject, modified PC needed approximately one minute and EMS-PC needed approximately two minutes to find both structure and direction.
